%!TeX root = ../incremental_SS_Translation.tex

\chapter{Śārīrasthāna 9:  An Analysis of the Pipes}

\section{Introduction}

\sse{dhamanī}{pipe}%
\sse{sirā}{duct}%
\sse{srotas}{tube}%
\sse{nāḍī}{tube}%
\sse{snāyu}{sinew}%
\sse{kaṇḍarā}{tendon}%

Ancient Indian physicians and surgeons were aware of the network of
vessels and tendons in the body and they debated their appearence and
functions. There are a number of different Sanskrit names for these
vessels, such as \emph{sirā}, \emph{dhamanī}, \emph{srotas},
\emph{nāḍī}, \emph{snāyu}, and \emph{kaṇḍarā}, and these terms pose a
challenge to the translator.  Descriptions of the first four
characterize hollow tubes, and other translators have used “vein” and
“artery” for at least two of them.  But the heart was not thought of
as a pump in the Āyurvedic view of the body, nor did ancient Indian
physicians think that the blood circulated in the post-Harveian sense.
Blood moved outwards radially from the centre of the body, like water
from a water-tank irrigating fields.  There was certainly no concept
of a contrast between venous and arterial circulation, and several of
these vessels are most commonly seen as being rooted in the navel, not
the heart.  So it seems best not to mis-read history and assign provocatively
modern names to these vessels.

\subsubsection{Function}

\label{intro:tubes} There is also the interesting question of what it
is that was actually considered to be flowing in these vessels.  The
\emph{sirā} vessels most often carry blood, and “blood-letting” is
often called “piercing the \emph{sirā}s."  The the \emph{dhamanī}
vessels most often conduct wind (Skt.\,{\small$\sqrt{}$}\emph{dham},
“blow”), an idea strongly reminiscent of the classical Greek doctrine
of Praxagoras of Cos (fl.\ ca.\ 300 \BCE), whose
\emph{pneuma}-carrying arteries started in the heart and spread out
into tubular \emph{neura} that dwindled into
nerves.\footnote{\cites[17--19,
        49--53]{stec-1958}[137]{phil-1973}[137--159 et passim]{wals-2016}.}
But all the Ayurvedic vessels seem also to transport other substances,
including humours, waste products, sensations, and
perceptions.\footnote{\citet[344--52]{dasg-hist} provided a clear
    account of this topic.} In āyurveda (as opposed to tantra or yoga),
the \emph{nāḍī} vessels are primarily discussed as the locus of the
pulse, though what it is that pulses is not made explicit.  In one
case Vāgbhaṭa used \emph{nāḍī} to refer to the windpipe.  Faced with
the word \emph{snāyu}, one is virtually obliged to use its English
cognate term “sinew.”  But the \emph{snāyu}s seem sometimes to refer
to what are today called nerves rather than to sinews or tendons, and
indeed this is the primary meaning of the cognate term that has
evolved in several contemporary Indian languages. The word
\emph{kaṇḍarā} more unambiguously refers to tendons.

\subsubsection{Terminology}

I have chosen to use neutral terms like “pipe,” “tube,” “duct,” and
“sinew” to translate some of the above terms, thus retaining the
original distinctions but avoiding an inappropriate and a-historical
identification with the anatomical distinctions developed especially
since the discoveries of William Harvey 
(1578--1657).\footnote{\citet[27--29]{kutu-1962} presented a thoughtful and 
perceptive discussion of this topic with sensitivity to its medical-historical 
context.}

Adhyāyas 7 and 8 of the \emph{Śārīrasthāna} described the \emph{sirā}s
and adhyāya 9 describes the \emph{dhamanī}s, with the \emph{srotas} being
described at the end of chapter~9.

\subsection{Fluids and Their Conduits}

The types of fluid in the āyurvedic body include \se{rakta}{blood},
milk, semen, \se{prāṇa}{breath}, the \se{rasa}{the juice of digested
    food}, and the humours \se{vāta}{wind}, \se{pitta}{bile}, and
\se{kapha}{phlegm}.

%Energy is present throughout the whole body, and is
%thought to be cold, oily, and solid.  It is of the \skt{cooling
%principle}{saumya}, and is believed to give the body its power and
%nourishment.

%from Tarabout.tex:

These fluids are transported from place to place by three principle
types of conduit: \se{sirā}{ducts}, \se{dhamanī}{pipes}, and
\se{srotas}{tubes} described above. Given the importance of this
system of fluid distribution to the Āyurvedic physiology, surprisingly
little work has been done on clarifying what these conduits do, and
how they are explained in āyurvedic theory.\footnote{Exceptions
    include \citet[ii.13]{dasg-hist} and \citet[ch.\,2]{kutu-1962}, who
    both invested effort to visualize and describe the ayurvedic bodily
    conduits.}

This lack of interest in the fluid mechanics of the anatomy is notSurendranath 
Dasgupta
new.  Early Chinese and Japanese medical books abound in diagrams of
the body carrying lines illustrating channels of \emph{chi}, but no
Indian book or manuscript to my knowledge shows maps of the medical
ducts, pipes, or tubes.  In fact, medical illustration does not become
established as a tradition in India until relatively recently, only
getting underway seriously with the production of Sanskrit and
vernacular medical printed texts in the second half of the nineteenth
century.\footcite{wuja-2008a}  And that initiative too, being inspired
and influenced by Western anatomy, largely ignores the traditional
conduits of the āyurvedic body.\footnote{\cite{birc-2024,lust-2018}.}

\subsubsection{\se{sirā}{Ducts}}

According to the \emph{Suśruta\-saṃhitā}, the function of the 700
ducts is to carry wind, bile, phlegm and blood around the body,
starting from their origin in the navel. In a vivid pair of metaphors,
one agricultural and one botanical, Suśruta's text described the ducts
as follows:
\begin{quote}
    As a garden or a field is irrigated by water-carrying canals, and
each part receives nourishment, so the ducts provide nutrition to
the body by means of their contraction and dilation. Their
branches are just like the veins on a leaf.\footnote{\SS\
    \Su{3.7.3}{376}.}
\end{quote}
A point of special interest is that the ducts are coloured according
to what they carry: those carrying wind are \se{aruṇa}{yellowish
    brown}, those carrying bile are dark blue, those carrying phlegm are
white, and those carrying blood are red.\footnote{Ibid.,
    \Su{3.7.18}{377}.} It seems likely that these distinctions are based
    on the observation of different-coloured vessels under the surface of
    the skin.  In yet another simile, Suśruta likens the distribution of
    these ducts from the umbilical centre through the body to the spokes
    radiating from the centre of a wheel.\footnote{Ibid., \Su{3.7.5}{376}.}

\subsubsection{\se{dhamanī}{Pipes}}

There are said to be twenty-four pipes in the body (Su.śā.9). Like the
ducts, they originate in the navel. From there, ten go up, ten down,
and four sideways.

Those which go up from the navel support the body by carrying
\se{viśeṣa}{particular items} such as sound, touch, vision, taste,
smell, \se{praśvāsa}{out-breath}, \se{ucchvāsa}{in-breath}, yawning,
sneezing, laughter, speech, crying, etc.  These ten pipes go from the
navel to the heart and there each one divides into three branches,
thus producing thirty pipes.  Ten of these are devoted to carrying the
humours, wind, bile, and phlegm, as well as blood and nutritive fluid
(two pipes for each substance).  Eight more carry sense impressions:
sound, form, taste, and smell (again, two pipes each).  Two pipes are
used for \se{bhāṣā}{speech}, two for making \se{ghoṣa}{sound}, two for
sleeping, and two more for waking up. Two pipes carry tears.  Two
pipes connected to the breasts carry women's breast-milk; curiously,
in men the same two pipes are said to carry semen from the breasts.

Those pipes which go down from the navel carry substances such as
wind, urine, faeces, semen, and menstrual blood.

At this point things become a little complicated.  On reaching the
receptacle of bile the pipes separate out the nutritive juice which
has resulted from digestion and carry it as refreshment to the whole
body, also supplying it to the upper and horizontal pipes. They fill
the receptacle of nutritive juice. And they separate out the three
principle impurities: urine, faeces, and sweat.

In between the receptacles of raw and digested food, the pipes
divide into three branches, as before.  The first ten pipes
have the same functions as the first ten upward pipes. The
next two carry food to the intestines, and another two carry
water.  Two carry urine to the bladder.  Two generate and
transport semen, and two make it ejaculate.  In women, the
same four pipes carry and discharge menstrual blood.  Two
pipes are connected to the intestines and function in
defecation. The remaining eight pipes supply sweat to the
horizontal pipes.

The four pipes which run sideways are said to subdivide hundreds of
thousands of times, holding the body together in a network. Their ends
are connected to the hair follicles, and through these sweat is
carried out and nutritive juice is carried in.  This is how massage
oils, showers, and ointments can move through the skin and affect the
body internally.  They are also the means by which pleasant and
unpleasant sensations of touch are experienced.

\subsubsection{\se{srotas}{Tubes}}

According to the \SS, there are initially twenty-two tubes in the
body, two for each of eleven substances.\footnote{\SS\
    \Su{3.9.12--13}{386--387}; cf.\ \CS\ \Ca{3.5}{249--252}.} Two of the
    \sepl{srotas}{tube} carry \se{prāṇa}{breath}, and are joined to the
    heart and the \sepl{dhamanī}{pipe} which carry nutritive juice. Two
    more carry food, and are joined to the food-carrying pipes and the
    stomach. Two carry water and are joined to the palate and the
    \se{kloman}{lung}. Two carry nutritive juice and are joined to the
    same places as those carrying breath.  Two carry blood, and are joined
    to the liver, the spleen, and the pipes which carry blood. Two carry
    flesh, and are joined to the ligaments, skin, and pipes which carry
    blood. Two carry fat and are joined to the \se{kaṭī}{waist} and the
    kidneys.  Two carry urine and are joined to the bladder and penis. Two
    carry faeces and are joined to the receptacle of digested food and the
    rectum. Two carry semen and are joined to the breasts and testicles. 
    Two carry menstrual blood and are joined to the womb and the pipes
    which carry menstrual blood. (There is no suggestion that these last
    pairs are specific to either gender.)

The \CS\ adds three more categories of tube: two carrying bone, two
carrying marrow (completing the set of seven basic \sepl{dhatu}{body
    element}), and two carrying sweat.  It omits menstrual blood. The
\CS's description of the various roots of the tubes is different in
small details from the \SS's, and it also asserts that the humours
wind, bile, and phlegm are carried in them indiscriminately and all
over the body.\footnote{Ca.vi.5.}

Like the horizontal pipes, the tubes in the body divide and
subdivide into innumerable tiny branches.

In contrast to the ducts and pipes, the description of these tubes is
embedded in a discourse of injury, and the symptoms arising from
damage to them are listed.

The \SS\ recorded the existence of an ancient disagreement amongst
physicians as to whether the pipes, ducts and tubes are really
separate types of vessel, and in particular whether there is a
significant difference between \sepl{dhamanī}{pipe} and
\sepl{srotas}{tube}. As the text argued, there is indeed a difference
between these three types of vessel: they look different, they have
different connections, and they have different functions.  The authoritative
tradition of medical science also asserts their difference.  It is
merely because of their close proximity, similarity, and small size
that they are conflated.  The \CS\ also testified to contemporary
debates about the nature of these vessels; it recorded -- and
rejected -- an extreme view that the human body consists only of a
conglomeration of tubes.
\begin{quote}
    Conduits such as the ducts grow in the flesh just like lotus roots
    located in muddy water grow in all directions in the ground.\footnote{\SS\ 
    \Su{3.4.9}{356}.}
\end{quote}

\begin{quote}
    From the heart it enters the twenty-four \sepl{dhamanī}{pipe}. Ten go
    up, ten go down, and four are horizontal.  Then, through an invisible
    agency, it nourishes the whole body, day in, day out, making it grow,
    holding it up, and making it go.  One can mark its passage as it
    courses through the body by inference based on whether diseases are
    caused by diminution or by superfluity.  This \se{rasa}{nutritive
        juice} courses through all parts of the body, through the humours,
    body tissues, impurities, and organs.\footnote{\SS\ \Su{1.14.3}{59}.}
\end{quote}


\section{Literature}

\tvolcite{1A}[258--259]{meul-hist} summarized this chapter of the \SS\
and documented the relevant research literature to 2002. On these
different conduits, see further
\cites[404--406]{wuja-2022}[xlvi--xlvii]{wuja-2003}[26--28]{ray-1980}[23--29 
and tables]{kutu-1962} and the admirably clear description by 
\tvolcite{2}[344--352]{dasg-medi}.

\newpage

\section{Translation}

\begin{translation}

\subsection{On the pipes in the body}

    \item [1]

    Next, we shall discuss the analysis  of the \sepl{dhamanī}{pipe}.

    \item[3]

    There are twenty-four pipes.  And they originate from the navel.

    Regarding that, some teachers have said that there is no difference
    between \sepl{sirā}{duct}, \sepl{dhamanī}{pipe} and
    \sepl{srotas}{tube}, because pipes and tubes are just types of duct.
    But here, it is said that that is not correct. Pipes, tubes  and ducts
    are quite different. Why? Because of the difference in their colours,
    in their connections to their roots, because of their divergent
    functions, and because of traditional doctrine. Their distinct
    functions only appear to have no functional difference because of the
    close proximity of their mutual ramifications, the similarity of their
    functions, and their smallness.

    \item[4]

    Those vessels originate in the navel, you see.  Ten of them go up, ten
    go down, and four go horizontally.

\subsubsection{Pipes that go up}
    \item[5]

    You see, the ones that go up support the body, conveying particular
    things like sound, touch, visible form, taste, smell, exhalation,
    inhalation, vision, sneezing, yawning, laughter, speaking, crying and
    so on.  But when they near the heart, they become three by three.
    There are thirty.  Ten of them carry two each of wind, bile, phlegm,
    blood and chyle. He grasps sound, touch visual form, taste and smell
    by means of eight. He speaks by means of two; he makes sound by means
    of two.\footnote{\Dalhana{3.9.5}{384} described “making a sound” as
        the production of an indistinct sound, as opposed to fully-articulated
        speech.}  By means of two, he sleeps, by means of two he wakes up. Two
    carry blood;  two connected to the breasts carry womens'
    breastmilk.\footnote{\dev{asra} “blood” can also mean tears, and this
        is the interpretation of the vulgate.  But the anatomy of women's milk
        and menstruation suggests “blood.” Ḍalhaṇa did not comment.} Two carry
    the semen of a man from the two breasts.

    Thus, these thirty, counting their subdivisions, have been described.
    They hold up and sustain the parts above the navel, i.e., the belly,
    flanks and back, the shoulders, neck head and arms.\footnote{The sequence
        of next passages in the Nepalese version differs from the vulgate. The vulgate's
        passage no.\,6 is not present in the Nepalese version.}

    \item[11] 
    
    And there is a verse about this:\footnote{In the
    vulgate, this verse appears near the end of the chapter, before
    the description of \dev{srotas}.}
    \begin{sloka}
        The five \sepl{adhibhūta}{primary element}, becoming five, bring
        into existence the five senses in five ways. After bringing the
        five senses into existence in five ways,  they dissolve into five
        items at the time of death.
    \end{sloka}


\subsubsection{Pipes that go down}

    \item [7]

    The ones that go downwards convey wind, urine, feces, semen and
    menstrual blood.  They arrive at the \se{pittāśaya}{gall bladder}.
    Then, separating out the essence of food and drink that is located
    there and has been matured by heat, they provide nourishment to those that
    go upwards and to the ones that go sideways, and they fill up the locations of the
    \se{rasa}{chyle}.\footnote{\Dalhana{3.9.7}{385} glossed ``the locations of the \dev{rasa}
        as ``the heart."}

    Every pair of these carry wind, bile, phlem and chyle.\footnote{Or ``The fluids wind, bile and phlegm.''} There are ten of these.  Two of them convey food.  Attached to the intestines, two water-carrying ones each connect to the urinary bladder.  And two are for making semen apear. In this manner, two of them carry blood.  For women, two of them are defined as being for menstrual fluid. Two are connected to the large intestine for expelling feces. Another eight horizontal ones deliver sweat. 

    These are the thirty divisions that have been explained. They support these items below the navel: the intestine, \se{kaṭī}{waist}, urine, bladder, anus, penis and thighs. 


\item[8]

And there is a verse on this:
\begin{sloka}
   The downward-moving ones perform all these functions completely.  Now I shall accurately explain the horizontal-moving ones, with their function. 
\end{sloka}

\subsubsection{Pipes that go horizontally}
\item [9]

Each one of the four horizontally-moving ones is split one hundred times
and a thousand times. And these are innumerable. They bind and stretch
this \se{gavākṣita}{lattice-like} body.  With their openings being
connected to the pores, they allow sweat to flow out and
\se{rasa}{chyle} to bring in refreshment  internally and externally. They
cause the \sepl{vīrya}{power} of anointment, showering, bathing, and
lotions, being matured on the skin, to be conveyed to the interior of body.
It is these, in fact, that give one the experience of touch, whether pleasant or 
unpleasant.
These four pipes that reach every part of the body, with their subdivisions, have 
been explained. 


\item[9.add]

And there are two verses on this:
\begin{sloka}
    All these pipes are interconnected like the senses. Whether waking
or sleeping they convey in the realm of the senses.\footnote{The
    meaning of this verse is not perfectly clear.  The verse is not
    present in the vulgate version of the text and the readings of
    some parts of the text in witness H are not certain.}
\end{sloka}


\item[10]

\begin{sloka}
They have openings through which \se{rasa}{chyle} is accumulated, just like
the naturally-occurring holes in lotus stalks and fibres.\footnote{The 
subsequent verse 3.9.11 in the vulgate appears earlier in this chapter in the 
Nepalese 
version.} 
\end{sloka}


\subsection{On the tubes in the body}
\item[12]

From this point on we shall explain the root of the
\sepl{srotas}{tube} that are defined for piercing.\footnote{The
    vulgate and the Nepalese version differ significantly on this topic. 
    See Table~\ref{tubecontents}.}  They are the ones responsible for
    carrying breath, food, water, chyle, blood, meat, fat, urine, feces,
    semen and menstrual blood.  Their particulars are many. They have
    their root in the heart, as do the \sepl{dhamanī}{pipe} that carry
    \se{rasa}{chyle}.\footnote{The subject of this sentence is not stated
        in the Nepalese version.  Perhaps this is an old eye-skip because of
        the repetitive sentence structure of this passage.  The vulgate reads
        \dev{prāṇavahe dve} “two vessels carry breath,” at this point.}
        Someone who is pierced there experiences screaming, convulsions, and
        confusion as well as unconsciousness and death.

There are two that carry \se{udaka}{water}.  They have two roots: the 

\begin{table}[t]
    \centering
    \begin{tabular}{>{\it}l>{\it}llll}
        {\rm Nepalese} & {\rm vulgate} &  &  &  \\
        \toprule
        & prāṇa   &  &  &  \\
        & anna    &  &  &  \\
        udaka    & udaka   &  &  &  \\
        & rasa    &  &  &  \\
        rakta    & rakta   &  &  &  \\
        māṃsa    & māṃsa   &  &  &  \\
        medas    & medas   &  &  &  \\
        mūtra    & mūtra   &  &  &  \\
        purīṣa   & purīṣa  &  &  &  \\
        śukra    & śukra   &  &  &  \\
        & ārtava  &  &  &  \\
        \bottomrule
    \end{tabular}
    \caption{Contents of tubes.}
    \label{tubecontents}
\end{table}
% got to here.


\end{translation}

