% !TeX root = incremental_SS_Translation.tex
\chapter{Sūtrasthāna 29: Prognostic signs relating to the messenger 
and to dreams}

\section{Literature}

Meulenbeld offered an annotated overview of this chapter and a bibliography
of earlier scholarship to 2002.\fvolcite{IA}[219--220]{meul-hist} 

\citeauthor{gosw-2011} studied the commentaries of Ḍalhaṇa and 
Cakrapāṇidatta on this and the following adhyāyas up to 32, focussing on the 
topic of \se{ariṣṭa}{omens}.  He concluded that both 
authors were influenced by the Indriyasthāna of the \CS\ in their commentaries 
on this topic.\footcite{gosw-2011}

\citet{lu-2025} discussed the reception of the materials of this
chapter by Chinese Buddhists, especially in the work of the second-
and third-century translators of Saṅgharakṣa's \emph{Yogācārabhūmi},
An Shigao (ca.\ 148--180 \CE) and Dharmarakṣa (fl.\,284 \CE).  As Lu
said, “The Sanskrit text fixes the baseline wording” of the Chinese
translations.\footcite[2]{lu-2025}  This fixes the reception of the
\SS\ in China to the mid- to late second century.\footnote{The \SS\ passages
    directly known to the Chinese translators include \SS\ 1.28.31--32 and 
    1.29.18--19ab.  Note that in \SS\ 1.29.19ab, the “fourth” day (\dev{caturthī})
    is not present in the Nepalese version, but is present in the Chinese 
    receptions of the text and in the vulgate.}
    
\section{Content of the chapter}

The chapter has two distinct halves.  The first describes the 
appearence and behaviour of medical messengers; the second 
describes the meaning of dreams.

\subsection{The messenger}

This part of the chapter was 
discussed by \citet[111--112]{wujad-2012}.

\subsection{Dreams}

The interpretation of dreams, oneiromancy, is the subject of the
second half of this chapter. % [chapter
%29](https://saktumiva.org/wiki/wujastyk/susrutasamhita/01-su.su-1-31/provisional-edition_sutrasthana-1-31?upama_scroll=SS.1.29.1)
This chapter is amongst the earlier writings on this subject in Indian
literature.  Although dream interpretation is evidenced as early as
the \emph{Ṛgveda}\footcite[77, n.\,68]{ping-1981}, the earliest
developed text is probably the two chapters on dream interpretation
included in the \emph{Śardūlakarṇāvadāna}
%(https://panditproject.org/entity/110404/work)*,
which was first fully translated into Chinese in the third century
\CE.\footcites{mukh-1954}[154-161, 161-164]{crui-2025}  It is
interesting that the \emph{Śārdūlakarṇāvadāna} is presented as the
account of a low-caste non-ārya, Triśaṅku, which may hint at a
non-Indian source for the material.  Two chapters on oneiromancy
\emph{Svapnādhyāya} appear in the
\emph{Atharvavedapariśiṣṭa}.\footnote{The text is hard to date;
    perhaps “modern and epic” \citep[767]{keit-1912}.}
    %(https://panditproject.org/entity/113224/work)*
    As Pingree also noted, the earliest full treatise on the topic is the
    twelfth-century
   \emph{Svapnacintāmaṇi}
   %(https://panditproject.org/entity/107707/work)* 
   of
    Jagaddeva,\footcites{nege-1912,sova-2025} while the best known
    is the \emph{Svapnādhyāya} of Bṛhaspāti.
    
        
In the vulgate version of the \SS\ there are often verses that do not
appear in the Nepalese version.  In this chapter of the Nepalese
version, however, it is the other way around: there is a stretch of
ten verses after verse 71 that appear in all three Nepalese witnesses
but that do not appear in the vulgate either here or anywhere else.  In
the digital edition of the Nepalese version we have numbered them
1.29.71.add1 - 1.29.71.add10.\footcite{wuja-2026}
          
At the time of writing, we have not identified any parallels for these verses 
elsewhere in Sanskrit literature.
            
1.29.71.add2 is notable for it's reference to Dhanvantari,
whose literary role in the Nepalese version is quite
different than in the vulgate. \footcite{birc-2021a}


\section{Translation}
    
\begin{translation}    
    \item [1] 
\end{translation}