% !TeX root = incremental_SS_Translation.tex
% Deepro

\chapter[Uttaratantra 65:  Rules of Interpretation]{Uttaratantra 65:  
Rules of Interpretation}

\section{Literature} 

Meulenbeld offered an annotated overview of this chapter and a bibliography
of earlier scholarship to 2002.\fvolcite{IA}[331]{meul-hist} 
Other explorations  of this topic have included \cite{
dasg-1952,
lele-1981,
mejo-2000,
nara-1949,
ober-1968,
scha-1993,
sing-2003,
muth-1976,
comb-1994}. 


 \citet{sche-1981} discussed the term \emph{yukti} in Buddhist
literature; see also \cite[444--446]{biar-1964} \cite[343--345]{pret-1991}, 
while
\cite{frau-1958} discussed the influence of the \emph{tantrayukti}s in the 
Sāṅkhya tradition.
\citet[105--106, fn.\,109]{prei-2013} provided further references to the
discussion of \emph{yukti} in Buddhist literatures. \citet{mane-2008}
gave examples of the use of tantrayuktis in Buddhist commentarial
literature. \citet{chev-2009} discusses the translation of the 
\emph{tantrayukti}s in Tamil literary tradition, with a specific focus on 
\emph{Tolkāppiyam} and its commentaries.

% Meulenbeld HIML IB p.431, note 960 refers to Ruben 1926

\section{Early Sources}

An ancient tradition of enumerating the \emph{tantrayukti}s served as a 
foundational source not only for medical texts but also for works in various other 
disciplines, including Arthaśāstra, philosophy, and even grammar. The \SS\  
stands 
as the earliest Āyurvedic text that presents a compilation of a list of 
\emph{tantrayukti}s followed by their definitions and usage. Mentions to 
Tantrayuktis are also found in the \CS\  \Ca{8.12}{} which introduce four 
additional 
\emph{tantrayukti}s. However, the \emph{tantrayukti}s remain undefined in 
the 
\CS. 

\subsection{The \emph{Arthaśāstra}} 
The enumeration and definitions of the \emph{tantrayukti}s in the 
\SS\  closely parallel their treatment in the \emph{Arthaśāstra}. 
\emph{Tantrayukti}s are discussed in the fifteenth and final chapter of the 
\emph{Arthaśāstra}, called the 
\emph{Tantrayukti}.\footnote{\cite[280--283]{kang-1960}} For a side-by-side 
comparison of the \emph{tantrayukti}s in the \SS\ and the \emph{Arthaśāstra}, 
please refer to Table \ref{table-SAV}.

\subsection{The \emph{Yuktidīpikā}} 
\emph{Yuktidīpikā} (circa late sixth to early eighth century), an anonymous 
commentary on Īśvarakṛṣṇa's \emph{Sāṅkhyakārikā}, initiates its discourse with 
a 
detailed discussion of the characteristics of a scientific treatise, some of which 
align 
with the \emph{tantrayukti}s.\footnote{See \cite[605--614]{ober-1968} for a 
detailed discussion of the use of the \emph{tantrayukti}s in the 
\emph{Yuktidīpikā}.} 
In the \emph{Yuktidīpikā}, these terms are referred to as \emph{tantraguṇa} or 
\emph{tantrasampat}. They are: (1) \emph{sūtropapatti} (2) 
\emph{pramāṇopapatti} (3) \emph{avayavopapatti} (4) \emph{anyūnatā} (5) 
\emph{saṃśayokti} (6) \emph{nirṇayokti} (7) \emph{uddeśa} (8) 
\emph{nirdeśa} (9) \emph{anukrama} (10) \emph{saṃjñā} and (11) 
\emph{upadeśa}.\footnote{
	\begin{verse}
		\dev{sūtrapramāṇāvayavopapattiranyūnatā saṃśayanirṇayoktiḥ/\\
		uddeśanirdeśamanukramaśca saṃjñopadeśāviha tantrasampat//\\}
	\end{verse}
\cite[3]{wezl-1998}}
Apart from these, the \emph{Yuktidīpikā} also exemplifies (12) \emph{utsarga} 
(general 
rule), (13) \emph{apavāda} (exception), and (14) \emph{atideśa} (extended 
application). 
However, \emph{utsarga} and \emph{apavāda} are not considered 
\emph{tantrayukti}s in 
other comprehensive lists. The \emph{Yuktidīpikā} further states that while 
other 
\emph{tantrayukti}s can be demonstrated in a similar manner, since they are 
peripheral 
topics, the text does not delve into their discussion.\footnote{\dev{evamārā 
anye'pi 
drastavyāh/ tadyathotsargo'pavādo'tideśa 
		ityādi/...ityevamanyā api tantrayuktayaḥ śakyā iha pradarśayitum/ 
atiprasaṅgastu prakṛtaṃ tirodadhātīti nivartyate/ siddhaṃ tantrayuktīnāṃ 
sambandhopapattestantram idam iti/}.\cite[8]{wezl-1998}} 

\subsection{Tamil literature}
Discussions on the \emph{tantrayukti}s are also found in Tamil technical 
literature, the earliest of which is the \emph{Tolkāppiyam}.\footnote{For a 
detailed discussion of the treatment of the \emph{tantrayukti}s in the 
\emph{Tolkāppiyam} see \cite{chev-2009}.} 
A list of 32 \emph{tantrayukti}s, called \emph{utti} or \emph{tantiravutti} in 
Tamil, are given in the 27\textsuperscript{th} (the final) chapter titled 
\emph{Marapiyal} “Chapter on conventions” of the last book called 
\emph{Poruḷ} 
“Matters” of the \emph{Tolkāppiyam}. There is no consensus regarding the 
dating 
of the \emph{Tolkāppiyam}. However, if we endorse Zvelebil's view, which posits 
that the final redaction of the \emph{Tolkāppiyam} occurred around the fifth 
century AD, it follows that this section of the \emph{Tolkappiyam} cannot 
postdate 
the fifth century. If we follow the dating of Zvelebil, we can safely argue that by 
that time, Sanskrit \emph{tantrayukti}s had already been translated into Tamil. 
Nevertheless, determining the correspondence between specific 
\emph{tantrayukti}s and Tamil \emph{utti}s poses a challenge. A major factor 
contributing to this challenge is the disagreement between two commentators of 
the \emph{Tolkāppiyam}, namely Iḷampūraṇar (11th or 12th century) and 
Pērāciriyar (possibly 13th century), regarding the interpretation of the list of 
\emph{utti}s. It is still not clear which list of 32 \emph{tantrayukti}s was before 
the author of the \emph{Tolkāppiyam}. 

After the \emph{Tolkāppiyam}, several other Tamil texts refer to the 
\emph{tantrayukti}s. Among them the \emph{Yāpparuṅkalam} (possibly 10th 
century), the \emph{Vīracoḻiyam} (11th century), 
\emph{Naṉṉūl} (late 12th or early 13th century), and 
their commentaries hold significant importance in this context.

\subsection{The \emph{Viṣṇudharmottarapurāṇa}}
The third book of the \emph{Viṣṇudharmattarapurāṇa}, believed to have
been composed between the fifth and seventh centuries, includes a chapter
dedicated to the \emph{tantrayukti}s.\footnote{Adhyāya 6,
    \cite[13--14]{shah-1958}.} Unlike the \emph{Arthaśāstra} and the \SS,
    this chapter lacks illustrative examples of the \emph{tantrayukti}s. The
    chapter lists 32 \emph{tantrayukti}s followed by definitions. Notably,
    the list and definitions given here -- we are using the critical
    edition by Priyabala Shah -- in most cases bear a striking resemblance to
    those found in the \SS. Given the striking alignment between the list and
    definitions of \emph{tantrayukti}s, one could suggest that the
    \emph{Viṣṇudharmottarapurāṇa}'s chapter on \emph{tantrayukti}s likely
    draws directly or indirectly from the \SS\ or from a common
    source. The designations and the order of the \emph{tantrayukti}s in the
    \emph{Viṣṇudharmottarapurāṇa} are almost identical. The only differences
    in the order are as follows:

\begin{enumerate}

	\item \emph{Viparyaya} is placed after \emph{vidhāna} whereas in the \SS\  it 
	follows \emph{arthāpatti}.
	
	\item \emph{Anumata} is placed after \emph{vyākhyāna} whereas in the \SS\ 
it  
	follows \emph{niṛṇaya}.

	\item \emph{Anāgatāvekṣaṇa} (\emph{anāgatāpekṣaṇa} in the Nepalese 
	version) occurs after \emph{atikrāntāvekṣaṇa} (\emph{atikrāntāpekṣaṇa} in 
the 
	Nepalese version) whereas the order is reverse in the \SS.
	
\end{enumerate}

For a side-by-side comparison of the \emph{tantrayukti}s in the \SS\  and the 
\emph{Viṣṇudharmottarapurāṇa}, please refer to Table \ref{table-SAV}.


\subsection{The \emph{Saddanīti}}
A list of the 32 \emph{tantrayukti}s accompanied by definitions also appear in 
the final chapter (\emph{Pariccheda} 28) of the final book (book 3: 
\emph{Suttamālā}) of the renowned Pali grammar \emph{Saddanīti} composed 
by 
Aggavaṃsa in Arimaddanapura (modern Bagan, Burma) in the 
twelfth-century.\footnote{\cite[920--921]{smit-1930}.} Just as the 
\emph{Viṣṇudharmottarapurāṇa}, this list also does not provide examples of the 
\emph{tantrayukti}s. Although written in Pali, the order and the definition of the 
\emph{tantrayukti}s (\emph{tantiyutti} in Pali) closely resemble those of the 
\SS. 
There are, however, a few differences:

	\begin{enumerate}

\item The \emph{tantrayukti} \emph{pradeśa} is referred to as 
\emph{paṭidesa} (Sanskrit \emph{pratideśa}) and is positioned after 
\emph{atidesa} (Sanskrit \emph{atideśa}) whereas in the \SS\ it follows 
\emph{apadeśa}. 

\item \emph{Atikrāntāpekṣaṇa} is designated as \emph{atītāpekkhana} 
(Sanskrit \emph{atītāpekṣaṇa}).

\item \emph{Svasaṃjñā} is designated as \emph{anaññā sakasaṃjñā} (Sanskrit 
\emph{ananyā svasaṃjñā}) and is defined with subtle variations.

\item \emph{Ūhya} is designated as \emph{upānīya}. 
	
	\end{enumerate}

For a side-by-side comparison of the \emph{tantrayukti}s in the \emph{Suśruta 
Saṃhitā} and the \emph{Saddanīti}, please refer to Table 
\ref{table-SAV}.\footnote{%
	For the reading of \emph{Saddanīti}, we used the edition by 
\citeauthor{smit-1930} who also provided an apparatus with variants. However, 
the edition by \citeauthor{ther-1909} sometimes consists of variants which do 
not 
appear in the apparatus of Helmer Smith's edition. In those cases, we noted the 
variants in footnotes.} 




\begin{longtable}{r@{\,}r
		@{\quad\quad}
		m{.25\textwidth} 
		p{.5\textwidth}}
	
	\caption{Tantrayuktis in \SS\  (S), \emph{Viṣṇudharmottarapurāṇa} (V), 
\emph{Arthaśāstra} (A), and \emph{Saddanīti} (N)} 
	
	\label{table-SAV}\\
	\toprule
	\multicolumn{2}{l}{Sequence} & Terms	& Definitions \\
	\midrule
	\endfirsthead
	
	\toprule
	\multicolumn{2}{l}{Sequence} & Terms	& Definitions \\
	\midrule
	\endhead
	
	
	%\rule{0pt}{0.5cm}
	
	(S) & 1. & \emph{adhikaraṇa} & \dev{tatra yamarthamadhikṛtyocyate 
		tadadhikaraṇam/} \\
	(V) & 1. & \emph{adhikaraṇa} & \dev{tatra yamarthamadhikṛtyocyate 
		tadadhikaraṇam/} \\
	(A) & 1. & \emph{adhikaraṇa} & \dev{yamarthamadhikṛtyocyate 
		tadadhikaraṇa/} \\
	(N) & 1. & \emph{adhikaraṇa} & \dev{tattha yaṃ adhikicca vuccati, taṃ 
adhikaraṇaṃ/} \\
	
	\rule{0pt}{0.5cm}(S) & 2. & \emph{yoga} & \dev{yena vākyaṃ yujyate sa 
yogaḥ/ yathā vyatyāsenoktānāṃ sannikṛṣṭaviprakṛṣṭānāṃ padārthānām 
ekīkaraṇam /} \\
	(V) & 2. & \emph{yoga} & \dev{yena vākyārtho yujyate sa yogaḥ/} \\
	(A) & 3. & \emph{yoga} & \dev{vākyayojanā yogaḥ/} \\
	(N) & 2. & \emph{yoga} & \dev{pubbāparavasena vuttānaṃ 
sannihitāsannihitānaṃ padānaṃ ekīkaraṇaṃ yogo;/} \\
	
	\rule{0pt}{0.5cm}(S) & 3. & \emph{padārtha} & \dev{yo'rtho'bhihitaḥ sūtre 
		pade vā sa padārthaḥ/ padasya padayoḥ padānāṃ vā yo'rthaḥ sa padārthaḥ/ 
		aparimitāśca padārthāḥ/} \\
	(V) & 3. & \emph{padārtha} & \dev{yo'rtho vidhikṛtaḥ sūtrapade 
		sa padārthaḥ/} \\
	(A) & 4. & \emph{padārtha} & \dev{padāvadhikaḥ padārthaḥ/} \\
	(N) & 3. & \emph{padattha} & \dev{suttapadesu pubbāparayogato yo attho 
vihito, so padattho/} \\
	
	\rule{0pt}{0.5cm}(S) & 4. & \emph{hetvartha} & \dev{yaduktaṃ sādhanaṃ 
		bhavati sa hetvarthaḥ/} \\
	(V) & 4. & \emph{hetvartha} & \dev{yadanyadyuktimadarthasya sādhanaṃ sa 
hetvarthaḥ/} \\
	(A) & 5. & \emph{hetvartha} & \dev{heturarthasādhako hetvarthaḥ/} \\
	(N) & 4. & \emph{hetuattha} & \dev{yaṃ vuttatthasādhakaṃ, so 
hetuattho/}\footnote{yaṃ vuttaatthasādhanaṃ ? so hetuttho. 
\cite[807]{ther-1909}.} \\
	
	\rule{0pt}{0.5cm}(S) & 5. & \emph{uddeśa / samuddeśa} & 
\dev{samāsavacanaṃ samuddeśaḥ/} \\
	(V) & 5. & \emph{uddeśa} & \dev{samāsavacanamuddeśaḥ/} \\
	(A) & 6. & \emph{uddeśa} & \dev{samāsavākyamuddeśaḥ/} \\
	(N) & 5. & \emph{uddesa} & \dev{samāsavacanaṃ uddeso/} \\
	
	\rule{0pt}{0.5cm}(S) & 6. & \emph{nirdeśa} & \dev{vistaravacanaṃ 
		nirdeśaḥ/} \\
	(V) & 6. & \emph{nirdeśa} & \dev{vistaravacanaṃ	nirdeśaḥ/} \\
	(A) & 7. & \emph{nirdeśa} & \dev{vyāsavākyaṃ nirdeśaḥ/} \\
	(N) & 6. & \emph{niddesa} & \dev{vitthāravacanaṃ niddeso/} \\
	
	\rule{0pt}{0.5cm}(S) & 7. & \emph{upadeśa} & \dev{evamityupadeśaḥ/} \\
	(V) & 7. & \emph{upadeśa} & \dev{evamevetyupadeśaḥ/} \\
	(A) & 8. & \emph{upadeśa} & \dev{evaṃ vartitavyamityupadeśaḥ/} \\
	(N) & 7. & \emph{upadesa} & \dev{evan ti upadeso/} \\
	
	\rule{0pt}{0.5cm}(S) & 8. & \emph{apadeśa} & \dev{anena 
kāraṇenetyapadeśaḥ/} \\
	(V) & 8. & \emph{apadeśa} & \dev{anena kāraṇenetyapadeśaḥ/} \\
	(A) & 9. & \emph{apadeśa} & \dev{evamasāvāhetyapadeśaḥ/} \\
	(N) & 8. & \emph{apadesa} & \dev{anena kāraṇenā ti apadeso/} \\
	
	\rule{0pt}{0.5cm}(S) & 9. & \emph{pradeśa} & \dev{prakṛtasyātikrāntena 
		sādhanaṃ pradeśaḥ/} \\
	(V) & 9. & \emph{pradeśa} & \dev{prakṛtasyānāgatena sādhanaṃ pradeśaḥ/} 
\\
	(A) & 11. & \emph{predeśa} & \dev{vaktavyena sādhanaṃ pradeśaḥ/} \\
	(N) & 10. & \emph{paṭidesa} & \dev{pakatassa anāgatena atthasādhanaṃ 
paṭideso/} \\
	
	\rule{0pt}{0.5cm}(S) & 10. & \emph{atideśa} & \dev{prakṛtasyānāgatena 
sādhanam atideśaḥ/} \\
	(V) & 10. & \emph{atideśa} & \dev{atikramaṇena atideśaḥ/} \\
	(A) & 10. & \emph{atideśa} & \dev{uktena sādhanamatideśaḥ/} \\
	(N) & 9. & \emph{atidesa} & \dev{pakatassa atikkantena sādhanaṃ atideso/} 
\\
	
	\rule{0pt}{0.5cm}(S) & 11. & \emph{apavarga} & 
	\dev{abhipramṛjyāpakarṣaṇamapavargaḥ/} \\
	(V) & 11. & \emph{apavarga} & \dev{abhiprāyānukarṣaṇamapavargaḥ/} \\
	(A) & 22. & \emph{apavarga} & \dev{abhiplutavyapakarṣaṇamapavargaḥ/} \\
	(N) & 11. & \emph{apavagga} & \dev{ativyāpetvā apanayanaṃ apavaggo/} \\
	
	\rule{0pt}{0.5cm}(S) & 12. & \emph{vākyaśeṣa} & \dev{yena 
		padenānuktena vākyaṃ samāpyate sa vākyaśeṣaḥ/} \\
	(V) & 12. & \emph{vākyaśeṣa} & \dev{yenārthaḥ parisamāpyate 
padenāhāryeṇa sa vākyaśeṣaḥ/} \\
	(A) & 17. & \emph{vākyaśeṣa} & \dev{yena vākyaṃ samāpyate sa 
		vākyaśeṣaḥ/} \\
	(N) & 12. & \emph{vākyadosa} & \dev{yena padena avuttena 
vākyaparisamāpanaṃ bhavati, so vākyadoso/} \\
	
	\rule{0pt}{0.5cm}(S) \-\- & \-\- & \-\- \\
	(V) \-\- & \-\- & \-\- \\
	(A) & 12. & \emph{upamāna} & \dev{dṛṣṭenādṛṣṭasya sādhanamupamānam/} 
	\\
	(N) \-\- & \-\- & \-\- \\
	
	\rule{0pt}{0.5cm}(S) & 13. & \emph{arthāpatti} & 
	\dev{yadakīrtitamarthādāpadyate sārthāpattiḥ/} \\
	(V) & 13. & \emph{arthāpatti} & 
	\dev{yadakīrtitamarthādāpadyate sārthāpattiḥ/} \\
	(A) & 13. & \emph{arthāpatti} & \dev{yadanuktamarthādāpadyate 
		sārthāpattiḥ/} \\
	(N) & 13. & \emph{atthāpatti} & 
	\dev{yad akittitaṃ atthato āpajjati, sā atthāpatti/} \\
	
	\rule{0pt}{0.5cm}(S) & 14. & \emph{viparyaya} & \dev{yadyasya 
prātilomyaṃ tadviparyayaḥ/} \\
	(V) & 20. & \emph{viparyaya} & \dev{tasya prātilomyaṃ viparyayaḥ/} \\
	(A) & 16. & \emph{viparyaya} & \dev{pratilomena sādhanaṃ viparyayaḥ/} \\
	(N) & 14. & \emph{vipariyaya} & \dev{yaṃ yattha vihitaṃ, tatra yaṃ tassa 
paṭilomaṃ, so vipariyayo/} \\
	
	\rule{0pt}{0.5cm}(S) & 15. & \emph{prasaṅga} & \dev{prakaraṇāntareṇa 
		samānaḥ prasaṅgaḥ/} \\
	(V) & 14. & \emph{prasaṅga} & \dev{prakaraṇābhihito'rthaḥ 
kenacidupodghātena punarucyamānaḥ prasaṅgaḥ/} \\
	(A) & 15. & \emph{prasaṅga} & \dev{prakaraṇāntareṇa samāno'rthaḥ 
		prasaṅgaḥ/} \\
	(N) & 15. & \emph{pasaṅga} & \dev{pakaraṇantarena 
		samāno attho pasaṅgo/} \\
	
	\rule{0pt}{0.5cm}(S) & 16. & \emph{ekānta} & \dev{yadavadhāraṇenocyate 
		sa ekāntaḥ/} \\
	(V) & 15. & \emph{ekānta} & \dev{yathā tathā sa ekāntaḥ/} \\
	(A) & 26. & \emph{ekānta} & \dev{sarvatrāyattamekāntaḥ/} \\
	(N) & 16. & \emph{ekānta} & \dev{sabbathā yaṃ tathā, so ekānto/} \\
	
	\rule{0pt}{0.5cm}(S) & 17. & \emph{anekānta} & \dev{kvacittathā 
		kvacidanyathā so'nekāntaḥ/} \\
	(V) & 16. & \emph{anekānta} & \dev{kvacittathā 
kvacidanyathā'sāvanekāntaḥ/} \\
	(A)\-\- & \-\- & \-\- \\
	(N) & 17. & \emph{anekānta} & \dev{yo pana katthaci aññathā so anekānto/} 
\\
	
	\rule{0pt}{0.5cm}(S) & 18. & \emph{pūrvapakṣa} & \dev{yastu 
		niḥsaṃśayamabhidhīyate sa pūrvapakṣaḥ/}\footnote{This definition of 
		\emph{pūrvapakṣa} in the Nepalese version is problematic.}\\
	(V) & 17. & \emph{pūrvapakṣa} & \dev{pratiṣedhavacanaṃ pūrvapakṣaḥ/}\\
	(A) & 24. & \emph{pūrvapakṣa} & \dev{pratiṣeddhavyaṃ vākyaṃ 
		pūrvapakṣaḥ/} \\
	(N) & 18. & \emph{pubbapakkha} & \dev{[yo] tu nissandeham abhidhīyate, so 
pubbapakkho/}\\
	
	\rule{0pt}{0.5cm}(S) & 19. & \emph{nirṇaya} & \dev{tasyottaraṃ 
nirṇayaḥ/}\\
	(V) & 18. & \emph{nirṇaya} & \dev{uttaravacanaṃ nirṇayaḥ/}\\
	(A) & 25. & \emph{uttarapakṣa} & \dev{nirṇayavākyamuttarapakṣaḥ/} \\
	(N) & 19. & \emph{niṇṇaya} & \dev{tassa yaṃ uttaraṃ, so niṇṇayo/}\\
	
	\rule{0pt}{0.5cm}(S) & 20. & \emph{anumata} & 
\dev{paramatamapratiṣiddhamanumatam/} \\
	(V) & 25. & \emph{anumata} & \dev{paramatamapratiṣiddhamanumatam/} \\
	(A) & 18. & \emph{anumata} & \dev{paravākyamapratiṣiddhamanumatam/} \\
	(N) & 20. & \emph{anumata} & \dev{paramatam appaṭisiddhaṃ anumataṃ/} 
\\
	
	\rule{0pt}{0.5cm}(S) & 21. & \emph{vidhāna} & 
\dev{prakaraṇānupūrvyādabhihitaṃ vidhānam/} \\
	(V) & 19. & \emph{vidhāna} & \dev{prakaraṇānupūrvaṃ vidhānam/} \\
	(A) & 2. & \emph{vidhāna} & \dev{śāstrasya prakaraṇānupūrvī vidhānam/} \\
	(N) & 21. & \emph{vidhāna} & \dev{pakaraṇānupubbaṃ vidhānaṃ/} \\
	
	\rule{0pt}{0.5cm}(S) & 22. & \emph{anāgatāpekṣaṇa} & \dev{evaṃ 
		vakṣyatītyanāgatāpekṣaṇam/} \\
	(V) & 22. & \emph{anāgatāpekṣaṇa} & \dev{paratra 
vakṣāmītyanāgatāvekṣaṇam/} \\
	(A) & 27. & \emph{anāgatāvekṣaṇa} & \dev{paścādevaṃ 
		vihitamityanāgatāvekṣaṇam/} \\
	(N) & 22. & \emph{anāgatāpekkhana} & \dev{evaṃ vakkhāmi ti 
anāgatāpekkhanaṃ/} \\
	
	\rule{0pt}{0.5cm}(S) & 23. & \emph{atikrāntāpekṣaṇa} & 
	\dev{ityuktamityatikrāntāpekṣaṇam/} \\
	(V) & 21. & \emph{atikrāntāpekṣaṇa} & \dev{ityuktamatikrāntāvekṣaṇam/} \\
	(A) & 28. & \emph{atikrāntāvekṣaṇa} & \dev{purastādevaṃ 
vihitamityatikrāntāvekṣaṇam/} \\
	(N) & 23. & \emph{atītāpekkhana} & \dev{iti vuttan ti atītāpekkhanaṃ/} \\
	
	\rule{0pt}{0.5cm}(S) & 24. & \emph{saṃśaya} & 
\dev{ubhayahetunidarśanaṃ saṃśayaḥ/} \\
	(V) & 23. & \emph{saṃśaya} & \dev{ubhayato hetudarśanaṃ saṃśayaḥ/} \\
	(A) & 14. & \emph{saṃśaya} & \dev{ubhayato hetumānarthaḥ saṃśayaḥ/} \\
	(N) & 24. & \emph{saṃsaya} & \dev{ubhayahetudassanaṃ saṃsayo/} \\
	
	\rule{0pt}{0.5cm}(S) & 25. & \emph{vyākhyāna} & 
\dev{tatrātiśayopavarṇanaṃ vyākhyānam/} \\
	(V) & 24. & \emph{vyākhyāna} & \dev{tatrātiśayavarṇanātivyākhyānam/} \\
	(A) & 19. & \emph{vyākhyāna} & \dev{atiśayavarṇanā vyākhyānam/} \\
	(N) & 25. & \emph{vyākhyāna} & \dev{saṃvaṇṇanā vyākhyānam/} \\
	
	\rule{0pt}{0.5cm}(S) & 26. & \emph{svasaṃjñā} & \dev{anyaśāstrāsāmānyā 
		svasaṃjñā/} \\
	(V) & 26. & \emph{svasaṃjñā} & \dev{parairasammataḥ śabdaḥ svasaṃjñā/} \\
	(A) & 23. & \emph{svasaṃjñā} & \dev{parairasamitaḥ śabdaḥ svasaṃjñā/} \\
	(N) & 26. & \emph{anaññā sakasaññā} & \dev{bhūtānaṃ pavattā 
ārambhacintā anaññā, sassa sādhāraṇā sakasaññā/} \\
	
	\rule{0pt}{0.5cm}(S) & 27. & \emph{nirvacana} & 
\dev{lokaprathitamudāharaṇaṃ nirvacanam/} \\
	(V) & 27. & \emph{nirvacana} & \dev{loke pratītamudāharaṇaṃ nirvacanam/} 
\\
	(A) & 20. & \emph{nirvacana} & \dev{guṇataḥ śabdaniṣpattirnirvacanam/} \\
	(N) & 27. & \emph{nibbacana} & \dev{lokappatītam udāharaṇaṃ 
nibbacanaṃ/} \\
	
	\rule{0pt}{0.5cm}(S) & 28. & \emph{nidarśana} & 
\dev{dṛṣṭāntavyaktirnidarśanam/} \\
	(V) & 28. & \emph{nidarśana} & \dev{tadyuktinidarśanaṃ dṛṣṭāntaḥ/} \\
	(A) & 21. & \emph{nidarśana} & \dev{dṛṣṭānto dṛṣṭāntayukto nidarśanam/} \\
	(N) & 28. & \emph{nidassana} & \dev{diṭṭhantasaṃyogo nidassanaṃ/} \\
	
	\rule{0pt}{0.5cm}(S) & 29. & \emph{niyoga} & \dev{idameveti niyogaḥ/} \\
	(V) & 29. & \emph{niyoga} & \dev{eveti niyogaḥ/} \\
	(A) & 29. & \emph{niyoga} & \dev{evaṃ nānyatheti niyogaḥ/} \\
	(N) & 29. & \emph{niyoga} & \dev{idam evā ti niyogo/} \\
	
	\rule{0pt}{0.5cm}(S) & 30. & \emph{vikalpa} & \dev{/} \\
	(V) & 30. & \emph{vikalpa} & \dev{idaṃ vedaṃ veti vikalpaḥ/} \\
	(A) & 30. & \emph{vikalpa} & \dev{anena vānena veti vikalpaḥ/} \\
	(N) & 30. & \emph{vikappa} & \dev{idaṃ vā ti vikappo/} \\
	
	\rule{0pt}{0.5cm}(S) & 31. & \emph{samuccaya} & \dev{/} \\
	(V) & 31. & \emph{samuccaya} & \dev{idaṃ cedaṃ ceti samuccayaḥ/} \\
	(A) & 31. & \emph{samuccaya} & \dev{anena cānena ceti samuccayaḥ/} \\
	(N) & 31. & \emph{samuccaya} & \dev{saṃkhepavacanaṃ samuccayo/} \\
	
	\rule{0pt}{0.5cm}(S) & 32. & \emph{ūhya} & \dev{yadanirdiṣṭaṃ 
		buddhigamyaṃ tadūhyam/} \\
	(V) & 32. & \emph{ūhya} & \dev{atra yadanirdiṣṭaṃ yuktigamyaṃ 
tadūhyam/} \\
	(A) & & \emph{ūhya} & \dev{anuktakaraṇamūhyam/} \\
	(N) & 32. & \emph{upānīya} & \dev{yad aniddiṭṭhaṃ buddhiyā 
avagamanīyaṃ, tad upānīyan ti/} \\
	
	\bottomrule
	
\end{longtable}


\subsection{Āyurvedic literature}

\subsubsection{Primary texts}
While references to \emph{tantrayukti}s can be found across various disciplines, 
Āyurveda places a particular emphasis on their discussion, 
especially evident in key texts of Āyurveda, 
such as the \emph{Caraka-} and the \emph{Suśruta- saṃhitā}s, 
as well as the \AS. The \emph{Carakasaṃhitā} and \AS\ present an identical list 
of \emph{tantrayukti}s contained in a stanza of four \emph{anuṣṭubh} 
verses.\footnote{\label{CaAsT}
	\begin{verse}
		\dev{tatrādhikaraṇaṃ yogo hetvartho'rthaḥ padasya ca/ \\
		pradeśoddeśanirdeśavākyaśeṣāḥ prayojanam// \\
		upadeśāpadeśātideśārthāpattinirṇayāḥ/ \\
		prasaṅgaikāntanaikāntāḥ sāpavargo viparyayaḥ// \\
	 	pūrvapakṣavidhānānumatavyākhyānasaṃśayāḥ/ \\
 		atītānāgatāpekṣāsvasaṃjñohyasamuccayāḥ// \\
 		nidarśanaṃ nirvacanaṃ niyogo'tha vikalpanam/ \\
 		pratyutsārastathoddhāraḥ sambhavastantrayuktayaḥ// \\}
	\end{verse}
	\AS\ \As{6.50.150--153}{959}. \emph{Carakasaṃhitā} 
\Ca{8.12.41b--45a}{736} reads almost the same. The only two variants are (1) 
\dev{atītānāgatāvekṣā...} and (2) \dev{nirvacanaṃ saṃniyogo vikalpanam}.
	} 
However, unlike the \SS\ they lack explicit definitions and examples. 
This list of the \emph{tantrayukti}s appear in the final chapter of the last book 
in both \emph{Carakasaṃhitā} (41b--45a, chapter 12, \emph{Siddhisthāna}) 
and 
\emph{Aṣṭāṅgasaṅgraha} (150--153, chapter 50, \emph{Uttarasthāna}). 
The same has been quoted by Aruṇadatta in his commentary 
\emph{Sarvāṅgasundarī} on the \emph{Aṣṭāṅgahṛdaya} 
while elucidating the concept of \emph{tantraguṇa} (qualities of the 
system)\footnote{Aruṇadatta on the \AHS\ \Ah{6.40.78}{946}.}  
and by Śrīdāsa Paṇḍita in the prefatory section of his commentary 
\emph{Hṛdayabodhikā} on the \AHS.\footcite[1--2]{muss-1940}
Notably, this list consists of 36 \emph{tantrayukti}s instead of 32 found in the 
\SS\ and other texts. The additional four are: \emph{prayojana} (objective), 
\emph{pratyutsāra} (rebuttal), \emph{uddhāra}, and \emph{sambhava} 
(origin). 

The presence of identical verses enumerating the \emph{tantrayukti}s in 
the \AHS, \CS, \emph{Sarvāṅgasundarī} and \emph{Hṛdayabodhikā} 
strongly suggests a shared origin.
 However, a critical issue arises due to 
 the absence of a comprehensive critical edition 
 of the chapter 12 of the \emph{Siddhisthāna} of the \CS, 
 leaving uncertainty about the total number of \emph{tantrayukti}s
  recognized by Dṛḍhabala in this section.\footnote{We know from 
  	internal textual evidence that the \emph{Siddhisthāna} of the \CS\ 
  	in which the list of the \emph{tantrayukti}s appear was 
  	originally authored by Dṛḍhabala, who lived in a town called Pañcanada 
sometime between 300 and 500 AD. \\Cf. 
	\begin{verse}
		\dev{akhaṇḍārthaṃ dṛḍhabalo jātaḥ pañcanade pure/ \\
		kṛtvā bahubhyastantrebhyo viśeṣoñchaśiloccayam// \\
		saptadaśauṣadhādhyāyasiddhikalpairapūrayat/} \\
	\end{verse}
	\Ca{8.12.39--40a}{735}
	} 
The problem arises from different readings of the half-verse that occurs right 
before the list of 36 \emph{tantrayukti}s. In \MScite{Kathmandu NAK 1/1648} 
(dated 1183 AD, the oldest dated manuscript of the \CS\ known to us), the 
reading 
of this verse is:	
			\dev{\textbf{ṣaṭtriṃśadbhi}rvicitrābhirbhū[ṣi]taṃ tantrayuktibhiḥ//}	
This number of 36 \emph{tantrayukti}s perfectly agrees with the following list of 
the 36 \emph{tantrayukti}s. A similar reading is found in Trikamji's 1933 
\CS\ edition which contains only the 
\emph{mūla}-text.\footnote{\dev{\textbf{ṣaṭtriṃśatā} vicitrābhirbhūṣitaṃ 
tantrayuktibhiḥ//}		
	8.12.70a \parencite[972]{cara-trikamji}.} 
However, although most of the other editions consist of the same reading, a 
number of editions show quite a lot of discrepancies with the number. For 
example, 
Trikamji's 1941 edition of the \CS\ reads the same half-verse as 
\dev{\textbf{ṣaḍviṃśatā} vicitrābhirbhūṣitaṃ tantrayuktibhiḥ/} 
\Ca{8.12.41a}{735}. 
In the same edition, the reading of Cakrapāṇi's \emph{Āyurvedadīpikā} supports 
the reading: 
\dev{\textbf{ṣaḍviṃśa}ttantrayuktibhirbhūṣitamapūrayaddṛḍhabala iti yojanā}. 
However, after this verse, the same edition consists of the versified list of the 36 
\emph{tantrayukti}s and commenting on these verses, the 
\emph{Āyurvedadīpikā} confirms the total number of the \emph{tantrayukti}s 
as 
36: 
\dev{ityetāḥ \textbf{ṣaṭtriṃśa}ttantrayuktayo 
vyāhṛtāḥ/}.\footcite[737]{cara-trikamji3} 
Moreover, the edition of Rāmaprasāda Vaidyopādhyāya reads the half-verse as--- 
\dev{pañcatriṃśadvicitrābhirbhūṣitaṃ 
tantrayuktibhiḥ}.\footnote{\cite[1913]{vaid-1911}.} 
Rāmaprasāda Vaidyopādhyāya excludes \emph{ūhya}.\footnote{Understanding 
the \emph{tantrayukti} \emph{samuccaya} as \emph{asamuccaya}, he reads 
the 
verse where \emph{ūhya} appears as--- 
	\dev{atītānāgatāpekṣā svasaṃjñā hyasamuccayāḥ}. Surely, this reading is 
erroneous as the plural ending after \emph{samuccaya} does not make sense.} 
The same reading is found in Satīśacandra Śarmā's third edition of the 
\CS.\footnote{\cite[1020]{sarm-1923}. His first edition, however, reads the 
half-verse the same as the reading in 
\cite{cara-trikamji}.\parencite[884]{sarm-1904}}
However, adding more troubles to it, Satīśacandra Śarmā, in his Bengali 
translation, 
says that there are 34 \emph{tantrayukti}s (even though the main Sanskrit text 
of 
his edition counts 35). Then he in fact illustrates 36 \emph{tantrayukti}s making 
a 
remark that states---

\begin{quote}
	
 “in Gaṅgadhara's reading, there are 36 \emph{tantrayukti}s because he 
counts \emph{saṃśaya} twice in his commentary. But 35 was reckoned in his 
\emph{mūla}-text. Another manuscript reckons 34 \emph{tantrayukti}s 
excluding 
\emph{apadeśa}. This edition reads thirty-five instead of thirty-four or 
thirty-six.”\footnote{\textbengali{“গঙ্গাধর পাঠ— তন্ত্রযুক্তি ছত্রিশ প্রকার। তিনি টীকাতে সংশয়কে 
দুই বার উল্লেখ করিয়া ছত্রিশ প্রকার গণনা করিয়াছেন, কিন্তু তাঁহার মূলে পঁয়ত্রিশ প্রকার আছে; গ্রন্থান্তরে ৩৪ 
প্রকার আছে; তাহাতে ‘অপদেশ’ ধর্ত্তব্য হয় নাই। এই অনুবাদের মূলে চতুস্ত্রিংশৎ বা ষট্‌ত্রিংশৎ স্থলে 
পঞ্চত্রিংশৎ লিখিত হইল।”} \cite[1022]{sarm-1923}.}
	
\end{quote}

 
In the edition of Narendranātha Senagupta and Balāicandra Senagupta that 
includes 
Cakrapāṇi's \emph{Āyurvedadīpikā} and Gaṅgādhara's \emph{Jalpakalpataru}, 
the 
Sanskrit \emph{mūla} and the \emph{Jalpakalpataru} enumerate 36 
\emph{tantrayukti}s. However, in the same edition, the \emph{Āyurvedadīpikā} 
reads, \dev{\textbf{pañcatriṃśa}ttantrayuktibhirbhūṣitamapūrayaddṛḍhabala iti 
yojanā}.\fvolcite{III}[3814]{sena-1928} 
Again, after the illustrations of the 36 \emph{tantrayukti}s it reads, \dev{ityetāḥ 
\textbf{ṣaṭtriṃśa}ttantrayuktayo vyāhṛtāḥ}.\fvolcite{III}[3822]{sena-1928} 
In his edition of the \emph{Tantrayuktivicāra}, Muthuswami also mentions that 
35 \emph{tantrayukti}s are reckoned in the 
\CS.\footnote{\dev{‘\textbf{pañcatriṃśa}dvicitrābhirbhūṣitaṃ 
tantrayuktibhiḥ/}’ \dev{iti carake/ dvātriṃśaditi suśrutaḥ/}\parencite[2, 
fn.\,2]{muth-1976}.} 
Jivānanda Vidyāsagara's edition gives no number at all--- \dev{tathā ca tā 
vicitrābhirbhūṣitaṃ tantrayuktibhiḥ}.\footcite[961]{bhat-1877} 

Commentaries on the \CS\ prior to Cakrapāṇi's \emph{Āyurvedadīpikā}, such as 
the 
\emph{Carakanyāsa} of Bhaṭṭāra Hariścandra (c. mid-sixth century) or 
\emph{Nirantarapadavyākhyā} of Jejjaṭa (c. 7th or 8th century AD) do not help 
much because the extant portions of these commentaries do not include the 
concerned section of the 12th chapter of the \emph{Siddhisthāna}. However, 
Hariścandra was possibly not aware of the total 
number and the list of the \emph{tantrayukti}s in the final chapter of the 
\emph{Siddhisthāna} because he discussed the \emph{tantrayukti}s right at 
the 
beginning of his commentary and showed no indication to the awareness about 
the 
discussion on the \emph{tantrayukti}s at the end of the text. Moreover, he 
discusses 40 \emph{tantrayukti}s instead of 36. It is not yet settled whether or 
not 
Hariścandra was aware of Dṛḍhabala's redaction of the \CS. However, 
Hariścandra's 
treatment of the \emph{tantrayukti}s supports the 
latter.\fvolcite{IA}[189]{meul-hist}.  It 
is clear from Cakrapāṇi's commentary on the \CS\ that in the version of the text 
he 
commented upon contained the four verses that list the 36 \emph{tantrayukti}s. 
It 
is, however, not improbable that the four verses that list the 36 
\emph{tantrayukti}s were later added to the \CS\ sometime between the sixth 
(the 
date of Hariścandra) and the eleventh century (the date of Cakrapāṇi) and the 
discrepancy appeared when the previous verse that gives the total number of the 
\emph{tantrayukti}s was not properly emended by the scribes complying with 
the 
following list of 36 \emph{tantrayukti}s. There is a need of a critical edition of 
the 
twelfth chapter of the \emph{Siddhisthāna} of the \CS\ to address these issues 
definitely.  

\subsubsection{Commentaries}\label{commentaries}

The commentators who extensively delved into the discussion of the 
\emph{tantrayukti}s are Hariścandra, the author of \emph{Carakanyāsa}, and 
Aruṇadatta (12th century),\fvolcite{IA}[663--664]{meul-hist}
who authored his commentary \emph{Sarvāṅgasundarī} on the 
\emph{Aṣṭāṅgahṛdaya} of Vāgbhaṭa. Hariścandra meticulously defined and 
analyzed 40 \emph{tantrayukti}s at the beginning of his work. The four 
additional 
\emph{tantrayukti}s are:  \emph{paripraśna} (question), \emph{vyākaraṇa} 
(grammatical clarification), \emph{vyutkrāntābhidhāna} (overpassing 
statement) 
and \emph{hetu} (means of knowledge).\footnote{This text has only been 
published once 
	(only until the third chapter of \emph{Sūtrasthāna}) by Masta Ram Shastri 
from Lahore in 1932/33.\fvolcite{IB}[290]{meul-hist} Unfortunately, it is 
currently 
inaccessible to us. Although some fragmented manuscripts of the 
\emph{Carakanyāsa} exist, for this section (Chapter 1, \emph{Sūtrasthāna}), we 
were able to consult only \MScite{MS Jamnagar GAU 114}. This is a recent 
apograph with several lacunae and corruptions. The list of the tantrayuktis 
provided in the \emph{Carakanyāsa} is as follows (with some emendations made 
in the reading): \dev{tantrasya yuktayo'dhikaraṇādyāścatvāriṃśat/... 
yuktayastāvadadhikaraṇaṃ yogo hetvartha uddeśo [nirdeśa] 
upadeśo'padeśo'tideśaḥ pradeśo nirṇayo'rthāpattirvākyaśeṣaḥ prayojanaṃ 
prasaṅga ekānto'nekānto viparyayo'pavargaḥ pūrvapakṣo vidhānamanumataṃ 
vyākhyānaṃ paripraśno vyākaraṇamatītāpekṣaṇamanāgatāpekṣaṇaṃ saṃśayaḥ 
svasaṃjñohyaḥ samuccayo nidarśanaṃ nirvacanaṃ niyogo vikalpaḥ pratyutsāra 
uddhāraḥ sambhavo vyutkrāntābhidhānaṃ heturiti/}}

Aruṇadatta, while discussing the concept of \emph{tantraguṇa} at the end of the 
\emph{Aṣṭāṅgahṛdaya}, 
provided an elaborate description of \emph{tantrayukti}s, 
considering them as part of a system of ninety-five \emph{tantraguṇa}s. 
Śrīdāsa Paṇḍita (14th century), a commentator on the \emph{Aṣṭāṅgahṛdaya}, 
echoed Aruṇadatta's exploration of \emph{tantrayukti}s 
in the beginning of his commentary, 
\emph{Hṛdayabodhikā}.\fvolcite{IA}[680]{meul-hist} 
Thus, both Hariścandra and Śrīdāsa Paṇḍita engage with this topic right at the 
beginning, 
underscoring the significance they attribute to the subject. 
Other noteworthy commentators who discussed the topic of \emph{tantrayukti} 
are Cakrapāṇi (11th century) 
and Indu (sometime between 8th and 12th century). 
Cakrapāṇi and Indu defined and illustrated the \emph{tantrayukti}s 
mentioned in the \emph{Carakasaṃhitā} and the \emph{Aṣṭāṅgasaṅgraha}, 
respectively. 
They affirm the inclusion of the four additional \emph{tantrayukti}s in 
Hariścandra's list. 
Cakrapāṇi, aligning them with existing concepts, 
incorporates \emph{paripraśna}, \emph{vyākaraṇa}, 
and \emph{vyutkrāntābhidhāna} under the \emph{tantrayukti}s 
\emph{uddeśa}, 
\emph{vyākhyāna}, and \emph{nirdeśa}, respectively. 
According to him, \emph{hetu} serves as an overarching term 
encompassing all \emph{pramāṇa}s (means of knowledge) 
such as \emph{pratyakṣa} (perception) and others. 
Indu, however, outlines three possible reasons 
for not incorporating these \emph{tantrayukti}s into the list: 
(1) they lack direct mention in the main text, 
(2) they could be considered as falling within the scopes of already enumerated 
\emph{tantrayukti}s, or 
(3) they are not recognized as \emph{tantrayukti}s. 

Detailed discussions on the \emph{tantrayukti}s also appear 
in the \emph{Jalpakalpataru}, a nineteenth-century commentary 
on the \emph{Carakasaṃhitā} by Gaṅgādhara Kavirāja from Bengal.
Gaṅgādhara included the commentary with his \emph{editio princeps} of the 
\CS. 
He defines the \emph{tantrayukti}s most often as defined in the \SS\ 
making explicit quotations from the \SS\ itself.
Hence, this commentary serves as a testimonium for most part 
of the \emph{tantrayukti} section of the \SS. 

\subsubsection{Monographs}
Two texts authored by Āyurvedic scholars exclusively delve into the topic of 
\emph{tantrayukti}. The first is the \emph{Tantrayuktivicāra} by a physician 
named Nīlamegha (also known as Vaidyanātha), while the second is called the 
\emph{Tantrayukti}, which is a sort of recast of the former by an anonymous 
author. The anonymous author describes himself as being from the same lineage 
as 
Nīlamegha and asserts that Nīlamegha belongs to the same lineage of Bhiṣagārya 
(also known as Nārāyaṇa Bhiṣaj). Both Nīlamegha and the author of Tantrayukti 
are 
likely from Kerala or coastal Karnataka.\footnote{Kolatteri Saṅkaramenon, 
	the first editor of the \emph{Tantrayuktivicāra}, believes that Nīlamegha hails 
from Kerala. This conclusion is drawn from Nīlamegha's reference to his guru as 
Sundara, whom Saṅkaramenon identifies as the same individual credited with 
composing the \emph{Lakṣaṇāmṛta}, a treatise on toxicology. This assertion is 
plausible because the only known manuscript of \emph{Tantrayuktivicāra} 
belongs to a member of one of the Aṣṭavaidya families of Kerala, aligning with the 
Vāgbhaṭa school, to which Nīlamegha also 
belongs.(\fvolcite{IIA}[143]{meul-hist}) 
On the other hand, the anonymous author of the \emph{Tantrayukti} associates 
Nīlamegha with the lineage of Bhiṣagārya, who hails from Uṇṭuru, a village 
located 
3 kilometers from Gokarṇa which is in coastal 
Karnataka.\parencite[30]{nara-1949}.}
According to Koḷatteri Śaṅkaramenon and Meulenbeld, Nīlamegha flourished in 
the first half of ninth century.\footnote{Nīlamegha mentions Vāhaṭa (Vāgbhaṭa), 
Indu, and Jejjaṭa in his work.
	 This places him definitively after the seventh century. The Buddhist influence 
in the Tantrayukti indicates a date not much later than 800 AD. 
(\cite[\dev{avatārikā 5–6}]{muth-1976}, \volcite{IIA}[143]{meul-hist}.)} 
The \emph{Tantrayukti} was very likely composed after the sixteenth 
century.\footnote{From 
	the explicit mention of Nīlamegha and Bhiṣagārya in the work 
\emph{Tantrayukti}, we can say that the author flourished after them. 
Determining 
the date of Bhiṣagārya is problematic. However, since the Kairalī commentary on 
the \AHS\ frequently quotes from Bhiṣagārya's \emph{Abhidhānamañjarī}, it 
indicates that Bhiṣagārya predates the composition of this commentary. 
Meulenbeld suggests the end of the seventeenth century as the terminus post 
quem for the Kairalī (\volcite{IA}[675]{meul-hist}). Moreover, he views 
\emph{Abhidhānamañjarī} as a work composed after the sixteenth century, 
citing 
details within it that affirm its posteriority to the \emph{Rājanighaṇṭu} and 
\emph{Bhāvaprakāśa} (\volcite{IIA}[442]{meul-hist}).} 

Nīlamegha's \emph{Tantrayuktivicāra} is a versified text accompanied by an 
autocommentary. The text comprises eighteen verses plus a hemistich, resulting 
in 
a total of 37 hemistichs. Each hemistich serves as a definition for a 
\emph{tantrayukti}. Nīlamegha enumerates a total of 36 \emph{tantrayukti}s, 
as 
mentioned in the \AS\ and \emph{Carakasaṃhitā}. The additional hemistich 
defines \emph{aviparyaya}, which, according to Nīlamegha, is sometimes 
considered instead of \emph{viparyaya}. This substitution occurs when one 
understands that the negative prefix \emph{a-} is deleted due to a 
\emph{pūrvarūpa sandhi}--- \emph{sāpavargaḥ + aviparyayaḥ} → 
\emph{sāpavargo viparyayaḥ} (See footnote \ref{CaAsT}.).

The text of the \emph{Tantrayukti} includes some verses at the beginning and 
end, where the author discusses the lineage of Nīlamegha. The author explicitly 
states that his text is a revised version of Nīlamegha's \emph{Tantrayuktivicāra} 
because the available manuscripts were mostly 
corrupt.\footnote{\dev{vaidyanāthopasṛṣṭānāṃ lakṣyalakṣaṇavāgjuṣām//\\
	tāsāṃ prāyaḥ prakāśānāṃ durlekhāpaṅkadūṣaṇāt/\\
	kriyate sāmprataṃ kṛcchrāduddhṛtya parimārjjanam//\\}
	\cite[1]{nara-1949}} 
It is evident that there are substantial reproductions of parts of the 
\emph{Tantrayuktivicāra} and its autocommentary. The total number of 
\emph{tantrayukti}s and their enumeration remains identical to that of the 
\emph{Tantrayuktivicāra}. What distinguishes it from the 
\emph{Tantrayuktivicāra} is the incorporation of a list of other 
\emph{tantraguṇa}s and 14 \emph{tantradoṣa}s. This list of 
\emph{tantraguṇa}s 
includes 15 types of \emph{vyākhyā}, 7 types of \emph{kalpanā}, 20 types of 
\emph{āśraya}, and 17 types of metaphoric and metonymic devices, such as 
\emph{tācchīlya} and so on.

\section{\emph{Tantrayukti}-inventories}

It is evident from the discussion on the early sources that all these listings of the 
\emph{tantrayukti}s in the early sources can be grouped into two categories. For 
the ease of our following discussion, we name these two inventories as (1) earlier 
listing and (2) later listing.

\subsection{Earlier Listing}

The four inventories of \emph{tantrayukti}s from the \emph{Arthaśāstra}, \SS, 
\emph{Viṣṇudharmottarapurāṇa}, and \emph{Saddanīti} belong to what we call 
the “earlier Listing.” The reason to call this listing as “earlier listing” is: two early 
Sanskrit texts, viz., the \emph{Arthaśāstra} and the \SS\ consists of this listing. 
The 
Tamil list of the \emph{tantrayukti}s as found in the \emph{Tolkāppiyam} also 
belongs to this group, even though not all of the \emph{utti}s in this list might 
correspond accurately to the Sanskrit and Pali lists. A defining characteristic of 
this 
listing is that each inventory explicitly states the total number of 
\emph{tantrayukti}s as thirty-two.\footnote{
	\dev{tad dvātriṃśad yuktiyuktam/} (\volcite[5.1.3]{1}[280]{kang-1969}) 
“that (\emph{Arthaśāstra}) is furnished with thirty-two logical methods of the 
system”,
	\dev{tatra dvātriṃśat tantrayuktayo bhavanti/} (\SS 6.64.2) “there are 
thirty-two logical methods of the system”,
	battiṃsa tantiyuttiyo bhavanti/ \parencite[\emph{Suttamālā}, 
28][920]{smit-1930} “there are thirty-two logical methods of the system”,		
	\texttamil{எண்ணான்கு உத்தியின்}... “It employs thirtytwo rules of criticism 
regarding writing.” \parencite[9--10]{srip-1995}} 
Even though there are sometimes different \emph{tantrayukti}s enumerated in 
different lists, the total count always remains consistent at 32. As demonstrated 
in 
Table \ref{table-SAV}, the Sanskrit and Pali lists are similarly ordered and are 
always accompanied by similar or identical definitions. This list appears across 
diverse disciplines.

\subsection{Later Listing}

The “later listing” is the one we find in the \AS, \CS, the commentaries on the \CS, 
\AS\ and \AHS\ and the two monographs, the \emph{Tantrayuktivicāra} and 
\emph{Tantrayukti}. This list has sprung from a single source--- a versified list of 
thirty-six \emph{tantrayukti}s comprising four verses that appear in the \AS, \CS\ 
and Aruṇadatta's commentary on the \AHS. It remains unclear whether these 
verses initially appeared in the Dṛḍhabala's redaction of the \CS\ or Vāgbhaṭa's 
\AS. 
Unlike the “earlier Listing,” this list lacks definitions of the \emph{tantrayukti}s. 
Definitions and illustrations are given by the authors of the commentaries and 
monographs as discussed in the previous section. Although Hariścandra's list 
includes 40 \emph{tantrayukti}s instead of 36, his enumeration aligns more 
closely with “later listing” than the earlier one. Despite the earlier listing's 
corss-disciplinary appearance, the later listing notably influences the field of 
Āyurveda, likely due to the popularity of Vāgbhaṭa's works. The \SS, 
incorporating 
the “earlier listing” distinguishes itself among Āyurvedic texts that list the 
\emph{tantrayukti}s.  

\section{Terminology}

\label{tantra-trans}
The terms have been translated into English in numerous books and articles. 
English renditions of the terms can be found in English translations of the \SS\ 
such 
as in \cite[171--172]{sing-1980}, and \volcite{3}[631--639]{shar-1999}; in 
translations of the \emph{Carakasaṃhitā} such as in \cite[436--444]{shar-2006} 
and in \cite[1050]{gula-1949}, in the translation of the \emph{Arthaśāstra} 
such 
as in \cite[459]{sham-1951}, \cite[593]{kang-1969}, \cite[1103]{unni-2006} 
and 
\cite[]{oliv-2013}, and by K. Srikanta Moorthy in \cite[Appendix 
xi--xxxiv]{muth-1976}. They are also found in various books and articles 
dedicated 
to discussing the \emph{tantrayukti}s such as in \cite[601--602]{ober-1968}, 
\volcite{1}[72]{solo-1976}, \cite[34--155]{lele-1981}, 
\citeyear[36--150]{lele-2006} and so on. German translations of the terms can 
be 
found in \cite[663--664]{meye-1926} (German translation of the 
\emph{Arthaśāstra}) and in \cite{pret-1991}.


The definitions of \emph{tantrayukti}s exhibit numerous variations across 
different texts. Here we will discuss each of the \emph{tantrayukti}s that occur 
in 
the \emph{Suśruta Saṃhitā} in comparison with their definitions in other texts. 
As 
indicated in Table \ref{table-SAV}, the definitions of \emph{tantrayukti}s in the 
\SS\ are frequently either identical or nearly identical to those found in the 
\emph{Arthaśāstra}, \emph{Viṣṇudharmottarapurāṇa} and \emph{Saddanīti}. 
Therefore, unless the definitions in these two texts notably deviate from those in 
the \SS, we will not make explicit references to them in the subsequent 
elucidation 
of the terms. 


\subsection{1. \emph{adhikaraṇa}}

\emph{Adhikaraṇa} appears as the first \emph{tantrayukti} in all traditional 
enumerations. It is among those \emph{tantrayukti}s for which there is little 
disagreement concerning its definition. This tantrayukti functions as a 
structural and interpretative device. With a tautological expression, the \SS\ 
defines \emph{adhikaraṇa} as something, with reference to which statements 
are made. While defining \emph{adhikaraṇa}, the text employs the same verb, 
\emph{adhi- kṛ-} (to refer), whence the noun \emph{adhikaraṇa} has been 
derived. The text supplies examples of \emph{rasa} (taste) and \emph{doṣa} 
(humour), for which two chapters of the \emph{Uttaratantra}, namely chapter 
62 (\emph{Kāyacikitsā} 27) and  chapter 65 (\emph{Kāyacikitsā} 30) are 
dedicated.\footnote{They are chapters 63 and 66 in \Su{A}{B}.} Clearly, 
\emph{adhikaraṇa} is the topic or theme. 

Cakrapāṇi and Gaṅgādhara define \emph{adhikaraṇa} in almost the same way 
as does the \SS.\footnote{\dev{yamarthamadhikṛtya pravartate kartā/ yathā 
“vighnabhūtā yadā rogā”
	ityādi/ atra rogādikamadhikṛtyāyurvedo maharṣibhiḥ kṛta iti ‘rogāḥ}’ 
\dev{ityadhikaraṇam/} \emph{Āyurvedadīpikā} 
\parencite[736]{cara-trikamji3}. 
\dev{tad yamarthamadhikṛtyocyate;} \emph{Jalpakalpataru} 
(\volcite{III}[3815]{sena-1928}).}
Aruṇadatta's definition is similar but he specifies that \emph{adhikaraṇa} can be 
of an entire discipline \emph{(śāstra)}, or a book \emph{(sthāna)} of it, or a 
chapter \emph{(adhyāya)}, or a section \emph{(prakaraṇa)}, or even of a 
sentence \emph{(vākya)}.\footnote{\dev{tatra 
	adhikaraṇaṃ nāma, yadadhikṛtya pravartate śāstraṃ sthānamadhyāyaṃ 
prakaraṇaṃ vākyaṃ vā/...} \parencite[947]{kunt-1939}.}
Śrīdāsa Paṇḍita follows Aruṇadatta.\footcite[2]{muss-1940} 
However, in the commentaries of Hariścandra and Indu, we explore two more 
aspects of the concept of \emph{adhikaraṇa}. According to Hariścandra, 
\emph{adhikaraṇa} is the reason or ground referring to which the authors direct 
their discourse. For example, diseases create misery and the authors of Āyurveda 
began their discussion addressing them.\footnote{\dev{tatrādhikaraṇaṃ nāma 
yannimittamadhikṛtya pravartate kartā/… uta vā
	vighnabhūtā yadā rogāḥ prādurbhūtāḥ tadidaṃ nimittamadhikṛtya 
jagadanukampayā maharṣibhirayamāyurveda āgamaḥ/ evamadhikaraṇavyākhyā 
varṇayitavyā/} \MScite{MS Jamnagar GAU 114}, p.4--5.}
Thus disease is the \emph{adhikaraṇa} or theme of their discussion. Indu 
identifies \emph{adhikaraṇa} as a binding force that links ideas. According to 
him, 
\emph{adhikaraṇa} as an introductory reference and it exposes a general 
statement to a specific context.\footnote{\dev{adhikaraṇaṃ prastāvaḥ 
sāmānyenoktamapyarthajātaṃ yadbalādviśeṣe'vasthāpyate tadadhikaraṇam/} 
\parencite[959]{atha-1980}.} 

Nīlamegha defines \emph{adhikaraṇa} using the Paninian terminology. 
According to him, \emph{adhikāraṇa} is the locus in which the \emph{tātparya}, 
“reference” lies.\footnote{\dev{tatrādharo'dhikaraṇaṃ tātparyaṃ tatra tiṣṭhati/} 
\cite[1][2]{muth-1976}.} 
\emph{Adhikaraṇa} is one of the six \emph{kāraka}-s (a sort of semantic roles 
recognized by Sanskrit grammarians). 
Pāṇini calls it a locus (\emph{ādhāra}).\footnote{\dev{ādhāro'dhikaraṇam/} 
\emph{Aṣṭādhyāyī} 1.4.45.} 
Through metaphorical extensions, the idea of a locus can apply to abstract 
domains and not merely to physical locus. 
In traditional Sanskrit grammar, a metonymic or metaphorical domain is called 
\emph{vaiṣayika adhikaraṇa}. 
Patañjali considers \emph{vaiṣayika} as one of the three types of 
\emph{adhikaraṇa}.\footnote{On \emph{Aṣṭādhyāyī} 6.1.72 
\volcite{3}[51]{kiel-1880}.} 
Nīlamegha applies this idea to his definition of the \emph{tantrayukti} 
\emph{adhikaraṇa}. 
He quotes \AHS\ \emph{Sūtrasthāna} 1.5b-6a and explains how the eight limbs 
of Āyurveda serve as the \emph{adhikaraṇa}s of \emph{cikitsā} “treatment”. 
In Nīlamegha's understanding, \emph{adhikaraṇa} “theme” is the domain of a 
reference. 
The \emph{Tantrayukti} repeats Nīlamegha's idea but it also adds different types 
of \emph{adhikaraṇa} as suggested by Aruṇadatta. 

In the \emph{Tolkāppiyam}, however, the equivalent expression for this 
\emph{tantrayukti} remains unclear, as commentators, namely Iḷampūraṇar and 
Pērāciriyar, list the item differently. In Sastri's translation of the 
\emph{Tolkāppiyam}, \emph{adhikaraṇa} was identified with \emph{atikāra 
muṟai}, the second element in Iḷampūraṇar's list. Sastri translates this expression 
as “deciding the extent where one serves as \emph{adhikāra sūtra} or a word or 
words in a sūtra taken along with the \emph{sūtra}-s that 
follow.”\footnote{\cite[233]{sast-1936}.} However, Dikshitar, in his brief article 
on 
the \emph{tantrayukti}s, equates \emph{adhikaraṇa} with \emph{nutaliyatu 
aṟital}, the first element in Pērāciriyar's list, and translates it as “that division of a 
book which centers around a chief topic and deals wholly with that 
topic.”\footnote{\cite[85]{diks-1930}} Clearly, Dikshitar's interpretation stands 
close to our definition of \emph{adhikaraṇa}. Sastri's interpretation, on the other 
hand, corresponds to the concept of \emph{adhikāra} “heading” and 
\emph{anuvṛtti} “recurrence” in the \emph{sūtra} literature, especially in 
Pāṇini's 
\emph{Aṣṭādhyāyī}.\footnote{See \cite[111]{chev-2009}.} 

The translators usually translated this \emph{tantrayukti} as “topic” or “subject 
matter”. 

\subsection{2. \emph{yoga}}

This \emph{tantrayukti} typically occupies the second position in most lists, 
except in the \emph{Arthaśāstra} where it appears third following 
\emph{vidhāna}. Functioning as a syntactic and semantic tool, \emph{yoga}, as 
defined in the \SS, represents the faculty responsible for the cohesion of a 
sentence. If we consider the main purpose of the \emph{tantrayukti}s as 
narrated 
in the \SS, namely, cohesion of a sentence (\emph{vākyayojana}) and cohesion 
of 
meaning (\emph{arthayojana}), it becomes evident that this \emph{tantrayukti} 
is one of the fundamental \emph{tantrayukti}s functioning as the device for 
\emph{vākyayojana}. The \SS\ further describes \emph{yoga} as a syntactic 
connection between words, facilitating the linking of words even when they are in 
reverse order or placed apart. However, this paraphrased statement is absent in 
the 
vulgate; instead, it appears in the commentary of Ḍalhana with a minor 
variation.\footnote{See \Su{6.65.9}{815}.} The definitions of \emph{yoga} in 
the 
\emph{Viṣṇudharmottarapurāṇa} and \emph{Arthaśāstra} closely mirror that of 
the \SS. However, the \emph{Viṣṇudharmottarapurāṇa}'s definition introduces a 
slight variation by including the term \emph{artha} “meaning”. According to this 
definition, \emph{yoga} is that by which the meaning of a sentence coheres. The 
\emph{Arthaśāstra} employs a nominalized verb in a compound noun instead of 
a 
relative clause--- \emph{vākyayojanā} “connecting a sentence”. The definition 
we 
find in the \emph{Saddanīti} is close to the paraphrased part of the definition of 
the \SS.\footnote{See Table \ref{table-SAV}.}

For illustration, a verse from chapter 18 of the \emph{Cikitsāsthāna} is quoted in 
the \SS---

\begin{quote}
	
	\dev{tailaṃ pibeccāmṛtavallinimbahaṃsāhvayāvṛkṣakapippalībhiḥ/\\
	siddhaṃ balābhyāñca sadevadāru hitāya nityaṃ 
galagaṇḍaroge//}\footnote{\Su{4.18.47}{474}.}
	
\end{quote}


In this verse, the noun \se{tailam}{sesame oil} appears at the beginning of the 
first hemistich, while its adjective \se{siddham}{cooked} is placed at the 
beginning of the final hemistich. Despite not being colocated, the 
\emph{tantrayukti} \emph{yoga} effectively connects them, facilitating our 
comprehension of the intended meaning. Evidently, this exemplifies a device for 
linking words within a sentence.


In the commentaries of Hariścandra, Indu, Cakrapāṇi and Aruṇa, however, the 
\emph{tantrayukti} \emph{yoga} is used in a broader sense. In these 
interpretations, \emph{yoga} serves not only as a device for cohesion within a 
sentence but also fosters coherence among sentences in a discourse. Hariścandra 
identifies three alternative interpretations of \emph{yoga}.\footnote{\dev{yogo 
nāma yojanā granthānāṃ yathārthasūtrabhāṣyasūtrayoḥ... pañcalakṣaṇo vā 
yogaḥ/ 
pratijñāhetūdāharaṇanigamanāni... yad iha yujyate sa yoga ityeke/} (MS 
Jamnagar 
GAU 114, p.5.)} 
Aruṇadatta also interprets \emph{yoga} in a similar fashion but instead of three 
alternatives he talks about the first two alternatives of Hariścandra. In the first 
alternative, \emph{yoga} is coherence between the main statement 
(\emph{sūtra}) and its gloss (\emph{bhāṣya}).
Aruṇadatta expands its scope to coherence between mention (\emph{uddeśa}) 
and description (\emph{nirdeśa}) as well.\footnote{\dev{yogo nāma yojanā, 
uddeśanirdeśayoḥ sūtrabhāṣyayorvā/} \emph{Sarvāṅgasundarī} on 
\Ah{6.40.80}{947}.}  
In the second alternative, \emph{yoga} is reasoning (\emph{yukti}) having five 
types: (1) \emph{pratijñā} “proposition”, (2) \emph{hetu} “reason”, (3) 
\emph{udāharaṇa} “exemplification” (4) \emph{upanaya} “application”, and (5) 
\emph{nigamana} “conclusion”, resembling the five-membered syllogism of 
inference (\emph{anumāna}) in the Nyāya-Vaiśeṣika school.\footnote{
	\dev{yuktirvā yogaḥ, pratijñā heturdṛṣṭānta upanayo nigamanamiti 
pañcavidhaḥ/} \Ah{6.40.80}{947}.}
Śrīdāsa Paṇḍita's comment is similar to that of 
Aruṇadatta.\footcite[2]{muss-1940}  
Hariścandra also notes a different understanding of this \emph{tantrayukti} by 
some others. In this sense, \emph{yoga} is connectedness. This alternative 
definition is close to that of the \SS. In Indu's interpretation, \emph{yoga} is 
lexical 
cohesion, as he understands \emph{yoga} as a relation between a word and its 
meaning or a sentence and its meaning.\footnote{\dev{yogo nāma yogaḥ 
sambandhaḥ sa ca padārthayorvākyārthayorvā/} \emph{Śaśilekhā} on 
\As{6.50.150a}{959}.} 
Cakrapāṇi, while defining \emph{yoga} in a fashion similar to the \SS, 
exemplifies it as a connection between five logical elements, namely 
\emph{pratijñā}, \emph{hetu}, \emph{udāharaṇa}, \emph{upanaya} and 
\emph{nigamana}, conflating the definition of \emph{yoga} with Hariścandra's 
second alternative i.e. \emph{yoga} is reasoning 
(\emph{yukti}).\footnote{\dev{yogo nāma yojanā vyastānāṃ 
padānāmekīkaraṇam/ udāharaṇaṃ tāvadyathā 
pratijñāhetūdāharaṇopanayanigamanāni/} \Ca{8.12.41}{736}.} 
Nīlamegha defines \emph{yoga} as connecting words one by one 
coherently.\footnote{\dev{yogaḥ padānāmekaikamarthaucityena yojanā/} 
\cite[2][3]{muth-1976}.} As he further explains in the autocommentary with 
examples from the \AHS, it is evident that he understands \emph{yoga} as 
coherence between a part of a sentence and the 
discourse.\footcite[3]{muth-1976} 
Neither V. R. Ramachandra Dikshitar nor P. S. Subrahmanya Sastri identified the 
\emph{tantrayukti} \emph{yoga} with any \emph{utti} mentioned in the 
\emph{Tolkāppiyam}.\footnote{\cite[84]{chev-2009}.} 

The word \emph{yoga} derives from the Sanskrit root √yuj “to connect” with the 
primary suffix \emph{GHaÑ}, which is often used for creating action nouns. In 
Sanskrit technical literature, the term \emph{yoga} is used in a broad sense to 
mean any kind of linguistic connection or connectedness. In the 
\emph{Aṣṭādhyāyī} of Pāṇini, it often refers to the connection with a word or a 
word-element.\footnote{\cite[64]{josh-1991}.} Hence, it refers to a 
morphosemantic or syntaco-semantic connection. Patañjali uses this term several 
times in his \emph{Mahābhāṣya}. In the \SS\, the word \emph{yoga} is primarily 
used to mean the connection between words in a sentence. According to this 
definition and illustration, it is primarily intra-sentential cohesion. Unlike the later 
commentators on the works of Caraka and Vāgbhaṭa, it does not extend the 
scope 
of this term to inter-sentential cohesion and coherence. Keeping in mind such 
definition given in the \SS, we translate the term as cohesion even though no 
other 
translators of the \emph{tantrayukti}s used this translation. In some other 
contexts, however, \emph{yoga} can be extended to coherence. Both coherence 
and cohesion are derived from the Latin verb \emph{cohaere-} (< \emph{con}- 
“with” \emph{haereō} “cling”) “to cling together.”  In other translations of the 
\emph{tantrayukti}s (see p. \pageref{tantra-trans}), \emph{yoga} is variously 
translated as employment, arrangement, conjoiner, connecting, concomitance, 
uniting, union, rational linking, joining and so on. We preferred the term cohesion 
because the other options are either too narrow or too vague. `Employment' is 
rather \emph{prayoga}, not \emph{yoga}. `Rational linking' disregards the 
grammatical aspect of \emph{yoga}. `Conjoiner', `connecting', `union', `uniting' 
or 
`arrangement' are vague and they do not reflect the technical import of the term 
\emph{yoga}.

\subsection{3. \emph{padārtha}}

In the earlier listing, \emph{padārtha} follows \emph{yoga}, while in the later 
listing, this \emph{tantrayukti} is enumerated after \emph{hetvartha}, possibly 
due to metrical requirements. The \SS\ dedicates more words to describing this 
\emph{tantrayukti} than any other early texts.

The description in the \SS\ commences with the definition of \emph{padārtha}, 
which is articulated as the meaning conveyed in an aphorism or a word. It then 
delves into the literal interpretation of the term \emph{padārtha}. The 
straightforward meaning of the compound \emph{padārtha}, obtained by 
dissecting its components---\emph{pada} “word," and \emph{artha} 
“meaning”---is “the meaning of one or more words.” After presenting the literal 
interpretation of \emph{padārtha}, the \SS\ provides the rationale why mere 
word 
meanings cannot suffice as the \emph{tantrayukti padārtha}---since a word or 
words may have multiple meanings. Therefore, as a \emph{tantrayukti}, the 
term 
\emph{padārtha} denotes the meaning of a word or words within a specific 
context.\footnote{See fn. \ref{fn.contrary}.} 
Ḍalhaṇa also supports this perspective, indicating that the term 
\emph{padārtha} refers to a specific meaning of a word or 
words.\footnote{\dev{adhunā bahuṣu padārtheṣu nirdhārya\
		viśiṣṭapadārthagrahaṇāya padārthabahutvaṃ pratipādayati}--- 
\dev{aparimitā ityādi/} \Dalhana{6.65.10}{816}.}
	
In \Su{}{}, a variant reading of the definition of \emph{padārtha} is noted: 
\dev{sūtrapade} in the place of \dev{sūtre pade vā}. It remains unclear which 
reading was available to Ḍalhaṇa. He proposed an etymological meaning of the 
word \emph{pada}, defining it as that by which a meaning is understood, and 
includes \emph{sūtra} under the semantic scope of \emph{pada}. Essentially, 
he 
viewed \emph{sūtra} as a type of \emph{pada} because, by conveying a 
meaning, 
a \emph{sūtra} falls under the category of \emph{pada}, which by definition 
signifies a meaning-conveying unit. This interpretation does not separate 
\emph{sūtra} and \emph{pada} as mutually exclusive entities.
Thus, if Ḍalhaṇa's reading of the text is \dev{sūtre pade vā}, he perceived 
\emph{pada} as a synonym or an alternative term for \emph{sūtra}. On the 
other 
hand, if the reading was \dev{sūtrapade}, he understood the meaning of the 
word 
\dev{sūtrapade} as \emph{pada} (a meaning-conveying unit) in the form of a 
\emph{sūtra}. The editor of \Su{}{} offered a more straightforward explanation 
of 
the variant reading \dev{sūtrapade}---a \se{pada}{word} in a \emph{sūtra} is a 
\emph{sūtrapada}.\footnote{\dev{anye tu sūtrapade iti paṭhitvā 
vyākhyānayanti--- sūtrasya padaṃ sūtrapadaṃ tasmin yo'rthaḥ sa padārthaḥ/} 
fn.2 
\Su{}{813}.} The \emph{Viṣṇudharmottarapurāṇa} adopts a definition of 
\emph{padārtha} close to this variant reading.\footnote{See table 
\ref{table-SAV}.}

The \emph{tantrayukti} \emph{padārtha} is illustrated with two examples in the 
\SS. In the first example, three polysemous words---\emph{sneha}, 
\emph{sveda} 
and \emph{abhyañjana}— are provided. Ḍalhaṇa gave various meanings of 
these 
words--- \emph{sneha} can mean lubricity or grease or affection; \emph{sveda} 
can mean sauna either with fire (\emph{sāgni}) or without fire (\emph{niragni}); 
\emph{abhyañjana} may mean the black eye make-up or \se{abhyaṅga}{oil 
rub}.\footnote{\dev{tatra snehaśabde nirdiṣṭe hi guṇapremasarpiṣāṃ 
trayāṇāmarthānāmupapattirdṛśyate, svedaśabdenāpi sāgniniragnikayorūṣmaṇoḥ 
prāptiḥ, añjanaśabdenāpi nayanāñjanābhyaṅgayoḥ prāptiḥ/} 
\Dalhana{6.65.10}{816}.} 
In the case of such polysemy, \emph{padārtha} or relevant meaning will be the 
meaning which coheres with the prior and subsequent elements 
(\emph{pūrvāparayogasiddha}). The definition of this \emph{tantrayukti} in 
Saddanīti corresponds to this explanation.\footnote{See table \ref{table-SAV}.} 
Thus, \emph{padārtha} is that meaning which fits the context.   

The second example is taken from the very beginning of the first chapter of 
the \emph{Sūtrasthāna} that says \dev{vedotpattimadhyāyaṃ 
vyākhyāsyāmaḥ}, 
“I shall narrate the chapter on the origin of \se{veda}{knowledge}.” 
The problem is, what does this word “\emph{veda}” refer to?  
Is it the \emph{Veda}, as in \emph{Sāmaveda}?  Or
something derived from the roots \root \emph{vi(n)d} or \root \emph{vid}?  
Context (“prior and subsequent elements”) can help us to know that 
“\emph{veda}”
means only \emph{āyurveda} and that the \SS\ is talking about the
origin of \emph{āyurveda}, specifically.  
The same issue is also addressed by Ḍalhaṇa at \Su{1.1.1}{1}.

Among the texts of the early listing, 
the \emph{Arthaśāstra} presents a notably distinct definition of 
\emph{padārtha}. 
Here, it is defined as that which has its limit within the word.\footnote{
	\dev{padāvadhikaḥ padārthaḥ/} (\volcite[15.1.10]{1}[280]{kang-1969}).}
Though somewhat ambiguous, this definition implies 
that \emph{padārtha} is the referent indicated by a word or in other words, 
\emph{padārtha} is the scope of meaning that corresponds to a word. 
Hence, this definition of \emph{padārtha} does not necessarily refer to a 
contextual meaning. 
It indirectly suggests that \emph{padārtha} is basically the meaning of a word. 

Similar to \emph{Arthaśāstra}'s understanding, the commentators 
Hariścandra, Indu, Cakrapāṇi, Aruṇadatta and Śrīdāsapaṇḍita 
interpret \emph{padārtha} as the referents indicated by a word. 
However, by this time, the term \emph{padārtha} became 
an important point of discussion among certain philosophical schools, 
particularly Vaiśeṣika and Nyāya.\footnote{
	The concept \emph{padārtha} is also discussed by grammarians such as 
Patañjali and others. For the treatment of the term \emph{padārtha} in different 
Indian philosophical schools, see \volcite{2}[153--154]{pret-1991}.} 
In the Vaiśeṣika ontology, padārtha is the term used for denoting the 
fundamental ontological categories. 
Such a wider use of the term among philosophical schools also influenced the 
\emph{Bṛhattrayī} commentators, 
most prominently Hariścandra, whose interpretation of \emph{padārtha} aligns 
with the framework of Vaiśeṣika philosophy. Following the Vaiśeṣika doctrine, he 
lists six types of \emph{padārtha}s, namely, \se{dravya}{substance}, 
\se{guṇa}{attribute}, \se{karman}{movement}, \se{sāmānya}{universality}, 
\se{viśeṣa}{individuality}, and \se{samavāya}{inherence}.\footnote{
	\dev{padārtho nāma ya ekena padenānekārtho gamyate/ 
		yathā dravyaṃ guṇaḥ karma sāmānyaṃ viśeṣaḥ samavāyaḥ/}. 
	\MScite{Jamnagar GAU 114}, p.6.} 
In his understanding, a \se{pada}{word} is a universal category 
that may have several referents called \emph{padārtha}.
While other commentators such as 
Indu, Aruṇadatta, Śrīdāsa Paṇḍita follow Hariścandra 
while defining this \emph{tantrayukti} and
cite Vaiśeṣika \emph{padārtha}s 
such as \emph{dravya} or \emph{guṇa} as instances of \emph{padārtha}, 
it is not clear whether they endorse the Vaiśeṣika interpretation.\footnote{
		\dev{padārtho nāma yenārtho gamyate/ 
			yathā gurvādayo guṇaśabdādavagamyate/} \emph{Śaśilekhā} on 
\As{6.50.150a}{959}.
		\dev{padārtho nāma, padenārtho gamyate/
			yathā dravyamiti padaṃ, tasyārtho bhūjalādiḥ/
			guṇa iti padaṃ tasyārtho gurvādiḥ/} \emph{Sarvāṅgasundarī} on 
\Ah{6.40.80}{947}.
		The	same reading appears in \emph{Hṛdayabodhikā} 
\parencite[2]{muss-1940}.} 
Cakrapāṇi adopts the literal definition of \emph{padārtha} as mentioned in the 
\SS. 
With examples, he emphasizes that \se{padārtha}{word-meaning} 
can stem from one word, two words or more.\footnote{
		\dev{padasya padayoḥ padānāṃ vā'rthaḥ padārthaḥ/
			tatra dravyamiti padena khādayaścetanāṣaṣṭhā
			ucyante; padayorartho nāma yathā}---‘\dev{āyuṣo veda}’ 
		\dev{iti padayorāyurbodhakaṃ tantramityarthaḥ, 
			evaṃ padānāmapyartha udāhāryaḥ/} 
		\emph{Āyurvedadīpikā} on \Ca{8.12.41}{736}.} 
Gaṅgādhara, however, defined \emph{padārtha} in the line of the \SS\ 
considering \emph{padārtha} as relevant word sense.\footnote{
		\dev{arthaḥ padasya ceti padārtho nāma tantrayuktiḥ 
		sā yo'rtho'bhihitaḥ sūtre pade vā/
		padārthastvanekastatra yo'rthaḥ pūrvāparayogasiddho bhavati
		so'rtho grāhyaḥ/} \emph{Jalpakalpataru} on 
\fvolcite{3}[3816]{sena-1928}.}

Nīlamegha presents a distinct perspective, 
defining padārtha as polysemy within a given context.\footnote{
	\dev{padārthastu padaikye'pi bhinnamarthaṃ prakāśayet/} 
\cite[4][4]{muth-1976}.} 
In his autocommentary, he also acknowledges another viewpoint 
that perceives \emph{padārtha} as synonymy within a context.\footnote{
	\dev{kecittu} “\dev{padārthaḥ padabhede'pi na bhedaḥ punararthataḥ/}/
	\cite[5]{muth-1976}.} 
The author of the \emph{Tantrayukti} merely quoted Nīlamegha's statements on 
this matter 
while incorporating the additional definition of \emph{padārtha} 
found in commentaries such as those of Aruṇadatta, Indu, or Śrīdāsa 
Paṇḍita.\footnote{
	\begin{verse}
		\dev{padārthastu ya aikye'pi bhinnamarthaṃ prakāśayet\\ \ldots
		padārtho'pi ca bhede'pi na bhedaḥ punararthataḥ/\\
		padena yo'rtho jñāyate yathā gurvādayo guṇe//\\} 
	\end{verse}
	\cite[8--10]{nara-1949}.}
Neither V. R. Ramachandra Dikshitar nor P. S. Subrahmanya Sastri 
identified the \emph{tantrayukti} \emph{padārtha} with any \emph{utti} 
mentioned in the \emph{Tolkāppiyam}.\footcite[84]{chev-2009}

From the discussion above, it is clear that \SS 's understanding 
of the \emph{tantrayukti} \emph{padārtha} is distinct from its definition 
offered by the later commentators of the works of Dṛḍhabala and Vāgbhaṭa. 
The \emph{Arthaśāstra}, 
which most often defines the \emph{tantrayukti}s similarly to the \SS, 
provides a definition of \emph{padārtha} that is closer 
to the understanding of the commentators such as Hariścandra and so on. 
In the \SS, \emph{padārtha} is not merely the meaning of a word or words 
but the meaning arising within a particular context or co-text. 
It is the result of word sense disambiguation. 
Singhal and Mitra translates \emph{padārtha} as “context.” 
However, padārtha denotes not the context but rather the meaning intended in a 
context or co-text. 
Most other translators render this term as “import of words,” 
which fits better in this case. 
The translation chosen here is “relevant meaning” 
because in the \SS, \emph{padārtha} refers to the meaning 
that is relevant within a context or co-text.

\subsection{3. \emph{hetvartha}}

\emph{Hetvartha} appears after padārtha in the earlier listing 
and before padārtha in the later listing. 
The word \emph{hetvartha} is a compound 
of two words---\se{hetu}{reason} and \se{artha}{purpose}. 
Although the term \emph{hetvartha} is present in all lists of the 
\emph{tantrayuktis}, 
it has not been lexicalized as a compound word. 
Thus, we need to understand the term 
through its components---\emph{hetu} and \emph{artha}. 
Depending on the meaning of the word \emph{artha}, 
the word \emph{hetvartha} can have different meanings. 
For example, Bhaṭṭoji Dīkṣita uses the term \emph{hetvartha} 
to mean simply the sense of a cause.\footnote{%
	\dev{hetvarthe tṛtīyā syāt |} \cite[137]{gada-1904}.} 
As a \emph{tantrayukti}, \emph{hetvartha} is a logical device 
that serves as the \se{artha}{purpose} of a \se{hetu}{reason}. 

The \SS\ provides an analytical definition of \emph{hetvartha}, 
where it is described as a statement functioning as a \se{sādhana}{premise}. 
The text supplies an example about moistening of wounds by milk etc. 
on the basis of the known fact that water moistens a lump of earth. 
Notably, the word \emph{ukta} in the definition 
likely does not imply an explicit statement in the text. 
As appears in the example, it can be a known fact from the outside world 
that aids in predicting a similar case related to our body.\footnote{%
	Cf. \dev{atra bāhyena mṛtpiṇḍadṛṣṭāntena 
		māṣadugdhayogādibhirābhyantaro vraṇaprakledaḥ sādhyate/} 
	\Dalhana{6.65.11}{813}.} 
The \SS\ does not mention that water moistens a lump of earth; 
this is understood from general empirical knowledge. 
The \SS\ prescribes moistening of a wound in certain cases 
but does not explicitly state how to do so. 
The knowledge that milk and similar substances 
can be used to moisten a wound derives 
from the empirical knowledge of moistening a lump of earth with water. 
The causal relationship between water and moistening a lump of earth 
serves a purpose elsewhere as a premise 
for understanding the causal relationship 
between milk or other similar substances 
and moistening of a wound.\footnote{%
	The definition is similar in \emph{Saddanīti}. 
	In the \emph{Arthaśāstra}, \emph{hetvartha} is defined 
	as a cause that serves a purpose. 
	Although phrased differently, 
	this definition refers to the same concept. 
	See table \ref{table-SAV}.} 
The definition is clearer in \Su{6.65.11}{813} 
because it includes the word \se{anya}{other}, 
emphasizing that an idea stated in one one context 
serves a purpose in another.\footnote{\dev{yadanyaduktamanyārthasādhakaṃ 
		bhavati sa hetvarthaḥ/}	\Su{6.65.11}{813}.
		The testimonium in Gaṅgādhara's \emph{Jalpakalpataru} 
		supports the reading of \Su{}{}. 
		In his own definition, Gaṅgādhara merely reproduces the definition of the 
\SS. 
		\volcite{3}[3815]{sena-1928}.} 
The author of the \emph{Viṣṇudharmottarapurāṇa} as well as 
commentators such as Hariścandra, Cakrapāṇi, Indu, Aruṇadatta, and Śrīdāsa 
Paṇḍita 
define this \emph{tantrayukti} similarly.\footnote{%
	The definition of \emph{hetvartha} is quite consistent
	across all the commentaries.
	Cakrapāṇi's definition: \dev{hetvartho nāma 
		yadanyatrābhihitamanyatropapadyate/} 
		\emph{Āyurvedadīpikā} on \Ca{8.12.41}{736}.
	Indu's definition: \dev{hetvartho nāma 
		yadekatrocyamānamanyatrāpi tathaivopayujyate/}
		\emph{Śaśilekhā} on \As{6.50.150a}{959}.
	Hariścandra's definition is also quite similar: 
		\dev{hetvartho nāma 
		yadanyaprastāvābhihitamarthajātamanyatrāpi 
		tathaivāpādyate/} \MScite{Jamnagar GAU 114} p.\,5.
	Aruṇadatta's definition: \dev{hetvartho nāma 
		yadanyaprastāvoktamanyatrāpi 
		tathaivāpādyate/}
		\emph{Sarvāṅgasundarī} on \Ah{6.40.80}{947}.
	Śrīdāsa Paṇḍita's definition: \dev{hetvartho 
		nāmānyaprastāvoktamanyatrāpi 
		tathaivāpādyate/}
		\emph{Hṛdayabodhikā} \parencite[2]{muss-1940}.
		} 
However, in their interpretation, the term \se{ukta}{stated} 
means an explicit statement in the text. 

A different definition appears in Nīlamegha's \emph{Tantrayuktivicāra}, 
where he defines \emph{hetvartha} as a situation 
where an entity is represented by its cause. 
For example, in the statement, \dev{rogas tu doṣavaiṣamyam}, 
“disease, however, is the disproportion of the humours,” 
the cause “disproportion of the humours” represents its effect, “disease.” 
It can also be the reverse, 
where disease is the cause of disproportion of the humours. 
When a cause and its effect are considered equivalent 
and one represents the other, it is called \emph{hetvartha}.\footnote{%
	\begin{verse}
		\dev{hetvartho hetunaiva syāt tattadarthaprakāśanam//3//}
	\end{verse}
	\dev{yathā—‘rogastu doṣavaiṣamyam’ (a.hṛ.sū. 1.29) 
		ityādau rogo nāma doṣavaiṣamyahetuḥ/ na tu doṣavaiṣamyam/ 
		tattu vṛddhiḥ kṣayo vā/ ato hetor eva rogaḥ/ 
		atra tuśabdena rogo'pi doṣavaiṣamyasya heturiti dyotayati/}. 
		\cite[4]{muth-1976}.} 
This definition by Nīlamegha is not found elsewhere. 
This may explain why the author of the \emph{Tantrayukti} 
does not refer to this definition at all, 
even though he frequently quotes Nīlamegha. 
Instead, the author of \emph{Tantrayukti} 
uses Aruaṇadatta's definition without attribution. 
He concludes with another definition, describing \emph{hetvartha} as a 
statement 
where a reason is expressed.\footnote{%
	\begin{verse}
		\dev{hetunā saha yatroktiḥ sa hetvarthaḥ prakīrtitaḥ//}
	\end{verse}
	\cite[8]{nara-1949}.} 
Neither V. R. Ramachandra Dikshitar nor P. S. Subrahmanya Sastri 
identified \emph{hetvartha} with any \emph{utti} 
mentioned in the \emph{Tolkāppiyam}.\footcite[84]{chev-2009} 

The term \emph{hetvartha} has been translated in various ways, 
including “extension of argument,” “implication,” 
“goal of a reason,” and merely “reason.” 
We chose to translate \emph{hetvartha} as “purpose of a reason” 
based on the components of the compound, 
which closely aligns with Olivelle’s translation,
“goal of a reason.”\footcite[436]{oliv-2013}
While other translations may capture the application of \emph{hetvartha}, 
they do not convey the lexical meaning of the term.

\subsection{5–6. \emph{uddeśa} and \emph{nirdeśa}}

It is necessary to consider the \emph{tantrayukti}s 
\emph{uddeśa} and \emph{nirdeśa} in relation to each other 
as they form a pair of relational antonyms. 
They consistently appear together in all listings---following \emph{hetvartha} 
in the earlier listing and \emph{pradeśa} in the later listing. 
The \SS\ defines \emph{samuddeśa} as a brief statement 
and \emph{nirdeśa} as a detailed statement. 
An example of \emph{uddeśa} is given as the simple mention of the word 
“\se{śalya}{spike}.” 
In contrast, the example of \emph{nirdeśa} is the phrase “in the body or 
exogenous,” 
where spike is described in more detail as being of two kinds. 
These two \emph{tantrayukti}s are stylistic structural devices 
used in scientific compositions for precision and clarity.

The texts of the earlier listing provide identical or near-identical definitions 
of \emph{uddeśa} and \emph{nirdeśa}.\footnote{See table \ref{table-SAV}.} 
They are also similarly defined and exemplified in the 
\emph{Yuktidīpikā}.\footcite[7]{wezl-1998} 
Commentators on the works of Dṛḍhabala and Vāgbhaṭa 
also defined \emph{uddeśa} and \emph{nirdeśa} similarly.\footnote{%
	\dev{uddeśo nāma saṅkṣepābhidhānam}\ldots
	\dev{nirdeśo nāma vistā...} 
		\MScite{Jamnagar GAU 114} p.\,6.	
	\dev{uddeśo nāma saṅkṣepābhidhānam}\ldots
	\dev{nirdeśo nāma saṃkhyeyoktasya (saṅkṣepoktasya ?) vivaraṇam/} 
		\emph{Āyurvedadīpikā} on \Ca{8.12.42}{736}. 
	\dev{uddeśo nāma saṅkṣepābhidhānam}\ldots
	\dev{nirdeśo nāma tasyaiva vistāroktiḥ/}
		\emph{Sarvāṅgasundarī} on \Ah{6.40.80}{947} and 
		\emph{Hṛdayabodhikā} \parencite[2]{muss-1940}.
	\dev{uddeśo nāma tantrayuktiḥ sā, yat samāsakathanam/}\ldots
	\dev{nirdeśo nāma tantrayuktiḥ sā, yad vistareṇocyate/}
		\emph{Jalpakalpataru}, \volcite{3}[3816]{sena-1928}.
	} 
However, Indu's definitions of these two \emph{tantrayukti}s are more 
informative. 
According to him, \emph{uddeśa} refers to objects mentioned merely by single 
words, 
while \emph{nirdeśa} involves restating those objects to show their specific 
features.\footnote{%
	\dev{uddeśo nāma yatrārthānāṃ śabdamātreṇaiva kīrtanamuddeśaḥ/}\ldots
	\dev{nirdeśo nāma yacchabdamātreṇa nirdiṣṭānāṃ 
		svarūpaviśeṣapradarśanāya punaḥ kīrtanaṃ nirdeśaḥ/}
		\emph{Śaśilekhā} on \As{6.50.150b}{960}.
	}

There is a difference between the examples of 
\emph{uddeśa} and \emph{nirdeśa} in the \SS\ and other texts. 
In the \SS, the example of \emph{uddeśa} is a single word, 
not a complete sentence, while in other texts, including the \emph{Arthaśāstra}, 
it is a complete sentence, usually a simple equative sentence. 
For instance, Aruṇadatta and Śrīdāsa Paṇḍita quote \Ah{1.1.6b}{6} as an 
example of \emph{uddeśa}: 
\dev{vāyuḥ pittaṃ kaphaśceti trayo doṣāḥ samāsataḥ/}, 
“the three humours are wind, bile, and phlegm.” 
As an example of \emph{nirdeśa}, they quote \Ah{1.1.11}{9}, 
which describes the characteristics of wind: 
\dev{tatra rūkṣo laghuḥ śītaḥ kharaḥ sūkṣmaścalo'nilaḥ/},
“wind is rough, light, cold, harsh, subtle, and mobile.”
Even in these cases, where \emph{uddeśa} is given with a complete sentence, 
\emph{uddeśa} is exemplified by the individual items in those sentences. 
Thus, there are three \emph{uddeśa}s in that sentence, 
and the description of each functions as \emph{nirdeśa}. 
What constitutes \emph{uddeśa} and \emph{nirdeśa} is contextually 
determined. 
In the example of the \SS, \se{śalya}{spike} is the \emph{uddeśa}, 
and its \emph{nirdeśa} is the expression that provides its two varieties: 
(1) in the body and (2) extraneous. 
When each variety is further described, each variety of the spike functions as 
\emph{uddeśa}. 
The author of the \emph{Tantrayukti} indicates 
this mutual relatedness of \emph{uddeśa} and \emph{nirdeśa} in a 
verse.\footnote{
	\begin{verse}
		\dev{uddeśanirdeśakayoranyonyāpekṣitā bhavet/}\\
		\dev{yattadoriva nityaiva śāstre sarvatra sarvadā//}\\
	\end{verse}
	\cite[12]{nara-1949}.} 
He also provides a versified definition of these two \emph{tantrayukti}s, 
in addition to repeating Nīlamegha's statements.\footnote{
	\begin{verse}
		\dev{prāk saṅkṣepeṇa kathanamuddeśa iti kīrtitam/}\\
		\dev{nirdeśo nāma tasyaiva vistaroktirudāhṛtaḥ//}\\
	\end{verse}
	\cite[12]{nara-1949}.} 
Nīlamegha also defined these tantrayuktis in a similar fashion.\footnote{
	\begin{verse}
		\dev{uddeśaḥ samavāyoktiriti prāhurmanīṣiṇaḥ/}\\
		\dev{nirdeśaḥ syād vivaraṇaṃ pūrvoktānāmanukramāt//}\\
	\end{verse}
	\cite[6]{muth-1976}.}  
Similar \emph{utti}s also appear in the \emph{Tolkappiyam}. 
However, Sastri and Dikshitar differently identified these \emph{tantrayukti}s 
with the \emph{utti}s of \emph{Tolkāppiyam}.\footcite[85]{chev-2009}

\emph{Uddeśa} is a common technical term in Sanskrit literature.\footnote{%
	For more references to the term \emph{uddeśa} check 
\volcite{2}[28--30]{pret-1991}.} 
Sometimes it appears paired with \emph{lakṣaṇa}, 
where \emph{uddeśa} means mentioning an item by name,
and \emph{lakṣaṇa} is its definition. 
In this sense, \emph{lakṣaṇa} and \emph{nirdeśa} are similar. 
However, \emph{nirdeśa} means any elaboration of the mentioned item, 
whereas \emph{lakṣaṇa} refers to precise features. 
In Pakṣilasvāmin's \emph{Nyāyabhāṣya}, the term \emph{uddeśa} appears with 
\emph{lakṣaṇa} and \se{parīkṣā}{investigation}, 
and the triad of these three is called the course of the discipline.\footnote{%
	\dev{trividhā cāsya śāstrasya pravṛttiḥ, uddeśo lakṣaṇaṃ parīkṣā ceti/
	tatra nāmadheyena padārthamātrasyābhidhānamuddeśaḥ/
	tatroddiṣṭasyātattvavyavacchedako dharmo lakṣaṇam/
	lakṣitasya yathālakṣaṇamupapadyate na veti pramāṇairavadhāraṇaṃ parīkṣā/} 
		\cite[14]{josh-1922}.}
	
We mentioned before that Hariścandra added four more \emph{tantrayukti}s to 
the later listing. 
Indu and Cakrapāṇi suggested that 
the scope of these \emph{tantrayukti}s could be included in those already 
enumerated. 
According to Cakrapāṇi \emph{paripraśna} could be included in \emph{uddeśa} 
and \emph{vyutkrāntābhidhāna} is a variety of \emph{nirdeśa}.\footnote{%
	\dev{tatra paripraśna uddeśe'ntarbhavati,}\ldots
	\dev{vyutkrāntābhidhānaṃ nirdeśaprabhedaḥ}
	\emph{Āyurvedadīpikā} on \Ca{8.12.41b–45a}{737}.} 
The only manuscript of Hariścandra's \emph{Carakanyāsa} available to us 
(\MScite{MS Jamnagar GAU 114}) 
is full of lacunae, making it challenging to determine its reading. 
It seems that asking a question about a topic is \emph{paripraśna}, 
and stating something without mentioning the order of its items or objects is 
\emph{vyutkrāntābhidhāna}. 
Including \emph{paripraśn}a and \emph{vyutkrāntābhidhāna} under 
\emph{uddeśa} and \emph{nirdeśa} respectively may be an oversimplification.

The tantrayukti \emph{uddeśa} has been variously translated as 
enunciation, concise statement, allusion, mention, etc. 
We chose the translation “mention.” 
The expressions enunciation, allusion, and concise statement are vaguer than 
“mention.” 
The \emph{tantrayukti} \emph{nirdeśa} is translated as 
elaboration, detailed statement, explanation, amplification, exposition, etc. 
All these translations are accurate. 
We translated \emph{nirdeśa} as “exposition.”

\section{Notes on Significant Variants}

\subsection{\ldots\dev{dvitīye pāde}\ldots}\label{yogan1}
	
The Nepalese version reads \dev{dvitīye pāde} which would
properly mean the second quarter of the first line; the vulgate reads
\dev{tṛtīye pāde} “third quarter” which seems more correct.

\subsection{\dev{yatra tu 
snehasvedābhyañjaneṣu}\ldots\dev{pūrvāparayogasiddho bhavati/}}		

There is a dangling relative clause,
\dev{yo'rthaḥ}, in the Nepalese version that is avoided in the vulgate
recension by the addition of \dev{sa grahītavyaḥ}. 
There are two possible explanations for this discrepancy: 
firstly, the missing main clause may have been present in the archetype 
but inadvertently omitted in the Nepalese version due to a scribal error. 
Alternatively, the main clause could have been elliptical in the archetype. 
The scribes of the Nepalese manuscripts accurately transmitted the text. 
However, at some stage during the transmission process, 
the main clause was supplied as an attempt to rectify the ungrammatical 
sentence.
The interpolation may also stem from Ḍalhaṇa's commentary in which the exact 
clause was used.\footnote{
\dev{tatra yo'rtha ityādi/ 
pūrvoktaparoktavākyasambandhenopapanno yo'rtho bhavati 
sa grahītavya ityarthaḥ} \Dalhana{6.65.10}{816}.}
Considering the principle \emph{lectio difficilior potior}, 
we may posit that the second scenario is more plausible. 
This is because the subject of the main clause can be inferred from 
the subject of the previous sentence, and within the context, 
the meaning of the sentence remains totally intelligible even without the explicit 
main clause.
	
\subsection{\dev{sāmavedādayaśca vedāḥ}}
	
Both \SuComma{}{} and the excerpts from the \SS\ cited in the 
\emph{Jalpakalpataru} read \dev{ṛgvedādayastu vedāḥ}, ê
“\emph{Ṛgveda} and so on are the Vedas.” 
 Traditionally, the \emph{Ṛgveda}, being the earliest composed \emph{Veda}, 
is often considered the prototype. 
However, the selection of \emph{Sāmaveda} as the prototype in the Nepalese 
Version of the \SS\ is intriguing. 
This choice brings to mind a verse from the \emph{Bhagavadgīta} where Kṛṣṇa 
declares, 
\dev{vedānāṃ sāmavedo'smi}\footcite[10.22][456]{pans-1936}, 
"I am the \emph{Sāmaveda} among the Vedas." 
With its incorporation of musical elements, 
the \emph{Sāmaveda} holds a unique charm compared to the \emph{Ṛgveda}. 
Thus, the decision to prioritize the \emph{Sāmaveda} as the prototype 
	may stem from its intrinsic appeal or enchantment rather than chronological 
precedence.\footnote{
	Cf. Madhusūdana Sarasvatī's comment on the same verse: 
	\dev{caturṇāṃ vedānāṃ madhye gānamādhuryeṇātiramaṇīyaḥ 
sāmavedo'hamasmi} 
	\parencite[10.22][456]{pans-1936}, 
	“amongst the four Vedas I am the \emph{Sāmaveda}, which is extremely 
delightful due to its musical charm.”}
However, this reading not just appear in the Nepalese version. 
Another early \SS\ manuscript from 1595 also keeps the same 
reading.\footnote{\MScite{Jodhpur RORI 20060} f.265r5.}
	 
\subsection{\ldots\dev{vida vinda ityetayośca dhātvoḥ}\ldots}
	 
Three issues need to be addressed here: 
	(1) nomenclature of the verb roots in Sanskrit, 
	(2) the homonymy of \emph{vid}, and 
	(3) variant readings in \SuComma{}{} and \emph{Jalpakalpataru}. 
	 		
(1) The text suggests the etymology of the word \emph{veda} by mentioning 
two verb roots, namely \emph{vinda} and \emph{vida}. 
	It is worth noting that there are multiple ways of representing Sanskrit verb 
roots, 
	even within traditional Sanskrit grammar like that of Pāṇini. 
	Verb roots are presented in various forms, including mere lexical root 
forms,\footnote{
		E.g. as \root \emph{gup}, \root \emph{tij}, and \root \emph{kit} are 
mentioned in 
			\dev{guptijkidbhyaḥ san} (\volcite{}[3.1.5]{shar-2002}).}	 		
   forms ending in \emph{-a},\footnote{
 		 		E.g. as \root \emph{gam}, \root \emph{han}, \root \emph{vid}, and 
\root \emph{viś} are mentioned in \dev{vibhāṣā gamahanavidaviśām} 
(\volcite{}[7.2.68]{shar-2002}).}	 		 
 		 ending in \emph{-i},\footnote{ 
 		E.g. as \root \emph{mṛj} is mentioned in \dev{mṛjervibhāṣā} 
(\volcite{}[3.1.113]{shar-2002}).}	 		 
	  or with the ending \emph{-ti} in the present stem,\footnote{
  		E.g. as \root \emph{as}, \root \emph{vac}, and \root \emph{khyā} are 
mentioned in
   		\dev{asyativaktikhyātibhyo'ṅ} (\volcite[]{}[3.1.52]{shar-2002}).} 		   
	    or sometimes with \se{anubandha}{indicatory sounds} as found in the 
\emph{Dhātupāṭha}\footnote{
   		E.g. as \root \emph{iṣ} is mentioned in \dev{iṣugamiyamāṃ chaḥ} 
(\volcite{6}[7.3.77]{shar-2002}). 
   		Cf. \dev{iṣum̐ icchāyām} (\emph{Dhātupāṭha} 6.78).} 
	In the reading of the Nepalese version of the \SS, the second option, 
	representing the verbs with a final \emph{-a}, is adopted. 
	 		
(2) The second issue pertains to the homonymy of \emph{vid} in Sanskrit 
\emph{Dhātupāṭha}s, 
	 		where at least four homonymous verbs are mentioned. 
	 		They all belong to different classes and signify different meanings: 
	 		\emph{\root vid} “to know” belongs to the second class (\emph{adādi} or 
the root class), 
	 		\emph{\root vid} “to find, to attain” to the sixth class (\emph{tudādi} or 
the suffixally accented thematic class), 
	 		\emph{\root vid} “to consider” to the 7th class (\emph{rudhādi} or the 
athematic nasal infix class) and 
	 		\emph{\root vid} “to exist” to the fourth class (\emph{divādi} or the 
thematic \emph{ya}-suffix class).\footnote{
	 			Cf. \dev{sattāyāṃ vidyate jñāne vetti vinte vicāraṇe/ vindate vindati 
prāptau śyanlukśnamśeṣvidaṃ kramāt//} 
	 			\parencite[402]{gada-1904}.}  
 			A 10th-class verb \emph{\root vid} is also mentioned in the 
\emph{Dhātupāṭha} 
 			but this appears to be derived from the other \emph{vid} verbs with a 
pleonastic causative suffix. 
 			The 4th-class \emph{vid} is also originally a derivative of the other 
\emph{vid} verbs formed with the passive suffix. 
 			The sixth-class verb \emph{vid} belongs to a subclass called 
\emph{mucādi}, characterized by a nasal infix. 
 			Thus, it is clear that \emph{vinda}, the first of the two verbs mentioned in 
the \SS, 
 			is the \emph{vid} of the 6th class. 
 			The form \emph{vinda} is, therefore, the present stem of the sixth-class 
verb \emph{vid}.\footnote{ 
 				Mentioning verbs in their present stem forms is not uncommon.
 				The same 6th class verb \root \emph{vid} is mentioned as 
\emph{vinda} in this rule:
 				\dev{anupasargāllimpavindadhāripārivedyudejicetisātisāhibhyaśca} 
(\volcite{}[3.1.138]{shar-2002}).}
 			 The other one may be the 2nd- or the 7th-class \emph{vid}. 
 			 The nominalized form of all of these verbs, using the suffix \emph{GHaÑ}, 
is \emph{veda}. 
 			 This is where the ambiguity appears. 
	 		
(3) In \SuComma{}{}, the verbs are represented as they appear 
	 		in the Paninian \emph{Dhātupāṭha}, with indicatory letters 
(\emph{anubandha}) and meanings attached to the roots:
	 		\dev{vida vicāraṇe, vidḷ lābhe}.\footnote{\Su{6.65.10}{813}.}
	 		However, the citation from the \emph{Jalpakalpataru} presents another 
variant--- 
	 		\dev{vid vicāraṇe vid vindati}\fvolcite{3}[3816]{sena-1928}. 
	 		In both of these variants, the meanings of the verb roots appear. 
	 		The verbs mentioned here are the 7th- and the 6th-class \root \emph{vid} 
respectively. 
	 		The absence of meanings attached to the verb roots in the Nepalese 
version suggests the preservation 
	 		of an older form of the text.\footnote{
	 				Scholars believe that meanings were not initially attached to verb 
roots 
	 				in the original Paninian \emph{Dhātupāṭha} 
	 				and were later additions, possibly by Bhīmasena. See 
\cite[161--163]{card-1976}.}
 				
\subsection{\ldots\dev{dhātvorekārthaḥ/ paścāt padaṃ bhavati}\ldots}
 			
The Nepalese version of this passage significantly diverges from other witnesses.
	For a comparison the readings of four witnesses are provided:
\begin{enumerate}
	\item \dev{pūrvāparam upalakṣya vinda vida ityetayośca dhātvorekārthaḥ  | 
			paścāt padam bhavati āyurvedotpattimayaṃ vivakṣuriti} (Nepalese 
version)
	\item \dev{tatra pūrvāparayogamupalabhya viṃdatītyetayośca 
dhātvorekārthayoḥ 
		paścāt padaṃ bhavati āyurvedotpattimayaṃ vivakṣuriti} 
(\emph{Śuśrutapāṭhaśuddhi} \MScite{London BL IOLR 1842})
	\item \dev{tatra pūrvāparayogamupalabhya vid vicāraṇe vid vindatītyetayośca 
		dhātvoranekārthayoḥ prayogaḥ paścāt pratipattirbhavati 
āyurvedotpattimayaṃ vivakṣuriti} (\emph{Jalpakalpataru})
	\item \dev{vida vicāraṇe, vidḷ lābhe, ityetayośca dhātvoranekārthayoḥ 
prayogāt, 
		tatra pūrvāparayogamupalabhya pratipattirbhavati āyurvedotpattimayaṃ 
vivakṣuriti} \Su{}{}
\end{enumerate}

Evidently, the readings gradually change across these manuscripts, 
with the Nepalese version representing the earliest 
and the last drawing from more recent manuscripts. 
Two key issues emerge: 
firstly, the contradictory readings of \dev{ekārthaḥ} or \dev{ekārthayoḥ} versus 
\dev{anekārthayoḥ}, 
and secondly, the syntactic structure of the sentence. 

(1) The Nepalese version and \MScite{London BL IOLR 1842} uphold 
the former of the contradictory readings, 
while the \emph{Jalpakalpataru} and \Su{}{} adhere to the latter. 
Ḍalhaṇa's commentary provides no definitive insight about his preferred reading. 
Nevertheless, the earlier reading appears more coherent. 
Although the two homonymous verbs bear distinct meanings, 
the context fails to privilege one over the other. 
In essence, we are not able to grasp the \se{padārtha}{relevant meaning} of the 
word \emph{veda} 
by choosing one or the other meanings of the homonymous verb \emph{vid}---be 
it \emph{Sāmaveda} or \emph{Āyurveda}, 
the meaning of the root \emph{vid} does not change. 
The comprehension of \emph{veda} as Āyurveda only occurs 
upon encountering the word “āyurveda” 
itself appearing after two sentences.\footnote{Cf. ‘\dev{āyurvedamicchāma 
ihopadiśyamānam}’ 
	\dev{ityasminnāyurvedaśabdaḥ śrūyate, 
		ato'tra veda āyurveda ityabhiprāyaḥ//} (\Dalhana{6.65.10}{816}).}. 
Hence, the reading \dev{anekārthayoḥ} appears less tenable. 
But if we take the meaning of the word \emph{artha} as “meaning,” 
the phrase \dev{vinda vida ityetayośca dhātvorekārthaḥ} 
does not make sense either because these two homonymous verbs indeed have 
two different meanings. 
However, interpreting the word \emph{artha} as “purpose” offers a more 
plausible explanation---both verbs, 
regardless of their individual meanings, 
serve the same purpose when nominalized with the suffix \emph{GHaÑ}, 
yielding the same form \emph{veda}. 

(2) The other issue pertains to the sentence’s syntax. 
In the Nepalese version, there are two separate sentences: 
“\dev{pūrvāparamupalakṣya vinda vida ityetayośca dhātvorekārthaḥ}” 
and “\dev{paścāt padam bhavati…}”, 
whereas all other witnesses present a unified sentence. 
The fact of having two different sentences in the Nepalese version 
is determined by the use of the nominative case in the word \dev{ekārthaḥ}, 
while the nominative of the second sentence is \dev{padam}. 
In other readings, the genitive case is employed (\dev{ekārthayoḥ}) 
and thus the entire chunk in question constitutes a single sentence. 
Under this interpretation, the sentence conveys the meaning as: 
“when the prior and the subsequent elements are considered, 
after the fact that the roots \emph{vin}d and \emph{vid} have the same 
meaning, 
the clue appears that he wants to talk about the origin of Āyurveda.” 
Here, the word \dev{padam} likely denotes a sign or a clue, 
rather than its conventional meaning “word”, 
which would be incongruous in this context.

\subsection{\dev{yaduktaṃ sādhanaṃ bhavati sa hetvarthaḥ}}

The reading in \Su{6.65.11}{813} is 
\dev{yadanyaduktamanyārthasādhakaṃ bhavati sa hetvarthaḥ}. 
The same reading appears in \MScite{London BL IOLR 1842} 
and the testimonium in the \emph{Jalpakalpataru}. 
\MScite{Jodhpur RORI 20060} reads 
\dev{yaduktamubhayārthasādhanaṃ bhavati sa hetvarthaḥ}. 
Clearly, the reading in the Nepalese version represents 
an older stage of the textual transmission, 
while the vulgate version indicates an attempt to clarify the definition. 
A comparison with the readings found in the texts of earlier listing 
may also suggest that the reading in the Nepalese version is older. 
	
\subsection{\ldots\dev{tathā māṣadugdhaprabhṛtibhirvraṇaḥ klidyate}}

The reading \dev{māṣadugdha}\ldots presents some challenges. 
The Nepalese manuscripts, \Su{}{}, Ḍalhaṇa's reading in \Su{}{}, 
and \MScite{London BL IOLR 1842}---all have \emph{māṣadugdha-}.
However, this reading does not make much sense. 
Should we interpret it as a \emph{dvandva} (beans and milk) 
or a \emph{tatpuruṣa} (milk mixed with beans or bean milk)? 
The first option (\emph{dvandva}) is untenable because 
a wound cannot be moistened with a solid substance like beans. 
The second option is also unlikely for two reasons:

\begin{enumerate}
	\item In the \SS, the word \emph{-prabhṛti-} typically follows 
	more than one item in a \emph{dvandva} compound. 
	Therefore, the reading \dev{māṣadugdhaprabhṛtibhiḥ}, 
	where \emph{māṣadugdha} signifies one item, 
	is statistically improbable.
	
	\item Does \emph{māṣadugdha} mean bean milk? 
	This expression is not found elsewhere. 
	Does it mean milk mixed with beans? 
	If so, it should be used for a specific remedial recipe 
	and mentioned in the \SS. 
	However, there is no such reference to \emph{māṣadugdha} in the text. 
	If māṣadugdha is not mentioned elsewhere, 
	it is unlikely the author intended such a complex example.
	
\end{enumerate}

We conjecture that the original term was not \emph{māṣadugdha}- 
but \emph{cājyadugdha}- or \emph{ājyadugdha}-. 
\MScite{Jodhpur RORI 20060} supports this reading. 
In this manuscript, The character before \emph{jya} is unclear, 
but it is clear that there is a medial vowel \emph{ā} before \emph{jya}. 
The expression \emph{ājyadugdha}- appears in other Sanskrit texts. 
If we read \emph{ājya}- or \emph{cājya}- instead of \emph{māṣa}, 
it would mean “a wound gets moistened with ghee, milk, etc.,” 
which makes much more sense. 
If \emph{ājya}- or \emph{cājya}- was the original reading, 
the scribal error likely occurred quite early 
when the ligature \emph{jya} looked similar to \emph{ṣa}.

\subsection{\dev{samāsavacanaṃ samuddeśaḥ |}}

The reading \dev{samuddeśaḥ} appears in the Nepalese manuscripts 
and \MScite{Jodhpur RORI 20060}. 
Everywhere else, the reading is \dev{uddeśa}, which matches 
the list of \emph{tantrayukti}s provided at the beginning of the chapter. 
The version of the \emph{Jalpakalpataru} includes another variant, 
\dev{samāsakathanam}, which is not supported by any other witnesses.



\section{Characteristics of the Manuscript Transmission}

% Deepro

\newpage
\section{Translation}

\begin{translation}

\item [1] Now we shall explain the chapter called, “the enunciation of the
\se{tantrayukti}{logical methods of the system}.”

\item [3] There are thirty-two logical methods of the system. They are as 
follows: 
\smallskip

\begin{multicols}{2}
    \raggedright
\begin{enumerate}
\item \se{adhikaraṇa}{topic}
\item \se{yoga}{cohesion}
\item \se{padārtha}{relevant meaning}
\item \se{hetvartha}{purpose of a reason}
\item \se{samuddeśa}{mention}
\item \se{nirdeśa}{exposition}
\item \se{upadeśa}{prescription}
\item \se{apadeśa}{statement of reason}
\item \se{pradeśa}{indication}
\item \se{atideśa}{prediction}
\item \se{apavarga}{exception}
\item \se{vākyaśeṣa}{ellipis}
\item \se{arthāpatti}{implication}
\item \se{viparyaya}{contraposition}
\item \se{prasaṅga}{recontextualization}
\item \se{ekānta}{invariable statement}
\item \se{anekānta}{variable statement}
\item \se{pūrvapakṣa}{objection}
\item \se{nirṇaya}{determination}
\item \se{anumata}{consent}
\item \se{vidhāna}{itemization}
\item \se{anāgatāpekṣaṇa}{future reference}
\item \se{atikrāntāpekṣaṇa}{past reference}
\item \se{saṃśaya}{doubt}
\item \se{vyākhyāna}{explication}
\item \se{svasaṃjñā}{field-specific term}
\item \se{nirvacana}{interpretation}
\item \se{nidarśana}{illustration}
\item \se{niyoga}{compulsion}
\item \se{vikalpa}{option}
\item \se{samuccaya}{aggregation}
\item \se{ūhya}{deducible}
\end{enumerate}
\end{multicols}
\bigskip

\item [4] It is said about this, “what is the purpose of these methods?”
The answer is, “cohesion of a sentence and cohesion of
meaning”.\footnote{\Dalhana{6.65.4}{815} explained “cohesion of a
    sentence” as “connecting up a sentence that is not connected,” and
    “cohesion of meaning” as “clarifying  or making appropriate a meaning
    that is implied or inappropriate.”}

\item [5-6] There are \diff{two} verses about this:
  
\begin{sloka}
The logical methods of the system prohibit
statements employed by people who do not speak the truth.
%untrue and unsuitable statements. 
They also bring about the validity of one’s own
statements.  And they also clarify meanings that are stated back to
front, that are implicit, unclear and any that are partially stated.
\end{sloka}

\item [8] Among them, “\se{adhikaraṇa}{topic}” refers to the object, with 
reference to which statements are made, such as \se{rasa}{flavour} or 
\se{doṣa}{humour}.\footnote{The idea here is that “\emph{rasa}” may be the 
topic of a chapter, and statements in that chapter are all understood to be about 
that topic}

\item [9]\label{yoga} “\se{yoga}{Cohesion}” is that by which a sentence
coheres, as when words that are in a reversed order, whether placed close
or apart, have their meanings unified.
%\dev{tailaṃ pibeccāmṛtavallinimbahaṃsāhvayāvṛkṣakapippalībhiḥ |\\
%siddhaṃ balābhyāñ ca sadevadāru hitāya nityaṃ galagaṇḍaroge ||\\}
\begin{quote}
    Sesame oil he should drink, with 
\gls{amṛtavalli}, 
\gls{nimba},
\diff{\gls{haṃsāhvayā}},
\gls{vṛkṣaka}, and
\gls{pippalī}
%with heart-leaved moonseed (Tinospora cordifolia), neem, the plant walking 
%maidenhair fern (Adiantum lunulatum or Adiantum philippense), long pepper, 
%heart-leaf sida, country marrow', and deodar.'' 
\\[2ex]
that is cooked with 
\gls{balā} and \gls{atibalā}, %two mallows
and
\gls{devadāru},
always for a benefit in the case of the disease goitre.
\end{quote}
In this verse, one ought to say, first, “one should drink
cooked\ldots.” However, the word “cooked” is used in the second
line.\footnote{See note \ref{yogan1} on \pageref{yogan1}.} 
Unifying the meanings of words in
    this way, even though they are far apart, is cohesion. 
 

\item [10]\label{padartha} 

The meaning that is conveyed in an \se{sūtra}{aphorism} or a word is 
called \se{padārtha}{relevant meaning}. The meaning that is attached to 
one or more words is the \se{padārtha}{meaning of one or more words}. 
There are innumerable meanings of a word or 
words.\footnote{\label{fn.contrary}Contrary to the translations by previous 
translators 
	\citeauthor{sing-1980} and \citeauthor{shar-1999}, we believe that there is a 
distinction in how the term \emph{padārtha} is employed in the first sentence, 
which serves as the definition of this \emph{tantrayukti}, and in the two 
subsequent sentences, which pertain to the non-technical understanding of the 
word \emph{padārtha}. We have translated the initial use of the word 
\emph{padārtha} as “relevant meaning” and the subsequent use as mere 
“meaning of one or more words.” Without presupposing the distinct usage of the 
word \emph{padārtha} in this passage, the statement, \dev{aparimitāśca 
padārthāḥ}, might seem out of context, as it would not make sense to assert that 
there are innumerable \emph{padārtha}s once the text has already specified 
that 
\emph{padārtha} refers to the particular meaning conveyed in a \emph{sūtra} 
or 
a word. Furthermore, the subsequent illustration featuring three polysemic 
words---\emph{sneha}, \emph{sveda}, and \emph{añjana}---also supports our 
argument. Through these examples, the text advocates for the perspective that 
in 
cases where a word has multiple meanings, only the interpretation that aligns 
with 
the preceding and subsequent elements should be regarded as 
\emph{padārtha}.} 

Where two or three meanings of words such as \emph{sneha}, \emph{sveda} or 
\emph{añjana} appear
to be possible, the relevant meaning is the one that coheres with prior
and subsequent elements.
For example, when it is said that, 
“We are going to explain the chapter on the \emph{veda}-origin” the mind may 
be confused about which “\emph{veda}” will be spoken about. 
\emph{Sāmaveda} 
and so on are the Vedas.
Taking note of the prior and subsequent elements, the two roots \emph{vind} 
”find” and \emph{vid} “know”
have a single meaning.\footnote{The Nepalese
    	text here is hard to follow, and the vulgate has a significantly
    	different reading. But the problem situation seems to be as follows.  The
    	\SS\ opens with a statement saying that it will describe the “origin of
    	the \emph{veda}” (\emph{vedotpatti}).  The problem is, what does this
    	word “\emph{veda}” refer to?  Is it the Veda, as in Sāmaveda?  Or
    	something derived from the roots \root vind or \root vid?  Context
    	(“prior and subsequent elements”) can help us to know that “\emph{veda}”
    	means only “\emph{āyurveda}” and that the \SS\ is talking about the
    	origin of ayurveda, specifically.  This same issue is also addressed by 
    	Ḍalhaṇa at \Su{1.1.1}{1}.} 
Subsequently, the understanding takes place that there is a wish to talk about 
the origin of \emph{āyurveda}.  So that is the meaning of the word.

\item [11] The \se{hetvartha}{purpose of a reason} is a 
statement that becomes
a \se{sādhana}{premise}. e
For example, just as a lump of earth is
moistened by water, so a wound is moistened by substances like milk with
\gls{māṣa}.%\footnote{The way this principle is expressed here seems to be
%    describing the application of a general principle (water makes things
%    wet) to a specific context. We can know the moistening of a wound because
%    we know the more general case of moistening earth. However,
%    etymologically, \dev{hetvartha} does not mean “analogy,” but rather,
%    something like “purpose of the reason.”  The phrase “the sense of cause”
%    that we have used leans on the use of the term in commentaries on the
%    \emph{Aṣṭādhyāyī} (\emph{Kaumudī} on 2.3.23). The vulgate of the \SS\
%    rewrites the principle, making it clearer that the principle means
%    “clarification by analogy.”  Cf.\ also Cakrapāṇi's discussion at
%    \Ca{Si.12.41}{736}, where he explained the principle as using an
%    explanation from one situation to clarify another situation. Cf.\
%    \emph{Arthaśāstra} 5.1.13 \citep[436]{oliv-2013}, which is also unclear.
%    }\q{See also Ḍalhaṇa at \Su{1.1.1}{1}}

\item [12] A \se{samuddeśa}{mention} is a brief statement such as 
“\se{śalya}{spike}”.\footnote{Generally, \dev{śalya} refers to any painful 
foreign body embedded in the flesh that requires surgical removal.}

\item [13] A \se{nirdeśa}{exposition} is a detailed statement. For example, “in 
the body or exogenous”.\footnote{This is a reference to \Su{1.26.4}{121}
    where \dev{śalya} is described in more detail as being of two kinds.} 

\item [14] “Prescription” (\emph{upadeśa})\sse{upadeśa}{prescription} refers 
to statements like 
``it should be 
this way.'' For example, one should not stay awake at night; 
one should not sleep during the day.  

\item [15] “\sse{apadeśa}{statement of reason}Statement of reason”
(\emph{apadeśa}) refers to statements like “this happens because of this.” For
example, in the sentence “Sweet substances increase phlegm,” the reason is
stated.\footnote{A techical term also in Nyāyaśāstra \citep[54]{jhal-1978}.}

\item [16] Substantiation of the subject matter through past evidence is
“\se{pradeśa}{indication}.” For example, he pulled out Devadatta's
\se{śalya}{splinter}, therefore he will pull out Yajñadatta's.

\item [17] Substantiation of the subject matter through a future event is
“\se{atideśa}{prediction}.” For example, if his wind moves upwards, that
will cause him to have colic.”\footnote{A techical term also in
    Nyāyaśāstra \citep[6--7]{jhal-1978}.}

% got to here with DW

\item [18] A deviation after generalization is \se{apavarga}{exception}. For 
example, those afflicted by poison should not go through sudorific treatment 
other 
than the cases of poisoning by urinary worms.

\item [19] \se{vākyaśeṣa}{Ellipsis} refers to an 
unstated word that 
completes a sentence. For example, despite not mentioning the word 'person',
when mentioning someone as 'the one having a head, hands, feet, flanks, and 
abdomen,' it's apparent that the reference is to a person. 

\item [20] Implication\sse{arthāpatti}{implication} refers to an unstated idea 
that becomes evident through context. For example, when one said, “We will 
eat rice” it becomes evident from the context that he did not wish to drink 
gruel. 

\item [21] When there is the reversal of it, it is \se{viparyaya}{contraposition}. 
For example, when it is said, "Weak, dyspneic, and fearful people are difficult to 
treat," the converse holds true: "Those who are strong and so on are easily 
treatable." 

\item [22] \se{prasaṅga}{Recontextualization} refers to a concept common to 
another section. For example, a concept belonging to another section is brought 
up 
by mentioning it repeatedly throughout. 

\item [23] \se{ekānta}{Invariable 
statement} is one that 
is stated with certainty. For example, \gls{trivṛt} causes purgation; 
\gls{madana} induces vomiting.

\item [24] Variable statement (\emph{anekānta})\sse{anekānta}{variable 
statement} is one that is true in one way in 
some cases and in another way elsewhere. For example, some teachers identify 
the main element as substance, others as fluid, some as semen, and some as 
digestion.\q{See chapter 40 of Sūtrasthāna.}

\item [25]  A \se{pūrvapakṣa}{first point of view} is something stated
with certainty.  For example, how are the four types of diabetes caused
by wind incurable?\footnote{The adverb \dev{niḥsaṃśayam} is problematic:
    the example expresses a query or doubt, the opposite of certainty, which
    is answered in the next passage.  It would seem to make more sense to
    read something like \dev{yas tu sasaṃśayam abhidhīyate sa pūrvapakṣaḥ},
    but our manuscripts are unanimous in their reading.}

\item [26] Its answer is determination. For example, afflicting the body and 
trickling downwards, it creates urine mixed with fat, fatty tissues, and 
marrow.\q{vasā / medas / majjan} Thus, those caused by wind are incurable. 

\item [28]  \sse{anumata}{consent}Consent (\emph{anumata}) refers to 
others' opinion that is not 
rejected. For example, when the assertor says that there are six flavours and 
that somehow gets accepted with affirmation, it is termed consent.

\item [29] Itemization (\emph{vidhāna})\sse{vidhāna}{itemization} refers to 
sequentially ordered 
statements within a chapter. For example, the eleven lethal points of thigh are 
mentioned sequentially in a chapter.

\item [30] A statement like “Thus will be stated” is \se{anāgatāpekṣaṇa}{future 
reference} such as when he says in the \emph{Sūtrasthāna}, “I will mention it in 
the \emph{Cikitsāsthāna}.” 

\item [31] A statement like “Thus has been stated” is 
\se{atikrāntāpekṣaṇa}{past reference} such as when one says in the 
\emph{Cikitsāsthāna}, “As mentioned in the \emph{Sūtrasthāna}\ldots.” 

\item [32] An indication pointing to causes on both sides is
\se{saṃśaya}{doubt}. For example, a blow to the
\se{talahṛdaya}{sole-heart}\footnote{\dev{talahṛdaya} is one of the
    muscle-group of lethal points mentioned in \Su{3.6.7}{370}.}  is fatal,
    whereas cutting hands and feet is not fatal.

\item [33] An elaborate description is \se{vyākhyāna}{explication}. For 
example, 
the twenty-fifth entity, \se{puruṣa}{person}, is being explicated here. Thus, no 
other Āyurvedic texts discuss  entities beginning with matters. \q{Does 
bhūtādi a compound or it means ahaṅkāra or ego?}

\item [34] A \se{svasaṃjñā}{field-specific term} is uncommon in other fields of 
studies. The term used in one's own systems is called field-specific term, such as 
in this system, \se{mithuna}{pair} denotes honey and ghee, and 
\se{mithuna}{triad} denotes ghee, sesame oil and fat.\q{triad? --DW}

\item [35] A customary potrayal is \se{nirvacana}{interpretation}. For example, 
one goes along the shade fearing heat. 

\item [36] Providing examples is \se{nidarśana}{illustration}. For example, just 
as fire spreads rapidly in a dry forest when accompanied by wind, a wound 
intensifies affected by wind, bile, and phlegm.  

\item [37] A statement like “This is the only way\ldots” \ldots\ 
\se{niyoga}{compulsion}. For example, one should consume only a healthy 
diet.     

\item [39] A statement like “This and this\ldots” is
\se{vikalpa}{option}. For example, in the section on meat, the major ones
are blackbuck, deer, quail and partridge.\footnote{The example here
    matches \dev{samuccaya} (next text), not \dev{vikalpa}.  There seems to
    have been a metathesis of terms. \citet[1005, footnote
    6]{susr-trikamji1945} notes that this text and the next have been swapped
    in the Calcutta edition that includes Hārāṇacandra's commentary
    \volcite{2}{bhat-1908}, in the same way as in the Nepalese version.}

\item [38] A summarized statement is
\se{samuccaya}{aggregation}.\footnote{As stated in the previous footnote, the 
example here is of \dev{vikalpa}, not \dev{samuccaya}.
%    The term \dev{samāsavacana}, which
%   means more or less the same as \dev{saṃkṣepavacana}, has already been
%    used in tantrayukti \dev{samuddeśa}
    } For example, let there be rice
    with meat broth, rice with milk, or burley with ghee.
    
    \begin{quote}
        A meaningful reading of these two rules would be
    
    39 idaṃ vedaṃ veti vikalpaḥ / yathā rasodanaḥ kṣīrodanaḥ saghṛtā vā 
    yavāgūr bhavatv iti //
    
    38 saṃkṣepavacanaṃ samuccayaḥ / yathā māṃsavarge eṇahariṇalāvatittirāḥ 
    pradhānā iti
    \end{quote}
    

\item [40] 

What is not explicitly stated but can be understood through discernment is 
\se{ūhya}{deducible}. For example, in the section on rules of foods and drinks, 
four types of foods and drinks are mentioned— \se{bhakṣya}{masticable}, 
\se{bhojya}{edible}, \se{lehya}{suckable}, and \se{peya}{drinkable}. Thus, 
while four types are needed to be stated, two types are actually mentioned. 
Here it is deducible that in the section on foods and drinks, by specifically 
mentioning two types, the four types are also mentioned. Furthermore, a 
masticable item is not excluded from the category of food because it shares the 
same characteristic of solidity. A suckable item is not excluded from being 
classified as a drink because it shares the same characteristic of liquidity. Four 
types of aliments are rare. They are usually just twofold. Therefore, lord 
Dhanvantari says “Twofold is popular”.   


\end{translation}
