% !TeX root = incremental_SS_Translation.tex
% Paras

\chapter{Cikitsāsthāna  4:  On the Treatment of Wind 
    Diseases}

\section{Literature} 

Meulenbeld offered an annotated overview of this chapter and a bibliography
of earlier scholarship to 2002.\fvolcite{IA}[265--266]{meul-hist} 

\section{Translation}

\begin{translation}
    
    \item [1] 
    Now we shall describe the treatment of wind diseases.
    
    %\item [2] 
    weight

    \item [3] When the wind enters the stomach, one should sequentially give to
the patient, \diff{who has vomited}, the \se{yoga}{formulation} with
\se{ṣaḍdharaṇa}{six-units},  together with tepid water, for seven
nights.\footnote{The vulgate has the reading \dev{chardayitvā} which means
    "after making [him] vomit". Thus, vomiting is a part of the treatment.
    Whereas \dev{chardita} in the H manuscript is ambiguous:  vomiting may be 
    part of the treatment or a symptom of the ailment.
    
The expression “six units” refers to the six ingredients listed in the
next passage.  \Dalhana{4.4.3}{420} noted that \dev{dharaṇa} in this
context means a particular weight characterized as equivalent to 21
medium-sized \gls{niṣpāva}.  \citet[303]{shar-1999} proposed that that
the formulation contains six ingredients each the weight of a
\emph{dharaṇa}.  See \Su{4.31.7}{508} where the term \dev{dharaṇa} is
defined in terms of other weights.  (In epigraphical Sanskrit, a
\dev{dharaṇa} may be a silver or gold coin \citep[91]{sirc-1966}.)

 \AH\ \Ah{4.21.14}{723} is the same verse, mutatis mutandis, but the editor 
 noted 
 (f.n.\,6) a variant reading \dev{ṣaṭcaraṇa} in the commentary of Śrīkaṇṭha.  
 There seems to be some confusion about this expression.

Ḍalhaṇa also noted that \dev{sukhāmbu} (“pleasant water”) means “slightly 
warm water.”}

    \item [4]
    
    “Six-unit” is traditionally the formulation that is 
\gls{citraka}, \gls{indrayava}, \gls{pāṭhā},
\gls{kaṭukā}, \gls{ativiṣā}, and \gls{abhayā}. It cures serious
diseases.



    \item [5]
    When the wind has entered the \se{pakvāśa}{abdomen} one should treat it 
    with an oil purge. One should also treat it with cleansing enemas and very 
    salty foods.



    \item [6] When the wind has entered the bladder, a cleansing enema method
should be carried out. And once an inflamed wind is in the ears and the like,
a procedure that destroys wind should be done.

    \item [7]
    When the wind has reached the skin, flesh, and \diff{blood}, one should do 
    an 
    \se{abhyaṅga}{oil rub}, apply a \se{upanāha}{poultice},  
    \se{mardana}{rubbing} and \se{ālepana}{ointments}. One should also 
    perform blood-letting.\footnote{On the translation of methods of 
    medical touch, such as \dev{abhyaṅga} and \dev{saṃvāhana}, see 
    \cite[122--131]{broo-2021}.  \dev{mardana, unmardana} mean “pressing or 
    vigorous rubbing.”  The vulgate includes 
    \se{sirā}{ducts} as an added place that wind can enter.}

    \item[8]
    When the wind has got into the ligaments, joints, and bones, an expert 
    should apply  \se{sneha}{oleation},  a \se{upanāha}{poultice},
    \se{agnikarma}{cauterization}, binding, and \se{unmardana}{rubbing}.



    \item [9] When the wind is deep within the bone, then a strong
physician should insert a \se{nāḍī}{tube} into the bone, which has
been split open by \se{pāṇimantha}{manual agitation}, and suck out
the wind.\footnote{Although gramatically the expression “which is
    split” could be construed with “wind,” it has to be understood
    here as referring to the bone.  The word order is not obvious. 
    \Dalhana{4.4.9}{420} interpreted \dev{pāṇimantha} as the name of a
    particular awl and described the bone being pierced by this awl so
    that a double-headed tube can be inserted into the resulting
    opening.
    
    This verse is in \emph{na} vipulā metre.}
         
         
          
    % Ask Dominik Sir about the vipula types of anustup metre. Refer Ch. 106 of the Gita Darpana book of Svami Ramasukhadasa. It has many vipula variants, etc. of the anustup.        
    % My earlier comment in footnote: Also, the metre of this hemistich of the verse is not according to the \emph{anuṣṭup} metre. This gives us more reason to think that the reading \dev{asthīni} is corrupted.   

    \item[10ab]
    When the wind has reached the semen, one should perform the treatment 
    for the defects of the semen.\footnote{Ḍalhaṇa comments 
    \citep[421]{vulgate} that this treatment for the defects of the semen is 
    mentioned [earlier] as the \dev{śukraśoṇitaśuddhi}, the purification of the 
    semen and the blood. This is the \emph{Śārīrasthāna} Ch. 2, 
    \dev{śukraśoṇitaviśuddhi}.}
    
    


    \item[10cd--11] 
    
When the wind has reached the whole body, an intelligent person should
conquer it by means of immersion, \se{kuṭī}{sauna}, \se{karṣū}{trench
    sweating}, \se{prastara}{blanket sweating}, oil massage, enema, and
blood-letting.\footnote{These forms of sweating treatment are
    described in the \CS\ (\Ca{1.14.39--63}{90--92}).
    
Regarding blood-letting, \Dalhana{4.4.11}{421} commented  that because
the verse has the plural form \dev{sirāmokṣaiḥ}, five blood vessels
have to be drained of blood if the wind is not pacified by oil
massage, etc.} Or, if is located in a single limb and is stuck there,
a thoughtful physician may conquer it with
cow-horns.\footnote{\dev{śṛṅga} “cow-horns” refers to bloodletting by
    horn; see the description at \SS\ \Su{1.13.5}{55}.}
        
       
            
    \item[12] 
    
    Or, if it is mingled with \se{balāsa}{phlegm}, bile, and
blood, the physician should treat it with 
remedies that are \se{avirodhin}{compatible}.\footnote{The word 
\dev{balāsa} is used here in the
    slightly unusual meaning “phlegm;” see \Dalhana{1.45.70,
    6.61.33}{202, 802} and \cite[553]{josi-maha}.} However, when the
    wind is inactive, he should perform blood-letting many
    times.\footnote{We read \dev{suptavāte} with witness H, but
        Ḍalhaṇa glossed \dev{supti-}, the reading of the vulgate, “it is
        wind characterized by \se{supti}{drowsiness} caused by a covering
        of blood.”}

    \item[13]
    
     And one should lick the milk cooked in \gls{pañcamūlī}
together with salt and \se{āgāradhūma}{soot from the chimney},
mixed with oil and also a \se{rasa}{juice} that has the sourness of
a fruit.\footnote{The vulgate reading \dev{dihyāt} for the Nepalese 
\dev{lihyāt} changes the meaning to “one should smear.”
    
    \Dalhana{4.4.13}{421} glossed \dev{pañcamūlī} as optionally the first  or 
    the second five roots.  On this therapy, cf.\ Cakrapāṇi's commentary on 
    \Ca{1.5.3}{36} for a similar therapy.
    
    The “juice” (\dev{rasa}) was glossed by Ḍalhaṇa as specifically
being a \se{māṃsarasa}{meat broth}.   He said that the sourness
may come from fruits such as pomegranate.
    
  \dev{rasāmla} may mean a vinegar made from fruit  
\citep[70]{moni-sans}, so the expression \dev{phalāmlo rasaḥ} in the
text here may mean a vinegar made from sour fruit. Cf.\
\dev{dhānyāmla}.}
    

    \item[14--16ab]
    
    Alternatively, cereal soup with a good amount of ghee is a
wholesome food that repels wind. However, a \emph{śālvala} poultice is 
famous 
as being a lukewarm and very salty substance made of the
\gls{kākolī} group combined with an item that repels wind and
together with all the sour drugs and the meat of creatures from
marshes and water that have all the oils.\footnote{Cf.\
    \dev{sālvaṇa} “sweat from a poultice” in \cite[898]{josi-maha}.
    \AS\ \As{1.26.3}{188} describes a poultice called “sālvala” made
    with numerous ingredients (the commentator Indu elaborates,
    p.\,189). \cite[1068]{moni-sans} glossed \dev{śālvaṇa} as “a
    poultice, cataplasm,” based on the \SS.}
            %\item[16ab]
        One should always apply a bandage with that to people who are 
        ill with wind. 
   
   \item[16cd-18ab]
   
   One should tightly bind someone who is bent, afflicted by pain,
   or whose limbs are \se{stabdha}{numb}, with a 
   \se{paṭṭa}{strip of cloth} made of bark, cotton or \se{ūrṇa}{wool}.
   
Alternatively, after massaging the affected body part and applying the
\emph{śālvala} poultice on it, one should insert it into a sack made
of the hide of a 
\gls{biḍāla}, 
\gls{nakula}, or 
\gls{udra}, or else of
\gls{mṛga}.\footnote{For \dev{goṇī}, \citeauthor[367]{moni-sans} recorded
``sack'' and ``torn or ragged clothes;'' it may have been a sling or
similar support.}
    
    
    
\item[18cd--19]
     
    Vomiting and an errhine done skilfully alleviate the wind that has
entered the chest, \se{trika}{between the shoulder-blades}, the
shoulders, or the nape of the neck.\footnote{On \dev{trika}, see
    \volcite{1}[387]{josi-maha}, citing \Dalhana{3.6.26}{374} “the
    junction between the shoulder-blades and the neck.”} The  wind  located in 
    the head is defeated by blood-letting and by the 
    \se{śirobasti}{application of oil to the head}.
    
    
    
     \item[20--21ab] 
     
     In that context, one should let the oil remain carefully for 
one thousand \se{mātrā}{measures}.\footnote{\Dalhana{4.2.20}{422}
    interpreted \dev{mātrā} as a measure of time, citing an
    unattributed verse defining it as the time of a blink, a snap of
    the fingers or the utterance of a single vowel. The expression
    might possibly be taken to refer to a measure of the oil's
    volume.} Only an \se{basti}{enema} can curtail the wind,  whether
    it is throughout the whole body or in just one limb.  Its
    \se{vega}{force} is like the wind.\footnote{This phrase is
        awkward. The idea here seems to be that an enema decisively stops
        the wind.  The vulgate revised this to make it more obvious:
        “only an enema can block the force of the wind, like a
        mountain.”}
    
% got to here 2024-05-29

    \item[21cd--26]
    
Oils, perspiration, oil massage, enema, unctuous purging of the
bowels, \emph{śirobasti}, oiling the head, unctuous smoke, gargling
with lukewarm water, errhines, an \se{kalka}{oily
    paste},
    %\footnote{\dev{kalka} ``oily paste'' here might also refer to
    %\gls{vibhītaka}.}, 
    milks, meats,\footnote{The plural indicates milk and
        meat from various animals.} soups, oils,\footnote{This is the second
            occurrence of the word \dev{snehāḥ} in this sentence. This seems to be
            an anomaly.} any unctuous substance, unctuous and salty meals that
            are made sour by fruits, bathing with lukewarm water, massages,
            saffron, \gls{aguru}, \gls{patra}, \gls{kuṣṭha}, \gls{elā},
            \gls{tagara}, garments made of silk, wool, and fur, soft cotton
            garments, inner rooms with sunlight, no wind flow, and a soft bed,
            taking the warmth of fire, and celibacy, etc., are to be collectively
            employed for patients with wind diseases.

    

    \item[27]
    
One should take 12 grams each of \se{kalka}{pastes} of \gls{trivṛt},
\gls{dantī}, \gls{śaṅkhinī}, \gls{suvarṇakṣīrī}, \gls{triphalā}, and
\gls{viḍaṅga}, with 48 grams of \gls{tilvaka}-root and \gls{kampilya},
and six kilos of the juice of \gls{triphalā} and curds and three kilos
of ghee.\footnote{The measure of 12 grams is expressed as an
    \dev{akṣa}.  See \cite[263]{wuja-2003}, which is based on the
    fourteenth-century \emph{Śārṅgadharasaṃhitā}.  The measure of 48\,g is
    a \dev{bilva}, more commonly called a \dev{pala} (\emph{idem}).  Six
    kilos is two \dev{pātra}, which is a liquid measure of approximately 
    2$\times$3\,kg (\emph{idem}).  There is one \dev{pātra} of ghee.
    \Dalhana{4.4.27}{422}
                commented on the measurements mentioned in this recipe.} One 
                should
                then mix these ingredients all together and cook the mixture. This is
                called the \gls{tilvaka}-aghee. They prescribe this oily bowel purge in
                cases of wind disorder.
                
                 This procedure for making \gls{tilvaka}-ghee
                    should followed for making \gls{aśoka}-ghee and
                    \gls{ramyaka}-ghee.

%Explaining the production of sahasrapāka [thousand-fold processed oil], the 
%author says: "Now," and so forth.
%One should drench the oil in an amount [measured by] the pañcamūla [five 
%roots], according to the dimensions of the ground area. "Dark" means a region 
%of black soil. "Resided" means remained. "Having removed" means having 
%broken apart. "The ground has ceased" means the heat in the soil has been 
%pacified.
%Regarding "the group beginning with vidārigandhā," and so forth: one should 
%sprinkle the ground with a hundred jars of oil cooked with the group beginning 
%with vidārigandhā—prepared in the manner of a paste (kalka) and decoction 
%(kaṣāya)—along with an equal amount of milk.
%In this context, "oil" refers to sesame oil (tilataila). "Anointed" means it has 
%been moistened with oil and milk. In the phrase "in a large cauldron," the 
%singular number is used to represent the entire class (jāti). Therefore, the 
%process can occur in one, two, or three [cauldrons]. "One should sprinkle" 
%means one should drench it.
%The medicines that pacify wind (vātahara) are those beginning with 
%bhadradāru; their decoction is prepared according to the standard rules for 
%decoctions. The meat broth is prepared according to the rules for broth from 
%animals of arid lands (jāṅgala). The milk is cow’s milk, and the acidic liquid is 
%fermented rice water (kāñjika).
%Regarding the phrase "by a thousand parts": here, a "part" refers to a 
%quarter-share. Thus, there are a thousand parts each of the wind-pacifying 
%decoction, meat broth, milk, and acidic liquid. By this, it is meant that for each 
%individual cooking (pratipāka), one uses a quarter-part of the decoction and 
%other ingredients relative to the amount of oil, so that collectively they equal 
%the amount of oil. Gayī states that one should cook it a thousand times in this 
%manner so that it becomes a "thousand-fold processed" oil.
%Jejjaṭa, however, believes that one should throw in a thousand quarter-parts of 
%the decoction and other ingredients collectively relative to the oil all at once 
%and 
%cook it thoroughly. He holds that "thousand-fold processed" is merely a 
%technical name, for if there were a thousand individual cookings, the oil would 
%be completely consumed.
%This [view of Jejjaṭa] was refuted by Gayadāsa as follows: it is not the case 
%that all the oil is consumed; rather, only a tiny amount is lost, because during 
%each cooking, a quarter-part of milk is introduced. For wealthy kings, such a 
%minor loss is negligible. Therefore, since both great masters have explained it, 
%both methods of cooking are authoritative.
%Although someone has said that one should cook it a hundred times with a 
%thousand parts of liquid, this should not be accepted, as it is a careless 
%explanation.
%In this context, "addition" (prativāpa) means the throwing in of the fragrant 
%paste. "Himalayan" refers to those substances produced in the Northern 
%region 
%(uttarāpatha), such as musk, śaṭī, costus, spikenard, sarala pine, deodar, 
%murā, 
%and so forth. Those from the Southern region (dakṣiṇapatha) include 
%sandalwood, nutmeg, kaṅkōla, cloves, and so forth. These should be made 
%into 
%a paste in an amount equal to a quarter of the oil.
%Regarding "and those that destroy wind": these are also those found in the list 
%of wind-pacifying groups, such as śatapuṣpā and miśri. This is done to produce 
%an extraordinary effect because of the inconceivable power [of the oil], such 
%as 
%[curing deafness so one can hear] the blowing of a conch.
%Others, however, read: "one should sprinkle with a hundred jars of oil along 
%with milk and fat." This should not be read, as it was not explained by the 
%teachers who were discerning analysts.
%Explaining the production of śatapāka [hundred-fold processed oil], the author 
%says: "In this way, by a hundred parts," and so forth. This is easily understood.




    \item[28] 
    
Take wood that has been used in instruments for grinding sesame seeds.
Chop those that have been used  to grind the sesame seeds for a long
time into tiny pieces.  Next, crush them, put them into a large pot of
water and make them into a paste. Then, collect the oil from the
surface of the water, either with a cup or by hand, and cook it into a
\se{pratīvāpa}{mixture} that has herbs that destroy wind added to it,
following the \se{oil-cooking}{snehapāka} method.\footnote{This
    special method of preparing oil in three grades is described at \SS\
    \Su{4.31.11}{509}. See also \volcite{1}[944]{josi-maha}.} This is the
    \se{aṇutaila}{fine oil} that is mentioned in wind disorders. It is
    called \se{aṇutaila}{fine oil} because it is produced from ingredients
    of \se{aṇu}{fine} oil.



    \item[29] 
    
% Dharmamitra:
%    Alternatively, one should burn a patch of ground with a large quantity of 
%wood from the five great roots (mahāpañcamūla). After one night, once the 
%fire 
%has died down, one should remove the ashes and drench the cleared soil with 
%a 
%hundred jars of oil prepared with vidārīgandhā and other bs. Having left it for 
%one night, one should then collect as much of the earth as has become oily 
%and 
%soak it in a large cauldron of water.
%    
%    Whatever oil rises to the surface there should be skimmed off with both 
%hands and stored in a well-secured place. That oil should then be processed a 
%thousand times (sahasrapāka) with a thousand parts of a decoction of 
%wind-subduing (vātahara) drugs, meat broth, milk, and acidic substances.
%    
%    The duration of the cooking process should be as long as necessary. During 
%this, the additives (prativāpa) should include aśvagandhā from the Himalayas 
%and the southern regions (dakṣiṇāpatha), along with other wind-subduing 
%agents.
%    
%    As the process reaches successful completion, one should blow conch 
%shells, hold up parasols, beat drums, and fan it with yak-tail whisks. Then, 
%once 
%it is perfectly prepared, it should be removed from the heat and stored in a 
%golden, silver, or earthen pitcher.
%    
%    This is the "thousand-times processed" (sahasrapāka) oil, which possesses 
%irresistible power and is worthy of kings. In the same manner, oil processed a 
%hundred times is called śatapāka.
%    
    
    
Alternatively, one should burn a patch of ground using many sticks of
wood belonging to the \gls{mahāpañcamūlī}.  After one night, when the
fire has gone out, the ash should be removed.  Then, when the earth has been 
extinguished, it should be soaked with a hundred pots of sesame oil cooked with
herbs like \gls{vidārigandhā}.  It should be left in that condition
for one night.\footnote{\tvolcite{1B}[104, n.\,123]{meul-hist} noted
    that the \CS\ \Ca{6.26.23, 74}{598, 602} stated that the group
    starting with \gls{vidārigandhā} (\dev{vidārigandhā}) is the same as
    the \dev{hrasvapañcamūla} or the \dev{sthirādi} groups.} Next, one
    should put all the earth that is oily in water and then pour it into a
    large cauldron. %
    %  \footnote{The reading in H,
    %            \dev{kaṭāhebhyaḥ siṃcet}, does not make sense here.
    % Thus,
    % we have
    %            accepted the vulgate reading \dev{kaṭāhe 'bhyāsiṃcet}
    % for
    % the
    %            translation.}
    The oil that rises to the surface should be taken out with both
    hands and stored in a private place.  Next, one should thoroughly
    cook that oil a thousand times for as long as possible %
    %  \footnote{The
    %                phrase "\dev{yāvatā kālena śaknuyāt paktum}"
    % appears
    % as a part of
    %                a new sentence in H. But, we should take it to be
    % a
    % part of the
    %                earlier sentence for it to make proper sense.}
    with a thousand parts of a decoction of wind-alleviating herbs,
    meat soup, milk, and \gls{amla}.\footnote{On
        “\se{sahasrapāka}{thousand cooking}” see
        \volcite{1}[891]{josi-maha}, \dev{sahasravāraṃ pācitaṃ tailam}
        “oil cooked a thousand times” citing the present passage.  The
        \CS\ has a similar sense (\Ca{3.6.16}{256}).
                    
        Ḍalhaṇa commented that the word \egls{amla} here means
\dev{kāñjika}, the sour, fermented water drained after boiling
rice \citep[423]{vulgate}.} An admixture is added to the oil
containing the \se{haimavata}{Himalayan herbs}, herbs of the
southern region, \gls{aśvagandhā}, and other herbs that reduce
wind.\footnote{\Dalhana{4.4.29}{423} commented that
    \dev{haimavatāḥ} refers to the herbs that grow in the northern
    region.  But the term can also be a more specific taxon, i.e.,
    \gls{haimavata}.}
    
While the oil is being cooked, one should blow conch shells loudly,
open umbrellas, beat kettle drums and wave chowries. Then, when the
oil is perfectly cooked, it should be removed from the heat and poured
into a golden or silver pot and stored. This
\se{sahasrapāka}{thousand-cooking} oil has irresistible potency and
is fit for kings. 

Similarly, something cooked a hundred times with a hundred ingredients
is called a “\se{śatapāka}{hundred-cook}.”


      \bigskip       
\begin{tt}
    \marginpar{Draft tr.\ from here}

    \item[30]
    One should collect fresh leaves of \gls{gandharvahasta}, \gls{aṭa}, 
    \gls{ruṣkaka}, \gls{muṣkaka}, \gls{naktamāla}, \gls{pūtikā}, and 
    \gls{citraka}.\footnote{In H, the ending \dev{nām} should be \dev{ṇām} 
    due to sandhi.} These leaves should be completely pounded along with salt 
    in a mortar. This mixture should be put in a pot filled with 
    oil\footnote{\dev{snehaghaṭa} can also mean a pot filled with ghee}. It 
    (pot) should be smeared\footnote{The H or vulgate do not specify with 
    words that it is the pot to be smeared. But, it is to be understood.} with 
    cow-dung. Thereafter, the pot should be heated.\footnote{The word 
    \dev{dāhayet} usually refers to burning, but sometimes it can refer to 
    heating.} This (resultant) is the \textit{patra-lavaṇa} (leaf-salt) that is 
    mentioned in wind disorders. 

    \item[31]
    In the same way, one should pound the stalks of \gls{snuhā}\q{Euphorbia 
    Antiquorum (Antique spurge)} and eggplants smeared with salt and fill a pot 
    with it.\footnote{In H, there should be a \textit{visarga} after 
    \dev{lavaṇā}.} In that pot, one should add ghee, oil, fat, and marrow. Then, 
    one should smear it\footnote{As earlier, the pot should be smeared with 
    cow-dung.} and heat it as earlier. This (resultant) is the 
    \textit{sneha-lavaṇa} (fat-salt) that is mentioned in wind disorders.

    \item[32]
    One should collect the fresh fruits, roots, leaves, and branches of all the 
    twenty [herbs]: \gls{gaṇḍīra}, \gls{palāśa}, \gls{kuṭaja}, \gls{bilva}, 
    \gls{arka}, \gls{snuhā}, \gls{pāmārga}, \gls{pāṭalī}, \gls{pāribhadraka}, 
    \gls{nādeyī}, \gls{kṛṣṇagandhā}, \gls{nīpa}, \gls{nirdahī}, \gls{ānyāṭa}, 
    \gls{rūṣaka}, \gls{naktamāla}, \gls{pūtīka}, \gls{bṛhatī}, \gls{kaṇṭakārikā}, 
    \gls{bhallātaka}, \gls{aśoka}, \gls{vaijayantī}. One should then mix them 
    with salt and heat them as earlier.\footnote{It is to be understood that all 
    these fresh branches, leaves, fruits, and roots of the herbs should be 
    completely pounded together with salt. The mixture should then be put into 
    a pot filled with oil or ghee. The pot should be smeared with cow-dung and 
    then heated.} The oil on top should be poured out completely with the salty 
    mixture intact [at the bottom]. This mixture should be cooked thoroughly. 
    The admixture added to it consists of \gls{pippalī}, etc. This (resultant) is 
    the salt called \textit{kalyāṇaka} that is mentioned in wind disorders and in 
    meals and drinks for the patients troubled by \textit{plīhāgnisaṃga}, 
    indigestion, loss of appetite, and piles.


    Thus ends the fourth chapter on the treatment of wind diseases. 

\end{tt}    

\end{translation}
