% !TeX root = incremental_SS_Translation.tex
\chapter{Kalpasthāna 3: Poisonous Insects and Animals}

\section{Literature}

Meulenbeld offered an annotated overview of this chapter and a
bibliography of earlier scholarship to
2002.\fvolcite{IA}[291--292]{meul-hist}


\section{Translation}

\begin{translation}
  
  \item[1] 
  
And now we shall explain the \se{kalpa}{formal procedure} that is the
required knowledge about mobile poisons.\footnote{In contrast to
    stationary, plant poisons.  No reference is made to Dhanvantari
    \citep[see][]{birc-2021}.}\q{Come back to the issue of "kalpa".  Look
        up passages in the Kośa.}

  \item[3] 
  
The full explanation about the sixteen \se{adhiṣṭhāna}{carriers} of
the mobile poisons, that have been mentioned by me in brief, will be
stated.\footnote{“Carrier” for \se{adhiṣṭhāna}{base, foundation} aims
    to capture the idea that the author will describe the creatures in
    which poisons inhere.}
  
\item[4] 

In that context, they are:\footnote{The content of this section is
    presented as a table, for clarity for the contemporary reader and
    mindful of the theoretical issues surrounding notational variation,
    including the “symbolic rewriting” and the modification of “expressive
    capacities” discussed by \citet[321\,ff]{saru-2016}.  For further
    discussion, see \cite[81--83]{wuja-2021}.}
  \begin{multicols}{2}
      
  \begin{enumerate} 
      
      \item gaze       				%1
      \item  breath,       			%2
  \item teeth, 						%3
      \item nails, 					%4
      \item mouth,      			%5
      \item urine,					%6
      \item feces,					%7
    \item \diff{menstrual blood},  %8
    \item semen,					%9
    \item \diff{tail},				%10  lāṅgūla 
   \item \diff{contact with saliva}, % 11 lālāsparśa 
  \item \se{mukhasaṃdaṃśā}{nipping with the mouth},  %12
  \item \se{avaśardhita}{fart},\footnote{This interpretation  comes from
    Ḍalhaṇa on \Su{5.3.4}{567}, but he reads \dev{viśardhita}.} %13
   \item \diff{anus},\footnote{\Dalhana{5.3.4}{567} noted this 
   reading.} % 14
\item  bones, %15
\item  bile, %16
\item  \se{śūka}{bristles}, and  %17
\item corpses.\footnote{This list has grown in transmission by two items.}  
%18
\end{enumerate} 
\end{multicols} 

\bigskip

\item [5]  

In that context, 

\noindent
{\centering \begin{longtable}{ 
>{\raggedright\arraybackslash}p{.33\textwidth} 
>{\raggedright\arraybackslash}p{.61\textwidth}} 
\toprule 
\emph{location of the poison} &  \emph{creatures}\footnotemark\\ 
\midrule 
\endfirsthead  
\toprule  
\emph{location of the poison} & \emph{creatures}\\ 
\midrule 
\endhead
\bottomrule 
\endfoot
\footnotetext{Many of these names are mere dubious  placeholders.}
 in their breath and gaze  & divine snakes \\[2ex]  
 in their fangs  & the ones on earth\footnote{Ḍalhaṇa on 
    \Su{5.3.5}{567} cited the otherwise unknown authority Sāvitra on the
    topic of  poisonous snakes
\citep[IA, 377, IB 497, n.\,105]{meul-hist}.} \\[2ex] 
 
 in their nails, mouths and fangs  a & cats, dogs, monkeys,
\se{nara}{men},\footnote{Probably dittography from  the previous
    word, \se{vānara}{monkey}. But it is supported in both Nepalese
    witnesses, so it must go back to an earlier exemplar.}  crocodiles,
    frogs, \gls{pākamatsya},\footnote{MS KL 699  separates
        the words \dev{pāka} and \dev{matsya} with a daṇḍa,  indicating that
        the scribe thought they were separate terms (see \pageref{pakamatsya}).} 
        monitor lizards, 
        \glspl{śambūka}, \se{pracalāka}{`poisonous
            snakes'},\footnote{\emph{Arthaśāstra} 14.1.14, 23
            \citep[448]{oliv-2013}, where it might also be a  chameleon; but the
            latter are not venomous.}
            \glspl{gṛhagoḍikā},\footnote{\label{grhagodika}The  scribe of
                \MS{Kathmandu NAK 5-333} noted in the  margin that some of his
                sources read \dev{galagoḍikā}, which is the name of  a snake known
                also in  the \CS\ and elsewhere in literature (cf.\ note 
                \ref{galagodika}, p.\,\pageref{galadodika}).} four-footed insects and 
                others \\[2ex] 
in 
their urine and faeces &  \se{kiṭipa}{lice}, \se{picciṭā}{`flat insects'}, 
\se{kaṣāyavāsika}{`orange-dwellers'}, \se{sarṣapaka}{`pepper snakes'}, 
\se{toṭaka}{`angry beetles'}, \se{varcaḥkīṭa}{dung beetles}, and  
\se{kauṇḍinya}{`pot insects'}\\[2ex] 
 in their semen & mice \label{sukravisa}\\[2ex] 
 in their \se{śūla}{stings}  &	scorpions,  
\se{viśvambhara}{`earth 
scorpions'},  \se{varaki}{wasps},\footnote{\dev{varaṭī} is  a wasp;  
\dev{varaki} in the Nepalese MSS may be  an alternant  of this word. 
Ḍalhaṇa 
on \Su{5.3.5}{568}  remarked  that some interpreted 
\dev{varakimatsya}  as 
two items, “wasp and fish,”  others as a single one, “wasp-fish.”}  fish, 
\se{ucciṭiṅga}{crabs}, and \se{patravṛścika}{`wing-scorpions'} 
\\[2ex] 
 in 
their saliva, nails, urine,  feces, blood, semen  and 
fangs &  spiders \\[2ex] 
  in the bites of their mouths  & flies, \se{kaṇabha}{wasps} and leeches \\[2ex] 
 in the bites of their mouths,  in their fangs, 
faces,  \dag,  
farts, anuses and feces &  \se{citraśīrṣa}{`speckle-heads'},  
\se{śārava}{`lids'}, \se{kukṣita}{`bellied'}, 
\se{dārukāri}{`wood-enemies'}, 
\se{medaka}{`liquors'}, and \se{śārikā}{`darts'}. \\[2ex]  
continue & continue \\[2ex] 
continue &  continue \\ 
 \bottomrule 
\caption{Passage 5, 
expressed in tabular  format.} \end{longtable} \par} 
% end centering  
\footnote{\cite{kaur-2018} is unhelpful, in  spite 
of a section on the \SS\ 
(pp.\,61--63).}  
\q{got to here - 2023-01 continue with table for  \#5  }

\subsection{Pollution of the environment}

\item
 [6] \diff{The enemies of the king pollute the  waters, roads and foodstuffs  in 
 enemy territory. The experienced  physician,  who has learned how to purify 
 things,  should clean up those polluted things.} 

\item
[7]  Polluted water is slimy and smells of  tears.\footnote{\dev{asra} normally 
means “tears,” but rarely means “blood.” }  It  is covered with froth and covered 
with streaks. The frogs and fish die,  the birds are crazed and, along with the 
wetland creatures, they wander about  aimlessly.  

\item [8]  Men, horses and elephants who swim in it  experience
vomiting, delusion, fever, swelling and sharp pains.\footnote{On  the
    polysemy of \se{nāga}{elephant\slash  snake}, see \cite{seme-1979}.} 
    He should try to purify that polluted water,  after curing their
    ailments.

\item [9]  
 
And so, he should burn \gls{dhava} and \gls{aśvakarṇa}, %
% Dipterocarpa turbanitus  Gaertn., GVDB 28
as well as \gls{pāribhadra}, %Erythrina suberosa,  GVDB 245 with
% \gls{pāṭalā}
% GVDB 242 Stereospermum  suaveolens DC.
and  \gls{sidhraka} and  \gls{muṣkaka}, %%%%%%%%%%%GVDB 312,
%%%%%%%%%Schrebera  swietenioides Roxb and
%Elaeodendron glaucum Pers.
and with \gls{rājadruma} % āragvadha, GVDB 37  Cassia fistula
%%%%%%%%%%%%%Linn
and \gls{somavalka}.\footnote{\label{water-detox}Cf.\ with the
    recipe at \SS\ \Su{5.6.3}{580} for a paste to put on drums etc., 
    p.\,\pageref{drum-detox} below.}
    Then he should sprinkle that ash, cold, on  the waters.

\item
 [10--11]  And in the same way, putting a handful of  the ash in a pot, one may 
 also purify water  that  one wants.  If any one of the limbs of cows, horses,  
 elephants, men or women, touch a place on the ground that enemies have 
 spoiled with  poison, or a ford or rock or a flat surface, then it swells up and 
 burns and its  hair and nails fall out on that place.\footnote{``Swells up" 
 translates an  unclear reading that was probably \dev{śūyati}, which may be 
 an irregular  form of $\surd$\dev{śū, śvā, śvi}  
 \citep[see][175--176]{whit-root}.}  

\item
 [12]  In that situation, he should grind up  \gls{anantā} together with all the  
 aromatic items, with alcoholic drinks.  And then he should sprinkle the paths 
 that need to be used with waters mixed with  mud.\footnote{Our “alcoholic 
 drinks”  translates \dev{surā}.  For a discussion of this  term at our period see 
 \cite[37--39 \emph{et passim}]{mchu-2021}.}  \diff{And if there exists 
 another path,  he should  go by that.}\footnote{Ḍalhaṇa on  \Su{5.3.12}{568} 
 cited a similar  reading for the fourth pāda, but with a negative particle, “and if 
 there is no other way, one  should go by that.”}  

\item [13]  

When grasses and foods are  polluted, people collapse, fall
unconscious. And others vomit. They get  \sse{viḍbheda}{loose stool}loose 
stool or they  die.\footnote{In “they get loose stool,” the verb
    \dev{ārcchanti} ($\surd$\dev{ṛ}), transmitted in both Nepalese
    manuscripts, has an irregular initial strong vowel. Alternatively, and
    perhaps more likely, it is a combination of \dev{ā+$\surd$ṛ},
    conjugated unusually as a class 6 verb, but with an appropriate sense
    of “to fall into (misfortune).”}  One should apply to them the therapy
    as described.

\item [14--15]  

Alternatively, one should smear  various musical instruments with
antidotes that remove poison and then play  them. What is called the
most  excellent paste for a musical instrument is
\gls{tārāvitāra}\footnote{“Certain  minerals” translates
    \dev{tārāvitāra}, the unanimous  reading of the Nepalese  witnesses.
    But the meaning of this expression is not  clear and may even refer to
    plants, like the other ingredients.  The vulgate  reads \dev{tāraḥ
    sutāraḥ},  which is also not very clear.  However, Ḍalhaṇa on
    \Su{5.3.14}{568} identified  these as “silver” and “mercury.” This is
    highly unlikely  to be a correct understanding of  the passage.
    Historically, mercury is not  naturally present in the South  Asian
    peninsula \pvolcite{5}[233]{watt-1896}  and the word \dev{pārada} that
    Ḍalhaṇa used is probably a loan-word from  Persian \citep[sub
    \emph{paranda, parranda}][244b]{stei-pers}. Mercurial compounds are
    not  reliably attested in South Asia until two or three  centuries
    after the composition  of the \SS\ at the earliest.  The currently
    available  “śāstric” recension of the  \emph{Arthaśāstra} that is
    datable to 175--300 \CE\  \citep[29--31]{oliv-2013} does  not mention
    mercury (\emph{ibid}, 534). See further the study by  Dagmar \citet[17,
    \emph{et passim}]{wuja-2013b}.}  together with  \gls{surendragopa},
    and a  portion of of \gls{kuruvinda} equal to that,  together with the
    bile called  “brown cow”.\footnote{\dev{surendragopa}  and
        \dev{kuruvinda} are both  uncertain, see index. Ḍalhaṇa's opinion has
        been followed here, but it  seems fair to say that all commentators
        were  guessing.} By the sound of  the musical instrument, even
        terrible poisons that may be present  at that place are destroyed.

\item [16]  

If there is smoke or wind that  is affected by poison then  birds are
dazed and fall to the ground.  People get  coughs, colds, and head 
illnesses, and acute eye diseases.\footnote{The syntax  of this verse
    is somewhat  loose; the vulgate has regularized it, smoothing  out
    the difficulties.}

\item
[17]  

The smoke and air can be  purified by putting into the  air:  \gls{lākṣā}, 
\gls{haridrā}, \gls{ativiṣā}, and \gls{abhayā}, with \gls{vakra}, \gls{kuṣṭha}, 
\gls{elā},\footnote{}\q{write  footnote:  don't repeat  ativiṣā; vulgate similar to  
H.} as well as \gls{hareṇu}, and \gls{priyaṅgu}.  
\end{translation}  


\subsection{The origin of  poison}  

\begin{translation}[resume]  

\item
 [18]  As it is told, the arrogant  demon called Kaiṭabha  created an obstacle for 
 lotus-born Brahmā, at  the very time that he  was creating these  
 creatures.\footnote{At  this point, the text  seems to make a new  beginning to 
 the topic  of toxicology, as if  starting a new  chapter.  It is notable  that no 
 reference is  made here to  the famous origin  story of poison in the  churning 
 of the primal  milk ocean; for  discussion of the  sources of this  account, see  
 \cite{bede-1967}. For reflections on this  passage, connecting it  with Rudra 
 and the  \emph{Śatapathabrāhmaṇa},  see  \cite{mana-2019}.}  

\item
[19] Pitiless Fury took a  body and burst out of  the mouth of furious  Brahmā's 
store of fiery  energy.\footnote{“Fury”  is here  anthropomorphised.}  

\item
[20]  He burned that  great, thundering,  apocalyptic  demon. Then,  after 
bringing about the annihilation of  that demon, his  amazing fiery  energy 
increased.  

\item
 [21]  And so, there was a  sinking down  (\emph{viṣāda}) of  the Daityas.  
 Observing that, it was named  “poison  (\emph{viṣa})”  because of it's  ability 
 to produce a  “sinking down.”  

\item
 [22] After  that, the Lord  created beings and  subsequently made  that fury 
 enter into creatures  still and moving.  

\subsection{The working of poison}

\item
 [23--24]  Water that falls  from the sky to the  earth has no  obvious flavour. 
 The savour of the  different places it  lands on enters into  it.  In the same way, 
 whatever substance  a poison reaches, it  establishes itself  there and by its  
 nature it  takes on that  substance's  savour.\footnote{The  scribal emendation  
 in  \MScite{Kathmandu  NAK  5-333} of  \dev{niyacchati}  to  \dev{nigacchati}  
 suggests that the  scribe had more  than  one manuscript  before him, one  of 
 them  representing the  reading of the  vulgate  recension.}  

\item
 [25]  Generally  speaking, in a  poison, all the  qualities are  really sharp.  For 
 this reason,  every poison is  known to irritate  all of the  humours.  

\item
[26]  Irritated and afflicted by  the poison, they leave  their natural functions.  
Poison does not get  digested, so it blocks the  breaths.\footnote{Probably  a 
reference to the five  breaths. Ḍalhaṇa  referred to winds  (\dev{vāta}), but this  
does not seem correct  since it is a reference to
  humours rather than breaths.}
  
\item[27] 
  
Breathing is obstructed because its pathway is blocked by phlegm. Even
if life continues, a man remains without consciousness.
  
\item [28]
  
Similar to semen, the poison of all angry snakes pervades the whole
body, and goes to the limbs like semen because of being stirred up.
  
% Cakra on
%Ca.ci.2.4.49;
%in 46 śukra
%is
%all-pervading
%
%where there
%is
%saṃsparśana
  
\item [29]
  
The fang of snakes is like a hook.  When it gets there, it sticks
inside them. That is why the unagitated poison of a snake is not
released.
  
\item [30] 
  
Sprinkling with very cold water is traditional for all cases of
poisoning, because poison is declared to be extremely hot and
sharp.\footnote{The verb \dev{paṭh} “is declared, read aloud” here
    could possibly suggest that the author is working within a written,
    not oral, tradition.}
  
\item [31]
  
Poison in insects is slow and not very hot, having a lot of wind and
phlegm. So in cases of insect poisoning, sweating is not forbidden.
  
\item [32cd]
  
In cases of a strike or a bite, the poison may, of its own accord,
stay there.
  
\item[33--35ab]
  
\dag Having come upon a body,\footnote{“Having come upon” translates
    \dev{prakhyāpya}, which is hard to interpret unless it is a rare form
    connected with the sense “to see.”} in the case of corpses that have
    been pierced by a poisoned arrow and bitten by a snake, someone who
    eats the poisoned flesh of a recent corpse out of carelessness will
    suffer with illness according to the poison, or even die. And
    therefore, the flesh of those should not be eaten when they have just
    died.
  
It is admissable after three quarters of an hour, but without the
poisoned arrow and the snakebite.
  
\item[35.1]

[At this point an Upajāti verse is added in the margin of K but is not
fully legible; the version of the text in H is also incomplete and not
fully comprehensible.]\footnote{\emph{Mādhavanidāna}, 69.20--21
    \citep[480]{madhava1} has verses that are directly parallel to this
    section: \dev{darvīkarāṇāṃ viṣam āśughāti sarvāṇi coṣṇe
    dviguṇībhavanti// ajīrṇapittātapapīḍiteṣu bāleṣu vṛddheṣu bubhukṣiteṣu 
    //20//  kṣīṇakṣate mohini kuṣṭhayukte rūkṣe ’bale garbhavatīṣu cāpi//
    śastrakṣate yasya na raktam eti rājyo latābhiś ca na saṃbhavanti //21//}.
    This passage is the only occurrence in the ayurvedic text corpus that
    relates to the Nepalese version of the \SS\ at this point. This
    suggests that Mādhavakara (fl.\,ca.\,700, Bengal) knew and used the
    Nepalese version.}
  

  \item[35.3]

\dag When, in a wound, the poison that is connected with these
qualities runs, \ldots Therefore, not everything that is damaged by
poison and eaten causes death.\footnote{At this point, witness H
    inserts a marginal Indravajrā verse about diseases that afflict
    immoral women.}
  
\item[35.1] 

  [ślokas in the MSS that aren't in the vulgate. The first line
doesn't scan. Witness K addsa part of the start of this in the
bottom margin. This material is repeated at 3.39.2in MS H. ]
  

  \item
  [35cd \& 36cd] 
  
One designates a person who has diarrhoea of feces looking like
\se{gṛhadhūma}{soot} with wind,\footnote{\dev{gṛhadhūma} is not a plant
    in this context, \emph{pace} \cite[362]{moni-sans}. See the discussion
    in note \ref{soot}, p.\,\pageref{soot} above.\label{grhadhuma}} and who
    vomits foam, as “someone who has drunk poison.” \item[37]  Therefore,
    fire burns a  heart that is  pervaded by poison. For, having pervaded
    of its own accord  the location of consciousness, it
    abides.\footnote{Ḍalhaṇa said that someone who has died from drinking
        poison has a heart that cannot be burned because it is pervaded by
        poison (\Su{5.3.37}{570}). But the sense of the Nepalese MSS is the
        opposite.}
 
 \subsection{Patients beyond help}  
 
 \item[38] 

Patients who should not be accepted include: those who have been
bitten under a \gls{aśvattha}, in a temple, in a cemetery, at an
ant-hill, at dawn or dusk, at a crossroads, under Yama's
asterism,\footnote{\dev{yāmye} means “southerly” but Ḍalhaṇa on
    \Su{5.3.38}{570} interpreted it as “\sse{yāmya}{in Yama's direction}in
    Yama's direction” as “under the seventh asterism.”} under the Great
    Bear and people who have been bitten in lethal spots.
 
 \item[39]  
 
The poison of cobras kills rapidly.  They all gain twice the intensity
in those who have indigestion, those who are afflicted by bile or
wind, old people, children and the hungry.
 
 
 \item[39.1]  
 
In those whose who are mad or intoxicated, or who suffer from anxiety,
or who are unable to tolerate its various strengths, it becomes sharp.
\dag \ldots
 
 \item[39.2] 
 
 
 
 \footnote{Material corresponds to SS.1.45.205ab, where it  
 describes how alcohol produces intoxication because it is fine, hot  
 and sharp and travels through the vessels disturbing the senses and  
 the mind and intoxicating the potency.}  
 
 \item [3.40cd--3.43ab]  
 
One should reject someone overcome by poison who \diff{does not bleed} 
when cut with a knife, where weals do not appear as a result of  
 lashes,\footnote{Ḍalhaṇa, on \Su{5.3.40}{570}, glossed \dev{latābhis} 
“by means of whips,” as “when the body is struck by whips.”} or where 
there is no horripilation because of cold water, whose mouth is 
\diff{crooked}, whose hair is falling out of his head.  A man who is 
fatigued and those who stammer,\footnote{nāsāvasāda \& plural 
sakaṇṭhabhaṅgāḥ}  
 % 3.42: 
one who has a black and red swelling at the site of the bite, with
lockjaw, should be avoided.  The same goes for someone who has a solid
plug emerge from their mouth  and someone who has blood running from
above and below.
% 
% \item[3.43ab]  
% 
The physician should also avoid a person who has fangs that have not
fallen out quickly.\footnote{The grammatical verb-form
    \dev{parivarjayīta} “he  should avoid,” opt., 3rd, sg., is unusual.
    \citet[10\,ff]{reno-1940} documented such forms from the
    \emph{Aitareyabrāhmaṇa} onwards.  \citet[\P  6.3.3 “Peculiar optative
    endings”, pp.\,176--177]{ober-2003} showed that the form is
    well-documented in \emph{manuscripts} of the \emph{Mahābhārata}, but
    has been edited out of the printed critical edition in almost all
    cases. Cf.\ also \cite{kuli-2006}.  
    
    The concern about a patient who “has fangs that have not fallen
out” is hard  to understand.  The word \dev{daṃṣṭrā} does not mean
human teeth  (\dev{danta}).  We therefore prefer to understand
this as describing a patient where the  fangs of a venomous
creature remain in the bite-wound.  This requires  construing the
expression as a \emph{bahuvrīhi} compound: \dev{daṃṣṭrā} or
\dev{daṃṣṭra} $+$ \dev{anipātaḥ}.}
\end{translation}
