% !TeX root = incremental_SS_Translation.tex
% Paras

\chapter{Cikitsāsthāna 5:  On the Treatment of Great Wind Diseases}

\section{Literature} 

Meulenbeld offered an annotated overview of this chapter and a bibliography
of earlier scholarship to 2002.\fvolcite{IA}[266]{meul-hist} 

\section{Translation}

\begin{translation}
    
    \item [1]
    Now we shall describe the treatment of great wind diseases.
    
    

    \item [3]
    
Some people say that there are two kinds of
\se{vātarakta}{wind-afflicted blood}, the
\se{uttāna}{superficial} and \se{avagāḍha}{deep}.\footnote{On
    \se{vātarakta}{wind-afflicted blood}, see the discussion at 
    \pageref{ni1:vatarakta}.
        
Ḍalhaṇa commented \citep[424]{vulgate} that \dev{uttāna} refers to
being situated in the skin and flesh, and \dev{avagāḍha} refers to
being situated internally.

The view referred to can be found at \CS\ \Ca{1.19.3}{109--110} and
\Ca{6.29.19}{628}.  See the contextual discussion by
\citet[101]{emme-1984}.} However, this is not correct. 

Why?

It is like \se{kuṣṭha}{pallid skin disease}: having been superficial, it becomes 
deep after an interval of time.  Therefore counting it as two things is rejected.

\item[4]

In this regard, it brings about pains that are caused by 
\se{vātaśoṇita}{wind-afflicted blood}. 

The wind of someone who habitually eats too much, and heavy, hot food,
gets irritated by specific things like conflicts with  strong
people.\footnote{\Dalhana{4.5.4}{424} noted that “conflicts with
    powerful people, etc." were listed in the \emph{Vraṇapraśna} chapter
    (\SS\ 1.21); see \SS\ \Su{1.21.19}{103}. A long list of wind-irritants
    is given in that passage.  \Dalhana{1.15.32}{73}, \emph{et passim},
    glossed \dev{adhyaśana} not as “overeating” but as “eating when
    previous food is not yet digested."} The wind's path is obstructed by
    corrupted blood that suddently unites with it and causes pains that
    are triggered by blood with wind in it.  Experts refer to that as
    “\se{vātaśonita}{wind-afflicted blood}.”  It initially establishes
    itself in the hands and feet.  Then it spreads throughout the body.
    Its preliminary symptoms are pricking, burning, itching, swelling,
    stiffness, rough skin, throbbing of the ducts, sinews and pipes, and
    weakness of the thigh. There is also the chance appearence of brown
    and red \se{maṇḍala}{round patches} that arise on the soles of the feet,
    fingers, ankles and wrists. This disease fully manifests in people who
    do not have treatment and who do not behave as they
    should.\footnote{This specific type of “treatment” (\emph{pratisāra})
        is described as being of four types at \SS\ \Su{4.40.69}{558}, i.e.,
        paste, linctus, honey and powder \pvolcite{1}[527]{josi-maha}.}

Its characteristics have been stated.   In the case of any of these, 
those who do not take precautions develop disabilty.



\item[5]

Wind-blood most commonly gets irritated in those who are very
delicate, those who have wrong diet and activity, those who are corpulent,
and those who live in comfort.\footnote{This same passage
    occurs in \Su{2.1.40}{263}, where Ḍalhaṇa questioned its
    appropriateness given its repeated occurrence here.  See footnotes
    \ref{ni1:vatarakta} and \ref{2.1.40} (pages \pageref{ni1:vatarakta}
    and \pageref{2.1.40}).}

    \item[6]
    
In such a case, one should treat a patient if they are strong, self-possesed, and 
well-resourced, as long as they are not also afflicted by \se{saṅkoca}{cramp}


 the patient who is not degenerating due to 
wasting of life air, thirst, fever, unconsciousness, dyspnea, trembling, and loss 
of appetite, is not oppressed by the contraction [of limbs], is strong, composed, 
and has the means.



  \newpage
\begin{tt}
    \raggedright
    \bigskip
    \marginpar{Draft tr.\ from here}





    \item[7]
    
    In the treatment, at the beginning itself one should do blood-letting of the 
    wind-affected body part little by little and more than once. That (slow 
    blood-letting) is because of the danger of further aggravation of wind. One 
    should avoid doing blood-letting of the part hardened or weakened by 
    excessive wind.\footnote{In H, the reading \dev{amlāna} does not make sense 
    given the context. Therefore, we have accepted the vulgate reading 
    \dev{mlāna} for the translation.} Thereafter, one should make the patient do 
    the remedies of vomiting, etc. If the wind that is mixed [with blood] or 
    separated is very aggravated then one should make him consume aged ghee 
    or goat-milk. Or, [one can give him] half a measure of oil added with an 
    \textit{akṣa} of \gls{madhuka} and cooked with 
    \gls{pṛśniparṇī}\footnote{Ḍalhaṇa glossed \citep[425]{vulgate} 
    \emph{śṛgālavinnā} as \emph{pṛśniparṇī}.}, or the oil that is sweetened by 
    sugar and honey and cooked with \gls{śuṇṭhī} and \gls{kaśeru}. Or, one 
    should boil milk with an eight times volume of the decoction of the following 
    herbs: \gls{śyāmā}, \gls{rāsnā}, \gls{suṣavī}, 
    \gls{pṛśniparṇī}\footnote{According to Ḍalhaṇa, \emph{śṛgālavinnā} is 
    \emph{pṛśniparṇī}.}, \gls{pīlu}, \gls{śatāvarī}, \gls{śvadaṃṣṭrā}, and 
    \gls{dvipañcamūla}. This milk should then be used to cook oil with the 
    admixture of pastes of \gls{meṣaśṛṅgī}, \gls{śvadaṃṣṭrā}, \gls{madhu}, 
    \gls{nāgabalā}, \gls{bhadradāru}, \gls{vacā}, and \gls{surabhi}. This 
    (resultant) should be utilised in drinks, etc. Or, one should use the oil that is 
    cooked with a decoction of \gls{śatāvarī}, \gls{apāmārga}\footnote{Ḍalhaṇa 
    glossed \citep[425]{vulgate} \textit{mayūraka} as \textit{apāmārga}.}, 
    \gls{dhavaka}, \gls{madhuka}, \gls{kṣīravidārī}, \gls{balā}, \gls{atibalā}, and 
    \gls{tṛṇapañcamūlī}\footnote{Ḍalhaṇa commented \citep[425]{vulgate} that 
    \gls{kuśa}, \gls{kāśa}, \gls{nala}, \gls{darbha}, \gls{kāṇḍa}, and \gls{ikṣuka} 
    are called \textit{tṛna} (grass).}, with the admixture of \gls{kākolī}, etc. Or, 
    one should use the \gls{balā}-oil that is cooked as 
    \emph{śatapāka}.\footnote{\emph{Śatapāka} seems to be an oil that is 
    prepared with a hundred parts of some things similar to \textit{sahasrapāka} 
    that is prepared with one thousand parts of some herbs. Refer 
    \textit{Cikitsāsthāna} Ch. 4 text 29 for the preparation of 
    \textit{sahasrapāka}.} Or, [the affected body part] should be moistened with 
    milk that is boiled with the roots of wind-alleviating herbs, or it should be 
    moistened with sour things.\footnote{Ḍalhaṇa commented 
    \citep[425]{vulgate} that the sour things (\textit{amla)} are \gls{surā}, 
    \gls{sauvīraka}, \gls{tuṣa}-water, etc. \textit{Surā} is some kind of liquor, 
    \textit{sauvīraka} is perhaps the fruit of the jujube tree, and \textit{tuṣa} is 
    perhaps Terminalia Bellerica (\dev{vibhītaka}).} In that regard, five remedies 
    prepared with milk are described. For preparing a poultice, milk should be 
    cooked in ghee, oil, fat, marrow, and \emph{dugdha}\footnote{In the \SS, the 
    word for milk is \textit{kṣīra} or \textit{payas} but not \textit{dugdha}. 
    Therefore, the word \textit{dugdha} here can mean the sap of plants or 
    something that is extracted.} separately with each of these powdered grains or 
    pulses---barley, wheat, sesame, \gls{mudga}, or \gls{māṣa}---that is mixed 
    with unctuous pastes of \gls{kākolī}, \gls{kṣīrakākolī}, \gls{jīvaka}, 
    \gls{ṛṣabhaka}, \gls{balā}, \gls{atibalā}, 
    \gls{pṛśniparṇī}\footnote{\emph{śṛgālavinnā}}, \gls{meṣaśṛṅgī}, \gls{piyāla}, 
    sugar, \gls{kaśeru}\footnote{For \emph{kaśerukā}}, \gls{surabhi}, and 
    \gls{vacā}. Or, the essence of unctuous fruits\footnote{Ḍalhaṇa commented 
    \citep[425]{vulgate} that the unctuous fruits mentioned here are sesame, 
    castor, \gls{atasī}, \gls{vibhītaka}, etc.} can be used as a poultice. Or, a 
    \textit{veśavāra}\footnote{In H, the reading \dev{vaiśavāro} does not make 
    sense. It should have been \dev{veśavāro}, as shown in the vulgate, which is 
    the reading we have accepted here.\\ \textit{Veśavāra} is boneless meat 
    minced, steamed, and added with spices, ghee, etc. Refer to 'Ayurveda 
    Medical Dictionary' by Ranganayakulu Potturu.\\Perhaps the word 
    \dev{vaiśavāra} is an earlier form of the word \dev{veśavāra}.} prepared from 
    the flesh of a fat \textit{cilicima} fish\footnote{H has the compound word 
    \dev{nalapīnamatsya}. \dev{nalamīna} is a particular fish known as 
    \textit{cilicima} (\dev{cilicimaḥ}). See \textit{Amarakośa}. Also, if the name is 
    \dev{nalamatsya} then the word \dev{pīna} (fat) within the name is not 
    according to proper Sanskrit. But, it can be allowed because the word 
    \dev{matsya} (fish), instead of being a part of the name, can be considered to 
    mean fish in general and thus the word \dev{pīna} becomes its modifier. Thus, 
    \dev{nalapīnamatsya} can mean "a fat fish that is a \dev{nala} 
    (\textit{cilicima})".\\ Ḍalhaṇa says in his comment \citep[425]{vulgate} that 
    \dev{nalamīna} is a type of \dev{rohita} (\textit{rohita}). Monier Williams says 
    that \textit{rohita} is a kind of fish: Cyprinus Rohitaka. Regarding the 
    \textit{rohita} fish, there is a \textit{subhāṣita}: \dev{agādhajalasañcārī na 
    garvaṃ yāti rohitaḥ | aṅguṣṭhodakamātreṇa śapharī pharpharāyate ||} This 
    indicates that \textit{rohita} is a deep water fish.}\q{The webpage 
    https://hindi.shabd.in/vairagya-shatakam-bhag-acharya-arjun-tiwari/post/117629
     says that this verse belongs to the \textit{Nītiratna}. I could not find this 
    text.} can be used instead. Or, [one can use] the poultice containing 
    \gls{bilva}-rind\footnote{The word \dev{pesikā} in H should be read 
    \dev{peśikā}.}, \gls{tagara}, \gls{devadāru}, \gls{saralā}, \gls{rāsnā}, 
    \gls{hareṇu}, \gls{kuṣṭha}, \gls{śatapuṣpa}, liquor, yogurt, and whey. Or, [one 
    can use] the ointment prepared by mixing \gls{mātuluṅga}, 
    \emph{amla}\footnote{Perhaps it could mean vinegar or sour curds. Refer to 
    Monier Williams Sanskrit Dictionary.}, salt, and ghee with honey and 
    \gls{śigru}-root. Or else, [one can use] the unctuous sesame paste. 

    \item[8]
    When the [condition of wind-blood] has a predominance of bile, the patient 
    should be made to drink a decoction of grapes, \gls{revataka}-fruit, 
    \gls{payasyā}, \gls{madhuka}, \gls{candana}, and \gls{kāśmarī}. This 
    decoction is sweetened with honey and sugar before consumption. Or, the 
    decoction of \gls{śatāvarī}, \gls{paṭola}, \gls{patra}, \textit{triphalā}, 
    \gls{kaḍurohiṇī}, and \gls{guḍūcī} should be given. [The patient should be 
    administered] ghee that is prepared with sweet, bitter, and astringent 
    [remedies].\footnote{Ḍalhaṇa commented \citep[425]{vulgate} that the sweet 
    remedies are \gls{kākolī}, etc., bitter remedies are \gls{paṭola}, etc., and 
    astringent remedies are \textit{triphalā}, etc.} 
    
    [The patient] should be sprinkled with a decoction of \gls{bisa}, \gls{mṛṇāla}, \gls{bhadraśriya}, and \gls{padmaka} mixed with goat-milk\footnote{The compound word ending with \dev{kaṣāyeṇa} is taken to be a \textit{bahuvrīhi} for \dev{ajākṣīreṇa} (goat-milk).}, or with rice water that is mixed with milk, sugarcane juice, honey, and sugar, or with whey and sour rice gruel mixed with a decoction of grapes and sugarcane. Or else, [the patient] should be sprinkled with ghee that is prepared with \textit{jīvanīya}\footnote{\textit{Jīvanīya} seems to be a group of medicinal herbs. There is an Ayurvedic preparation called \textit{jīvanīya-ghṛta}. Refer to the \textit{Āyurvedīya Śabdakośa} vol. 1.} or sprinkled with ghee that is purified for one hundred times.

    The poultice [to be applied] should be made of rice flour or of the paste of sour rice gruel mixed with \gls{nala}, \gls{vañjula}, \gls{tālīśa}\footnote{\dev{tālīsa} should be read \dev{tālīśa}}, \gls{śṛṅgāṭaka}, \gls{kālodyā}, \gls{gaurī}, \gls{śaivāla}, \gls{padma}, etc. The poultice should be mixed with ghee.

    \item[9]
     The [condition of wind-blood] with a predominance of blood should be treated in the same way. Also, blood-letting should be done repeatedly.

     \item[10]
    However, when the [condition of wind-blood] has a predominance of phlegm, the patient should be made to consume a decoction of \gls{āmalaka} and \gls{haridrā} that is sweetened with honey, or a decoction of \textit{triphalā}, or a paste of \gls{madhuka}, \gls{śṛṅgavera}, \gls{harītakī}, and \gls{tiktarohiṇī}. He should be made to drink \gls{harītakī} with water mixed with a little urine. He should be sprinkled with oil, urine, salty water, and liquor that are acidic\footnote{Reading the word \dev{sukta} in H as \dev{śukta}}. Or, he should be sprinkled with a decoction of \gls{āragvadha}, etc. 

    The patient should be massaged with ghee cooked with sour cream, urine, liquor, \gls{śuktā}\footnote{Monier Williams states Rumex Vesicarius for \textit{śuktā}}, \gls{madhuka}, \gls{śārivā}\footnote{DCS has this entry: Cryptolepsis buchananii Roem. et Schult. (Surapāla (1988), 453) Decalepis hamiltonii Wight et Arn. (Surapāla (1988), 453)}, and \gls{padmaka}.
    
     The poultice should be made of either the paste of white mustard, or the paste of sesame and \gls{aśvagandhā}, or the paste of \gls{dāru}\footnote{According to V. S. Apte, \dev{dāru} can mean \dev{devadāru}.}, \gls{śelu}, and \gls{kapittha}, or the paste of honey, \gls{śigru}, and \gls{punarnavā},\footnote{H has a short \dev{a} at the end instead of the long \dev{ā}.} or the paste of dry ginger, long pepper, black pepper,\footnote{\dev{vyoṣatiktā} refers to the group of these three pungent spices. Also see \textit{Sūtrasthāna} 14.35.} \gls{pṛthakparṇī}, and \gls{bṛhatī}.\footnote{In H, the Sanskrit syntax does not match up with what the author is trying to say. The name of the fifth paste should also have been in the nominative case, as the other four pastes.}\q{The provisional edition should be modified accordingly.} These five poultices are prepared with salty water. Thus, they have been described.

     \item[11] 
     In case of combined aggravation of two humours or simultaneous aggravation of all three humours, the stated methods of treating those aggravations should be combined.\footnote{It means that the respective methods of treating the aggravation of individual humours should be combined.}  

     \item[12] 
    In all [aggravations], one should consume \gls{harītakī} with jaggery. Or, one should have a diet of rice cooked in milk for ten days and should drink a mixture of \gls{pippalī}s crushed in milk, with increasing by five \gls{pippalī}s each night. Then one should reduce them again by the order of five more [each night].\footnote{In H, the letter \dev{ñ} in \dev{bhūyañca} should have been \dev{ś}.} In this way, one should [reduce] all the \gls{pippalī}s. This is called \textit{Pippalīvarddhamānakam} (Increasing Long Peppers). It indeed cures wind-blood, intense fever,\footnote{Perhaps \dev{viṣamajvara} could mean irregular fever.} loss of appetite, jaundice, abdominal affection, piles, heavy breathing, cough, wasting disease, weak digestion, and heart disease. 

    The poultice is a paste of \gls{sahā}, \gls{candana}, \gls{mūrvā}, \gls{piyāla}, \gls{śatāvarī}, \gls{kaśeru},\footnote{H has \dev{kaśerukā}.} \gls{sahadevā}, \gls{padmaka}, \gls{madhuka}, \gls{śatapuṣpā}, \gls{kuṣṭhā}, \gls{niśaileya}, \gls{kāṭarūṣaka}, \gls{balā}, \gls{atibalā}, and \gls{jīvantī} mixed with milk. Or it is a paste of \gls{kāśmarī}, \gls{madhuka}, and \gls{tarpaṇa} mixed with ghee and cream. Or it is olibanum cooked with milk that is mixed with \gls{madhūcchiṣṭa}, \gls{śārivā}, \gls{sarjarasa}, \gls{madhuka} and the group of sweet herbs. 

    Old ghee that is cooked with \gls{āmalaka} and \gls{sarala} and sweetened with sugar and honey is for drinking. Old ghee that is cooked with \textit{jīvanīya} or that is cooked with a decoction of \gls{suṣavī} is for sprinkling. Cooked \gls{balā} oil is for sprinkling, bathing, enema, and eating\footnote{Perhaps it means that one should eat foods cooked in that oil.}. One should eat food preparations made of rice, \gls{ṣaṣṭika}, barley and wheat accompanied with milk, meat soup, or \gls{mudga} soup that is not sour. Blood-letting also [should be done]. The treatments of vomiting, purging of bowels, enema, and oily enema should be conducted when the humours are highly aggravated.

    \item[13] 

    \item[14] There are verses in this regard.\footnote{The word \dev{bhavati} in H should have been \dev{bhavanti}.}
    \begin{sloka}
    There is immediate relief by the application of remedies such as these by which the physicians cure the chronic condition of wind-blood. 
    \end{sloka}

    \item[15-16]
    \begin{sloka}
    Poultice, sprinkling [oil], plaster, oil massage,\footnote{In H, the part \dev{vyajanānilāḥ} does not make proper sense in the verse. Emending it to \dev{vyajanāni ca} could be a consideration, but fanning (\dev{vyajana}) a patient with wind-blood is not good, as understood from the recommendation that such a patient should stay in a non-windy room. Therefore, we have accepted the vulgate reading for the first half of this verse.} spacious and comfortable rooms\footnote{In H, read the \dev{sa} \dev{saraṇāni} as \dev{śa}.} with no wind, soft pillows, comfortable beds, and soft massages are recommended in the condition of wind-blood.
    \end{sloka}

    \item[17]
    \begin{sloka}
     Exercise, mating, anger, eating hot, sour, or salty foods, sleeping during the day, and food that is slimy or heavy should be avoided.
\end{sloka}

    \item[18]
    One should treat the person who is affected with spasmodic 
    contraction,\footnote{In H, the reading \dev{apatākinam} should have been 
    \dev{apatānakinam}.} who does not have droopy eyes and crooked eyebrows, 
    whose fingers have not become rigid, who is not perspiring or trembling, who is 
    not in a state of delirium, who is not bed-ridden,\footnote{V. S. Apte has 
    \dev{khaṭvayati}. The \textit{Āyurvedīya \'{S}abdakośa} has the entry 
    \dev{khaṭvāpātin} which means \enquote{one who is inclined to fall from 
    bed.} Perhaps the reading in H has an error of the letter \dev{yā} which 
    should have been \dev{pā}.} and who is not restrained externally. There at 
    the beginning itself,\footnote{In H, \dev{prāgaiva} should have been 
    \dev{prāgeva}.} after rubbing the patient with oil and making him perspire, 
    one should treat him with a strong \textit{avapīḍa}\footnote{The 
    \textit{Āyurvedīya \'{S}abdakośa} has the entry \dev{avapīḍa} that means 
    administering an oily paste through the nose. Refer \textit{SS Cikitsāsthāna} 
    Ch. 40 text 44 for a better understanding of \textit{avapīḍa}.}\q{There, 
    Ḍalhaṇa commented that deliberation on \textit{avapīḍa} had been done 
    earlier when it was mentioned. Find that description to know more details.} in 
    order to clear his head. Then, the patient should be made to drink filtered ghee 
    that is properly cooked with a decoction of \gls{vidārigandhā} and other herbs, 
    sugarcane juice, milk, and yogurt. In that way, the wind does not spread 
    exceedingly. 

    Thereafter, one should gather wind-alleviating herbs such as \gls{bhadradāru}, etc. and other constituent parts, along with \gls{yava}, \gls{kola}, and \gls{kulattha}, and the flesh of a freshwater aquatic creature all at one place and prepare a decoction of them. One should take this decoction and mix it properly with sour substances and milk, and then cook the \textit{pratīvāpa}\footnote{It refers to an admixture of substances to medicines either during or after decoction. Refer to Monier-Williams's Sanskrit dictionary.} of \gls{madhuka} in this mixture along with ghee, oil, body fat, and bone marrow. This is \textit{trivṛt} that should be recommended in treatments of sprinkling, oil massage, applying a poultice, oral consumption, oily enema, and errhine for patients having spasmodic contractions.

    The patient should then be made to sweat by the methods described earlier. If the wind is stronger then the patient should be immersed in [a vessel] filled with lukewarm fluid used for sprinkling (\textit{trivṛt}). Or he should be kept in the hot fireplace of a blacksmith.\footnote{H has the reading \dev{rathākāracullyām} that means \enquote{fireplace shaped like a chariot}, but the vulgate reading \dev{rathakāracullyām} makes more sense here. Thus, we have accepted it.} Or else he should be made to sweat by [a mixture of] \gls{kṛśara}, \textit{veśavāra},\footnote{Refer the above text no.7 for \textit{veśavāra}. In H, the syllable \dev{vai} should have been \dev{ve}.} and milk.
 
    Oil cooked with the juice of \gls{mūlaka}, \gls{uruvūka}, \gls{sphūrjaka}, \gls{saptalā}, and \gls{śaṅkhinī} should be used in sprinking, etc. for patients with spasmodic contractions.\footnote{The word \dev{tailam} is not present in H but is present in the vulgate. We have accepted it.} Sour yogurt mixed with \gls{marica} and drunk on an empty stomach alleviates spasmodic contractions. Or else, ghee, oil, body fat, or bone marrow [can be consumed on an empty stomach]. 
    
    This procedure of treatment thus described is for spasmodic contractions caused only by wind. When mixed humours cause it then the treatment should also be mixed. And when the spasms subside the patient should be given \textit{avapīḍa}-s. One should also consider the fats of cock, crab, black fish, and porpoise.\footnote{H has the reading \dev{rasān} which means \enquote{juices}. It seems unrealistic that juice would be extracted by crushing these whole animals. Vulgate has the reading \dev{vasāḥ} instead of \dev{rasān} which appears to be the more probable reading. Thus, we have accepted it.} Milk prepared with wind-alleviating medicines. Gruel prepared with barley, \gls{kola}, \gls{kulattha}, \gls{mūlaka}, yogurt, ghee, and oil. 

    One should treat this recurring spasm for ten nights with oil massage, purging of bowels, enemas, and oily enemas. One should also look up the treatment of diseases caused by wind. One should also undertake preventive measures.   

    \item[19]
    One should treat the paralytic (hemiplegic) patient whose limbs are not languid, who is in pain, and who is self-composed. There, at the beginning itself the patient should be massaged with oil and made to sweat. After cleansing the patient with a mild purifier,\footnote{According to P. V. Sharma, this refers to mild evacuatives (purgatives).} he should be administered with an oily enema and then a non-oily enema. Then at the appropriate time, he should be treated with special enemas of the brain and the head according to the method prescribed in the treatment of \textit{ākṣepaka}.\footnote{Refer \textit{Nidānasthāna} 1.50-51 for \textit{ākṣepaka}.}\q{Search for the section where the treatment of \textit{ākṣepaka} is described.} \textit{Anutaila} should be used for massage.\footnote{For the procedure of preparing \textit{anutaila}, refer \textit{Cikitsāsthāna} 4.28.} \textit{Sālvala} should be used for poultice.\footnote{For the procedure of preparing \textit{sālvala}, refer \textit{Cikitsāsthāna} 4.14-15.} \gls{balā} oil should be used for oily enema.\q{Make the first letter of sentence capital.} In this way, the unremitting patient should take the treatment for three to four months.

    \item[20]

    \item[21] 

    \item[22]
    One should treat the patient with \textit{ardita}\footnote{Refer 
    \textit{Nidānasthāna} 1.71-72 for \textit{ardita}.} who is strong and possesses 
    the means with the method prescribed in treating wind diseases. The unique 
    thing is the treatment with enemas of the brain and the head, errhine, smoke, 
    poultice, and steam bath through tubes. Then, one should take the great five 
    roots (\textit{pañcamūlī}) with grass and prepare its decoction in milk mixed 
    with twice the water. Then, the decoction with the milk remaining\footnote{It 
    means that the water has evaporated.} should be brought down [the stove] 
    and filtered. It should then be mixed with a \textit{prastha}\footnote{Ḍalhana 
    commented \citep[425]{vulgate} that a \textit{prastha} is a measure of 
    weight that is equal to 32 \textit{pala-s}.} of oil and again placed over fire and 
    cooked thoroughly. Then, the oil mixed with milk should be brought down [the 
    stove] and then churned after it cools down. This is called \textit{kṣīrataila} 
    that should be used in drinks, etc. for patients with \textit{ardita}.  

    \item[23]
    In the diseases of \textit{gṛdhrasī}, \textit{viścañcī}, \textit{kroṣṭukaśīrṣa}, \textit{paṅgukalāya}, lameness, \textit{vātakaṇṭaka}, burning sensation in the foot, numbness of the foot, \textit{avabāhuka}, deafness, and \textit{dhamanīvāta}, one should pierce the blood vessel as described earlier and, barring the case of \textit{avabāhuka}, one should look up the treatment for wind diseases.

    \item[24]
    However, in the case of \textit{karṇamūla},\footnote{The vulgate has the reading \dev{karṇaśūle} which appears to be a more credible reading according to the context.} lukewarm juice of \gls{śṛṅgavera}\footnote{\dev{śṛṅgavera} appears to be a name of ginger. Refer to the Sanskrit dictionary of Monier Williams.} mixed with \gls{madhuka}, oil, and salt should be put into the ears.\footnote{In H, the reading \dev{rasaiḥ} does not seem to make sense here. Hence we have accepted the vulgate reading \dev{rasam}.} Or else one can use goat urine, \gls{madhuka}, and oil. Or else one can use oil that is cooked with \gls{mātuluṅga}, \gls{dāḍima}, \gls{tintiḍīka} juice, and urine.\footnote{In H, the word \dev{taila} should have been \dev{tailam} to make proper sense. The vulgate has this reading. Thus we have accepted it.} Or else one can use oil that is cooked with sour liquor, buttermilk, and urine. 
    
    One should also make the patient sweat with a steam bath through tubes. One should also look up the treatment for wind diseases. More will be said later.

    \item[25]
    In the case of \textit{tūnī} and \textit{pratitūnī}, one should make the patient drink ghee and salt with hot water. Or else one should administer the powder of \gls{pippalī} and other herbs with hot water. Or else one should make the patient drink ghee that is made thick with asafoetida and \gls{yavakṣāra}.\footnote{\dev{yavakṣāra} is an alkali prepared from the ashes of burnt green barleycorns. Refer to the Sanskrit dictionary of Monier Williams.} One should also treat the patient with enemas. 

    \item[26]
    In the case of \textit{ādhmāna},\footnote{Refer to \textit{Nidānasthāna} 1.88. V. S. Apte explains it as \enquote{swelling of the belly}. P.V. Sharma has translated it as flatulence.} however, one should do \textit{avatarpaṇa},\footnote{We are unclear about its meaning. The vulgate has the reading \dev{apatarpaṇa} that means fasting.} heating the hands, \textit{phalavartikriyā},\footnote{The entry \dev{phalavarti} has the meaning \enquote{suppository} in the Sanskrit dictionary of Monier Williams. The Cambridge dictionary explains suppository as \enquote{a small, solid pill containing a drug that is put inside the anus, where it dissolves easily.} Refer to the link https://dictionary.cambridge.org/dictionary/english/suppository. Last accessed 30-Oct-2023.} stimulation of digestion, and [administer] digestives. One should also employ the purging of bowels and enemas. 

    In the case of \textit{pratyādhmāna},\footnote{Refer to \textit{Nidānasthāna} 1.89. According to the Sanskrit dictionary of Monier Williams, it is a kind of tympanites or wind-dropsy.} one should employ vomiting, fasting, and stimulation of digestion.

    \item[27]
    In the case of \textit{aṣṭhīlā} and \textit{pratyaṣṭhīlā},\footnote{Refer to \textit{Nidānasthāna} 1.90 and 1.91.} the procedure is that of \textit{gulma} and internal abscess.

    \item[28]
    The beneficial asafoetida, the three pungent spices (long pepper, black pepper, and dry ginger), \gls{vacā}, \gls{ajamoda} grains, \gls{ajagandhā}, \gls{dāḍima}, \gls{tintiḍīka}, \gls{pāṭhā}, \gls{citraka}, \gls{cavyā}, \gls{saindhava}, \gls{viḍa}, \gls{sauvarcala}, \gls{yavakṣāra}, \gls{suvarcikā}, \gls{pippalī} root, \gls{amlavetasa}, \gls{śaṭhī}, \gls{puṣkara}, \gls{hapuṣa} (juniper berry), and \gls{ajājī} (cumin seeds) should be powdered. This powder should be mixed with a lot of \gls{mātuluṅga} juice. Then it should be made into pills each weighing one \textit{akṣa}. Thereafter the patient of wind disease should consume one pill every morning. This medicine indeed cures \textit{gulma}, rapid breathing, cough, loss of appetite, heart disease, \textit{ādhmāna}, \textit{pārśvodara}, \textit{bastiśūla}, \textit{anāhamūtra}, painful piles, \textit{plīhodara}, and \textit{pāṇḍuroga}. Also, this medicine is excessively used in cases of \textit{tūnī} and \textit{pratitūnī}.
 
    \item[29]
    There are verses in this regard. 
    \begin{sloka}
    The wind that has entered into the body tissues should be correctly understood as either pure or vitiated by humours\footnote{In H, the reading \dev{lakṣaṇonyām̐śca} does not make sense. Hence I cannot translate it. Perhaps the correct reading could be \dev{lakṣaṇābhyāñca}. This would connect with the two conditions of the wind as stated in the verse.} and should be cured accordingly. 
    \end{sloka}

    \item[30]
    \begin{sloka}
    The wind that is accompanied by fat causes a swelling that is painful, hard, and cold. The physician should properly treat it like a treating a swelling.
    \end{sloka}

    \item[31]
    \begin{sloka}
    When the wind accompanied by phlegm and fat enters the thighs, it causes pain in and immobility of the thighs due to numbness, pain, and fever. 
    \end{sloka}

    \item[32] 
    \begin{sloka}
    Also, the thighs become pained, stiff, cold, and do not quiver due to sleep. They become heavy and as if belonging to someone else.\footnote{In H, the verb \dev{vartate} should have been in the dual. Also, the word \dev{āsthirau} does not make sense. The vulgate has the sensible reading \dev{asthirau} which we have accepted here.}
    \end{sloka}

    \item[33]
    \begin{sloka}
    That is called \textit{ūrūstambha}. Others call it \textit{āḍhyavāta}. In that case, one should drink the \textit{ṣaṇḍharaṇa} powder with cool water. 
    \end{sloka}

    \item[34]
    \begin{sloka}
    Similarly, consuming the powder of \gls{pippalī} and other herbs with hot water is beneficial. Or else, one should consume the powder of \textit{triphalā} with \gls{kṣaudra} and \gls{kaṭukā}. 
    \end{sloka}

    \item[35-38]
    \begin{sloka}
    Or else, one should drink the best \gls{guggulu} or \gls{śilājatu} with urine. Such a person cures the wind that is afflicted by phlegm and accompanied by fat, as well as heart disease, loss of appetite, \textit{gulma}, and internal abscess. 

    One should employ salty urine [therapy], sudation, and hard rubbing. One should also apply [the paste of ] mustard and \gls{karañja} fruits mixed with urine.\footnote{The word \dev{dihet} in H is not a proper Sanskrit word. We have taken its proper form \dev{dihyāt} as given in the vulgate.}

    One should eat old \gls{śyāmāka}s, \gls{kodrava}, \gls{uddālaka}, etc. along with uncooked\footnote{The vulgate has the reading \dev{aghṛtaiḥ} that means without ghee.} flesh of wild animals and unsalted vegetables that are beneficial.
    \end{sloka}

    \item[39]
    \begin{sloka}
    When the phlegm and fat become amply reduced one should again employ the treatment of oil massage, etc. for the patient.  
    \end{sloka}


\end{tt}
    
\end{translation}
