% !TeX root = surgery.tex

% turn off the footnotes, for the conference handout
%\renewcommand\footnote[1]{\relax }



%INTRODUCTION
%A. Preliminaries
%1. Aim of the Article
%2. Importance of 1.16 in the History of Āyurveda 
%B. Text
%1. The Nepalese Version
%2. Vulgate
%3. Differences between the Two (as exemplified by 1.16)
%C. Edition 
%1. Manuscripts
%2. Editorial Principles
%EDITION OF 1.16 

  \newcommand{\animal}[4]{#1 (\emph{#2}%\footnote{#3 (#4)})
}
\newcommand{\plant}[4]{#1 (\emph{#2}%\footnote{#3 (#4)})
}
\newcommand\skt[2]{#1 (#2)}


 \section{Translation of Sūtrasthāna 16}
%\subsection{Sūtrasthāna, adhyāya 16}
\renewcommand{\se}[2]{#2 (\dev{#1})} % English with Devanagari in parens
\newcommand{\seEngOnly}[2]{#2} % just the English

\begin{translation}    
  
\item [1] 

Now we shall expound the method for piercing the ear.\footnote{The topic of 
    \se{kaṛnavyadha}{piercing the ear} is not discussed in the \emph{Carakasaṃhitā}
    (\cite[IB, 326, n.\,175]{meul-hist}), but it is mentioned in some texts that
    followed the \emph{Suśrutasaṃhitā}, such as the \emph{Kaśāpyasaṃhitā} \citep[IIA,
    30]{meul-hist}. Also, the instrument for piercing the ear is described in the
    \emph{Aṣṭāṅgahṛdayasaṃhitā} \Ah{1.26.26}{321}. In the versions of the text known
    to Ḍalhaṇa \citep[76]{vulgate} and Cakrapāṇidatta \citep[125]{acar-1939}, the
    heading of this chapter is “the method of piercing and joining the ear”
    (\dev{karṇavyadhabandhavidhi}), instead of the Nepalese version's “the method of
    piercing the ear” (\dev{karṇavyadhavidhi}). The topic of joining the ear
    (\dev{karṇabandha}) is discussed in passages 17--20 of the Nepalese version.
    However, it appears that only subsequent redactors reflected its importance by
    including it in chapter headings.

 The Nepalese version also omits the opening remark on Dhanvantari that appears in
subsequent versions of the text. For a discussion of the frame story in the
Nepalese version, see \cite{birc-2021}. Ḍalhaṇa \citep[76]{vulgate} and
Cakrapāṇidatta \citep[125]{acar-1939} state that only the ears of healthy people
should be pierced, and they quote the lost authority Bhoja to affirm this: “When
piercing the ears of children who are free of disease at these times, their ear
flaps and apertures, as well as limbs, increase” (for the Sanskrit, see
\cite[76]{vulgate}).

Some texts use the adjective \dev{karṇa-vedhanī} rather than \dev{°vyadhanī}.}

\item [2] 

One may pierce a child's ears for the purpose of preserving and decorating. During the
bright fortnight, when the child is in the sixth or seventh month, on renowned
days, half days, hours and constellations, the physician, with a calming presence,
sits the boy, who has received a benediction and the recitation of a
blessing,\footnote{The causative form \dev{vyadhayet} is known in Classical
    Sanskrit \citep[166]{whit-root}.

The compound \dev{kṛtamaṅgalasvastivācanaṃ} “who has received a benediction and
the recitation of a blessing” is an emendation based on the similar text at
\Su{3.2.25}{346}.  Cf.\ also \Su{3.10.8, 24}{388, 390} that have slightly
different formulations.} on the lap of a wet-nurse.\footnote{The versions of
    1.16.3 known to Cakrapāṇidatta \citep[126]{acar-1939} and Ḍalhaṇa
    \citep[76]{vulgate} have the additional compound \dev{kumāradharāṅke} (“on the
    lap of one who holds the child”) after \dev{dhātryaṅke}. The gender of
    \dev{kumāradhara} is made clear by  Ḍalhaṇa's gloss “a man who holds the child.”
    Also, both versions add \dev{bālakrīḍanakaiḥ pralobhya} (“having enticed with
    children's toys”) to indicate that the child should be tempted with toys to stay
    on the assistant's lap. According to Ḍalhaṇa on \Su{1.16.3}{76}, the toys include
    replica elephants, horses, bulls and parrots. Ḍalhaṇa further mentions that others
    read \dev{bhakṣyaviśeṣair vā} (“or by special treats”) before
    \dev{bālakrīḍanakaiḥ}, but we see no trace of these small kindnesses in our
    witnesses.} Then, he should pull the ear with his left hand 
    and pierce straight through with his right hand at a naturally-occurring
    cleft.\footnote{The versions of 1.16.3 of Cakrapāṇidatta \citep[126]{acar-1939}
        and Ḍalhaṇa \citep[76]{vulgate} add that this naturally-occurring cleft is
        illuminated by a ray of sunshine  (\dev{ādityakarāvabhāsite}).

The syntax of this slightly long sentence is unusual in beginning with the dual
object \dev{tau} “the two (ears)” at the start of the sentence, which is remote
from the main verb.  The other singular accusatives referring to the ear being
pierced are governed by absolutives.} For a boy, do the right ear first; for a
girl, do the left one. Use a needle on a thin ear; an \seEngOnly{ārā}{awl} on a thick
one.\footnote{Ḍalhaṇa on 1.16.3 \citep[76]{vulgate} clarifies that the awl is a
    shoe-maker's knife for piercing leather.  He also cites the authority of “the
    notes of Lakṣmaṇa” (\emph{Lakṣmaṇaṭippaṇaka}) on the issue of the thickness of the
    needle. \textit{The Notes of Lakṣmaṇa} is not known from any earlier or
    contemporary sources and was presumably a collection of glosses on the \SS\ that
    was available to Ḍalhaṇa in twelfth-century Bengal. See \citet[IA,
    386]{meul-hist}.}
    
    
\item [3]  
    
One may know that it was pierced in the wrong place if there is excess blood or
too much pain. The absence of side-effects is a sign that it has been pierced in the right
place.\footnote{At this point, \MScite{Kathmandu KL 699} is missing a folio, so
    the rest of this chapter is constructed on the basis of witnesses
    \MScite{Kathmandu NAK 5-333} and \MScite{Kathmandu NAK 1-1079}.}
    
\item [4] 
 
In this context, if an ignorant person randomly pierces a \seEngOnly{sirā}{duct}
there will be fever, burning, \seEngOnly{śvayathu}{swelling}, pain, 
\seEngOnly{granthi}{lumps},
\seEngOnly{manyāstambhā}{paralysis of the nape of the neck}, 
\seEngOnly{apatānaka}{convulsions},
headache or sharp pain in the ear.\footnote{This passage is significantly
    augmented in Cakrapāṇidatta's and Ḍalhaṇa's versions, to outline the specific
    problems caused by piercing three ducts called \dev{kālikā}, \dev{marmikā} and
    \dev{lohitikā} (1.16.4 \citep[126]{acar-1939} and 1.16.5 \citep[77]{vulgate}
    respectively). In fact, the order of the problems mentioned in the Nepalese
    version has been retained in the other versions and divided between each duct.
    Cakrapāṇidatta's commentary on 1.16.4 \citep[126]{acar-1939} cites several verses
    attributed to Bhoja on the problems caused by piercing these three ducts in the
    ear flap: `\dev{lohitikā}, \dev{marmikā} and the black ones are the ducts
    situated in the earflaps.  Listen in due order to the problems that arise when
    they are pierced. Paralysis of the nape of the neck and convulsions, or sharp pain
    arise from piercing \dev{lohitikā}. Pain and lumps are thought to arise from
    piercing \dev{marmikā}. Piercing \dev{kālikā} gives rise to swelling, fever and
    burning.'}
    
\item[5]     
    
Having removed the \se{vartti}{wick} because of the accumulation of
humours or an unsatisfactory piercing at that location,\footnote{In addition to these reasons,
    1.16.6 of Ḍalhaṇa at \Su{1.16.6}{77}, added “because of piercing with a painful,
    crooked and unsatisfactory needle” (\dev{kliṣṭajihmāpraśastasūcīvyadhāt}) and 
    “because of a wick that is too thick” (\dev{gāḍhataravartitvāt}). Ḍalhaṇa was
    aware of the reading in the Nepalese version because in his commentary on
    \Su{1.16.6}{77} he noted that some read “because of the accummulation of humours”
    rather than “because of piercing with a painful, crooked and unsatisfactory needle
    or because of a wick that is too thick.” On the concept of humoral
    \se{samudāya}{accumulation}, see the important analysis by \citet{meul-1992}.} he
    should smear it with barley, liquorice, \seEngOnly{mañjiṣṭhā}{Indian madder},
    and the root of the \seEngOnly{gandharvahasta}{castor oil tree}, thickened with 
    honey and
    ghee. And when it has healed well, he should pierce it again.\footnote{The
        description of the drug is ambigious: the word “root” could be taken with each
        plant, or just with the last.  The vulgate reads just “castor oil root” so we
        assume that is the traditional interpretation.}
    
\item[6] 
    
He should treat the properly-pierced ear by sprinkling it with raw sesame
oil.   After every three days one should make a thicker \seEngOnly{varti}{wick} and
do the very same sprinkling.
%\footnote{The manuscripts support the reading
%    \dev{sthūlatarīṃ} that is either a non-standard form or a scribal error.}
    
\item[7] 
    
Once the ear is free from humours or side-effects, one should put in a light
\se{pravardhanaka}{dilator} in order to enlarge it enough.\footnote{Cakrapāṇidatta
    on 1.16.6 \citep[127]{acar-1939} and Ḍalhaṇa on 1.16.8 \citep[77]{vulgate} pointed
    out that the dilator can be made of wood, such as that of the
    \seEngOnly{apāmarga}{prickly chaff flower}, the \seEngOnly{nimba}{neem tree} 
    and the
    \seEngOnly{kārpāsa}{cotton plant}. Ḍalhaṇa added that it can also be made of
    \seEngOnly{sīsaka}{lead} and should have the shape of the 
    \seEngOnly{dhattūrapuṣpa}{datura
    flower}. The manuscripts have variant readings for \dev{laghupravardhanakam 
    āmuñcet} at this point that include a scribal
    emendation, none of which construe plausibly. It is possible that the unusual
    verb form \dev{ā}+$\surd$\dev{muc} puzzled the scribes and caused 
    the implausible scribal readings and emendations.} 
    
\item[8]
    
\begin{sloka}
A person's ear enlarged in this way can split in two, either as a result of the 
humours\footnote{Ḍalhaṇa on 1.16.9  \citep[77]{vulgate} notes that the word \dev{doṣa} 
here can refer to either a humour, such as \seEngOnly{vāta}{wind}, as we have 
understood it, or a 
disease generated from a humour.} or a blow.\\ Listen to me about the 
\seEngOnly{sandhāna}{ways of joining} 
it can have. 
    \end{sloka}
    
\item[9]
    
Here, there are, in brief, fifteen ways of mending the ear flap.\footnote{The Nepalese version 
uses the word \dev{sandhāna} to refer to joining a split in an ear flap, which is consistent 
with the terminology in the verse cited above (8). However, 1.16.10 of Ḍalhaṇa's version 
\citep[77]{vulgate} uses the term \dev{bandha} here and at the very beginning of the 
chapter (i.e., 1.16.1) to introduce the topic of repairing the ear.}  They are as follows:
    \se{nemīsandhānaka}{Rim-join}, \se{utpalabhedyaka}{Lotus-splittable}, 
    \se{vallūraka}{Dried Flesh}, \se{āsaṅgima}{Fastening}, \se{gaṇḍakarṇa}{Cheek-ear}, 
    \se{āhārya}{Take away}, \se{nirvedhima}{Ready-Split}, \se{vyāyojima}{Multi-joins}, 
    \se{kapāṭasandhika}{Door-hinge}, \se{ardhakapāṭasandhika}{Half door-hinge}, 
    \se{saṃkṣipta}{Compressed}, \se{hīnakarṇa}{Reduced-ear},
    \se{vallīkarṇa}{Creeper-ear}, \se{yaṣṭīkarṇa}{Stick-ear}, and \se{kākauṣṭha}{Crow's lip}.\footnote{For an artist's impression of these different kinds of joins in the ear flap, see \cite[290]{majn-1975} (reproduced as Figure 3.2 in \cite[154]{wuja-2003}).}
    
    In this context, among these, 

    \begin{description}
        
\item[\mdseries``Rim-join'':]
        both flaps are wide, long, and equal.
        
\item[\mdseries``Lotus-splittable'':]
        both flaps are round, long, and equal.
        
\item[\mdseries``Dried flesh'':]
        both flaps are short, round, and equal.
        
\item[\mdseries``Fastening'':]
        one flap is longer on the inside.
        
\item[\mdseries``Cheek-ear'':]
        one flap is longer on the outside.\footnote{For an artist's impression of this join, see \cite[291]{majn-1975} (reproduced as Figure 3.3 in \cites[155]{wuja-2003}).}
        
\item[\mdseries``Take-away'':]
        the flaps are missing, in fact, on both sides.
        
\item[\mdseries``Ready-split'':]
        the flaps are like a \se{pīṭha}{dais}.
        
\item[\mdseries``Multi-joins'':]
        one flap is small, the other thick, one flap is equal, the other unequal.
        
\item[\mdseries``Door-hinge'':]
        the flap on the inside is long, the other is small.
        
\item[\mdseries``Half door-hinge'':]
        the flap on the outside is long, the other is small.
    \end{description}

`These ten options for \seEngOnly{sandhi}{joins} of the ear should be
bound.  They can mostly be explained as resembling their
names.\footnote{Cakrapāṇidatta on 1.16.9–13 \citep[128–129]{acar-1939} and Ḍalhaṇa
    on \Su{1.16.10}{77–78} provide examples of how the names of these joins describe
    their shapes. For example, the \se{nemīsandhānaka}{rim-join} is similar to the
    join of the \se{cakradhārā}{rim of a wheel}.}  The five from
    \se{saṃkṣipta}{compressed} on are incurable.\footnote{Ḍalhaṇa on
        \Su{1.16.10}{77–78} mentions that some do not read the statement that only five
        are incurable, and they understand the causes of unsuccessful joins given below
        (i.e., heat, inflammation, suppuration and swelling) as also pertaining to the
        first ten when they do heal.}  Among these, “Compressed” has a dry ear canal and
        the other flap is small.   “Reduced ear” has flaps that have no base and have
        wasted flesh on their edges. “Creeper-ear” has flaps that are thin and uneven.
        “Stick-ear” has \seEngOnly{granthita}{lumpy} flesh and the flaps are stretched thin 
        and
        have \seEngOnly{stabdha}{stiff} \seEngOnly{sirā}{ducts}.  “Crow-lip” has a flap 
        without flesh
        with \seEngOnly{saṃkṣipta}{compressed} tips and little blood. Even when they are 
        bound
        up, they do not heal because they are hot, inflamed, 
        \seEngOnly{srāva}{suppurating}, or
        swollen.\footnote{The version of 1.16.11–13 known to Ḍalhaṇa \citep[78]{vulgate}
            has four verses (\dev{śloka}) at this point that are not in the Nepalese
            manuscripts. The additional verses iterate the types of joins required for ear
            flaps that are missing, elongated, thick, wide, etc. All four verses were probably
            absent in the version of the \emph{Suśrutasaṃhitā} known to Cakrapāṇidatta. He
            cites the verses separately in his commentary, the \emph{Bhānumatī}
            \citep[128–129]{acar-1939}, introducing each one as 'some people read' (\dev{ke
            citpaṭhanti}). However,  in Trikamajī Ācārya's edition of the \emph{Sūtrasthāna}
            of the \emph{Bhānumatī}, the root text is largely identical to the one commented
            on by Ḍalhaṇa (\cite{vulgate}), even in instances like this where Cakrapāṇidatta's
            commentary indicates that he was reading a different version of the
            \emph{Suśrutasaṃhitā}. See further the discussion on p.\,\pageref{skinflap}
            above.} 
            %            \q{The vulgate verses missing in the Nepalese version here
            %                are worth noting because they are explicit about a skin-flap
            % graft remaining
            %                connected to the site of removal.}
            %
            \item[10]
    
    % 15
A person wishing to perform a join of any of these should therefore have supplies
specially prepared according to the recommendations of the “Preparatory Supplies”
chapter.\footnote{\emph{Suśrutasaṃhitā} \Su{1.5}{18–23}, probably verse 6
    especially that lists the equipment and medications that a surgeon should have
    ready.}  And in this regard, he should particularly gather\footnote{The reading in
        the Nepalese manuscripts of \dev{viśeṣataścāgropaharaṇīyāt} has been emended to
        \dev{viśeṣataścātropaharet} to make sense of the list of ingredients, which is
        in the accusative case. Also, the repetition of \dev{agropaharaṇīyāt} in the
        Nepalese version suggests that its second occurrence, which does not make good
        sense here, is a dittographic error.} \se{surāmaṇḍa}{decanted liquor}, milk,
        water, \seEngOnly{dhānyāmla}{fermented rice-water}, and 
        \se{kapālacūrṇa}{powdered
            earthenware crockery}.\footnote{The term \dev{kapālacūrṇa} is unusual. Ḍalhaṇa
            \citep[79]{vulgate} defines it as the powder of fragments of fresh earthen pots
            and Cakrapāṇidatta \citep[129]{acar-1939} as the powder of earthenware vessels.}
    
Next, having made the woman or man tie up the ends of their hair, eat lightly and
be firmly held by qualified attendants, the physician considers the
\seEngOnly{bandha}{joins} and then applies them by means of 
\seEngOnly{chedya}{cutting},
\seEngOnly{bhedya}{splitting}, \seEngOnly{lekhya}{scarification}, or
\seEngOnly{vyadhana}{piercing}.\footnote{There are syntactic difficulties in this
    sentence.  We have %    It appears that a verb has
    %dropped out of this sentence in the Nepalese version because each word of the
    %sentence is in the accusative case with no apparent verb. Therefore, we have
    adopted the reading in Ḍalhaṇa's version \citep[78]{vulgate}, which has
    \dev{ca kṛtvā} following \dev{suparigṛhītaṃ}. It is likely that a verb, such
    as \dev{kṛtvā}, dropped out of the Nepalese transmission.}  Next, he should
    examine the blood of the ear to know whether it is \seEngOnly{duṣṭa}{tainted} or not.
    If it is tainted by wind, the ear should be bathed with
    \seEngOnly{dhānyāmla}{fermented rice-water} and water; if tainted by choler, then
    cold water and milk should be used; if tainted by phlegm, then
    \seEngOnly{surāmaṇḍa}{decanted liquor} and water should be used, and then he 
    should
    scarify it again.
       
After arranging the join in the ear so that it is neither proud, depressed, nor
uneven, and observing that the blood has stopped, one should anoint it with
honey and ghee, bandage each ear with \seEngOnly{picu}{cotton} and 
\se{plota}{gauze}, and
bind it up with a thread, neither too tightly nor too loosely.  Then, the
physician should sprinkle earthenware powder on it and  provide
\se{ācārika}{medical advice}. And he should supplement with food as taught in  the
“Two Wound” chapter.\footnote{\emph{Suśrutasaṃhitā} 4.1 \citep[396–408]{vulgate}.}
    
\item[11]
\begin{sloka}
        One should avoid rubbing, sleeping during the day, exercise, overeating,\\
        sex, getting hot by a fire, or the effort of speaking.
    \end{sloka}

\item[12]
    
    % 17
One should not make a join when the blood is too pure, too copious, or too
thin.\footnote{1.16.17 of Ḍalhaṇa's version \citep[79]{vulgate} reads “impure” for
    the Nepalese “too pure,” which would appear to make better medical sense. Emending
    the text to \dev{nāśuddha-} for \dev{nātiśuddha-} in the Nepalese recension would
    yield the same meaning as the Ḍalhaṇa's version.} For when the ear is tainted by
    wind, then it is \seEngOnly{raktabaddha}{obstructed by blood}, unhealed and will
    peel. When tainted with choler, is becomes \se{gāḍha}{pinched},
    \seEngOnly{pāka}{septic} and red.  When tainted by phlegm, it will be
    \seEngOnly{stabdha}{stiff} and itchy.  It has excessively copious
    \seEngOnly{srāva}{suppuration} and is \seEngOnly{śopha}{swollen}.  It has a small
    amount of \se{kṣīṇa}{wasted} flesh and it will not grow.\footnote{In his edition
        of \emph{Suśrutasaṃhitā}, Ācārya \citep[79 n. 1]{vulgate} includes in parentheses
        the following treatment for these conditions, which according to a footnote is not
        found in the palm-leaf manuscript he used: 'One should sprinkle it with raw sesame
        oil for three days and one should renew the cotton bandage after three days'
        (\dev{āmatailena trirātraṃ pariṣecayettrirātrācca picuṃ parivartayet}).}
    
\item[13] When the ear is properly healed and there are no complications,  one may
very gradually start to expand it.  Otherwise, it may be \se{saṃrambha}{inflamed},
burning, septic or painful.  It may even split open again.
    
\item [14]
    
Now, massage for the healthy ear, in order to enlarge it.
     
One should gather as much as one can the following: a \seEngOnly{godhā}{monitor 
lizard},
%{godhā}{Varanus bengalensis, Schneider}{Daniel 1983:58},
\seEngOnly{pratuda}{scavenging} and \seEngOnly{viṣkira}{seed-eating} birds, and
\seEngOnly{ānūpaudaka}{creatures that live in marshes or water},\footnote{For such
    classifications, see the analyses by \citet{zimm-1999} and \citet{smit-1994}.} fat, 
    marrow, milk,
    and sesame oil, and white mustard oil.\footnote{Ḍalhaṇa's version of \Su{1.16.19}{79}
        includes \seEngOnly{sarpis}{ghee}. However, Ḍalhaṇa's remarks on this passage 
        and
        Cakrapāṇidatta's on 1.16.18 \citep[130]{acar-1939} indicate that they knew a
        version of this recipe, perhaps similar to the Nepalese one, that did not include
        ghee. Ḍalhaṇa also noted that others simply read four oils, beginning with fat
        and without milk, whereas Cakrapāṇidatta said that some say it is made with four
        oils and milk.}\q{think more about the compound structure here?} Then cook the oil
        with an \seEngOnly{prativāpa}{admixture} of the following: \se{arka}{purple 
        calotropis},
        %{Calotropis gigantea, (L.) R. Br.}{ADPS 52, AVS
        % 1.341, NK \#427, Potter 57, ID 306},
        \se{alarka}{white calotropis}, %{Calotropis procera, (Ait.) R. Br.}{NK
        % \#428,
        % GIMP 46b, ID 306},
        \se{balā}{country mallow}, %{Sida cordifolia, L.}{ADPS 71, NK \#2297},
        \se{atibalā}{`strong Indian mallow'}, %{Abutilon indicum, (L.) Sweet; Sida
        % rhombifolia, L.?}{NK \#11, IGP ,4 1080; NK \#2300},
        %\plant{Indian sarsaparilla}{anantā}{Hemidesmus indicus, (L.) R.
        %  Br. \textnormal{and} Cryptolepis buchanani, Roemer \&
        %  Schultes}{ADPS 434, AVS 3.141, NK \#1210},
        %the \skt{Indian sarsaparillas}{sārive}
        \se{anantā}{country sarsaparilla}, %{Hemidesmus indicus, (L.) R. Br.}{ADPS
        % 434,AVS 3.141--5, NK \#1210}
        %and
        %\plant{black creeper}{pālindī}{Ichnocarpus frutescens, (L.)
        %    R.Br. \textnormal{or} Cryptolepis buchanani, Roemer \&
        %    Schultes}{AVS 3.141, 3.145, 3.203, NK \#1283, \#1210, ADPS
        %    434}),
        %
        %%\plant{prickly chaff-flower}{apāmārga}{Achyranthes aspera,
        %%    L.}{GJM 524f., IMP 1.39, ADPS 44f., IMP 3.2066f., Dymock 3.135},
        %\plant{Withania}{aśvagandhā}{Withania somnifera (L.) Dunal}{IMP
        %    5.409f., Dymock 2.566f., Chevallier 150.},
        \se{vidāri}{beggarweed}, %{Desmodium gangeticum (L.) DC}{Dymock 1.428, GJM
        % 602, cf.\ NK \#1192; ADPS 382, 414 and IMP 2.319, 4.366 are confusing},
        %\plant{giant potato}{kṣīraśukla  $\rightarrow$ kṣīravidārī}{Ipmoea
        % mauritiana,
        %    Jacq.}{ADPS 510, AVS 3.222, IMP 3.1717ff.},
        \se{madhuka}{liquorice} and hornwort (\dev{jalaśūka} $\rightarrow$
        \dev{jalanīlikā}\footnote{Ceratophyllum demersum, L. %(IMP 2371, AVS
            % 2.56,
            % IGP 232).
            This name is not certain. In fact, Ḍalhaṇa on 1.16.19
            \citep[79]{vulgate} notes that some people interpret it as a
            poisonous, hairy, air-breathing, underwater creature.}).%
            %items having the `sweet' savour (\dev{madhuravarga}\footnote{The
            %items which exemplify the `sweet' savour \label{kakolyadi}
            % (\dev{madhuravarga}) are enumerated at SS.1.42.11.})
            %and `milk
            %flower'(\dev{payasyā}  $\rightarrow$ \dev{vidārī}\footnote{Pueraria
            %tuberosa (Willd.) DC.
            %(ADPS
            %    510, IMP 1.792f., AVS 4.391; not Dymock 1.424f. See GJM
            % supplement 444,
            %    451, IMP 1.187, but IMP 3.1719 = Ipmoea mauritiana, Jacq.).
            \footnote{The version of 1.16.19 known to Ḍalhaṇa \citep[79]{vulgate}
                adds several ingredients to this admixture, including \dev{apāmārga,
                aśvagandhā, kṣīraśuklā, madhuravarga} and \dev{payasyā}. Also, it has
                \dev{vidārigandhā} instead of \dev{vidāri}. When commenting on
                1.16.19, Ḍalhaṇa \citep[79]{vulgate} notes that some do not read
                \dev{madhuravarga} and \dev{payasyā}. Therefore, there were probably
                other versions of this recipe with fewer ingredients, as seen in the
                Nepalese version.} %
                This should then be deposited in a well-protected spot.
    
\item[15]% 20
    \begin{sloka}
The wise man who has been sweated should rub the \seEngOnly{mardita}{massaged} 
ear with
it.\\ Then it will be free of complications, and will enlarge properly and be
strong.\footnote{For these aims (i.e., healing and enlarging the ear), the text
    known to Ḍalhaṇa \citep[79]{vulgate} has an additional verse and a half describing
    an \seEngOnly{udvartana}{ointment for rubbing the ear} and 
    \seEngOnly{taila}{sesame oil} cooked
    with various medicines for massage. Cakrapāṇidatta \citep[131]{acar-1939} does not
    comment on these verses, nor verse 15 of the Nepalese version, and so the version
    of the \emph{Suśrutasaṃhitā} known to him may not have included them.}
    \end{sloka}
    
\item[16]
    % 22cd-23
        \begin{sloka}
Ears which do not enlarge even when sweated and oiled,  should be scarified  at the
\seEngOnly{apāṅga}{edge of the hole}, but not outside it.\footnote{Dalhaṇa's version of
    \Su{1.16.23}{79--80} adds another hemistich that states more explicitly that the
    scarification should not be done on the outside of hole as it will cause
    derangement.}
        \end{sloka}
    
\item[17]
          \begin{sloka}
    In this tradition, experts know countless repairs to ears.  So a 
    physician who is \seEngOnly{suniviṣṭa}{very intent} on working in this way 
    \seEngOnly{yojayed}{may repair} them.\footnote{After verse 17, the 1938 edition of 
    Ācārya 
    \citep[80]{vulgate} has in parentheses nineteen verses on diseases of the ear lobes, 
    treatments and complications. It is possible that these verses were in some of the 
    witnesses used by Ācārya to construct the text as they occur in other manuscripts, such as  
    MS Hyderabad Osmania 137-3 (b). However, Cakrapāṇidatta \citep[132]{acar-1939} and 
    Ḍalhaṇa \citep[80]{vulgate} state that some read about the diseases of the ear lobes in 
    this chapter whereas others read about them in the chapter on 
    \se{miśrakacikitsa}{various treatments} (SS 5.25), which does indeed begin with a 
    discussion of the disease \dev{paripoṭa}.  Ḍalhaṇa goes on to say that some believe that 
    these verses were not composed by sages and, therefore, do not read them.}
        \end{sloka}
    
\item[18]
    % 25
           \begin{sloka}
    If an ear has grown hair, has a nice hole, a firm join, and is strong and
    even, well-healed, and free from pain, then one can enlarge it slowly.\footnote{The order of verses 17 and 18 are reversed in Ḍalhaṇa's version \citep[80]{vulgate}.}
        \end{sloka}
    
\item[19]
    
    %28, 29.
   \begin{sloka}
        Now I shall describe the proper method of making a repair when a nose is severed.
    First, take from the trees a leaf the same size as the man's nose and hang it
    on him. 
   \end{sloka}
    
\item[20] 

\begin{sloka}
    Next, having cut a \se{vadhra}{slice of flesh},\footnote{The
    version of 1.16.28b known to Ḍalhaṇa \citep[81]{vulgate} reads “\se{baddham}{bound, 
    connected}” instead of “\se{vadhra}{slice of flesh}.”
    This is a critical variant from the surgical point of view.  If the slice remains
    connected, it will have a continuing blood supply.  This is one of the effective 
    techniques that so astonished surgeons witnessing a similar operation in Pune in
    the eighteenth century \citep[see][67--70]{wuja-2003}.} with the same
    measurements, off the cheek, the end of the nose is then scarified.\footnote{Or 1.16.20 
    could be mean, 
    `\ldots\ off the cheek, it is fixed to the end of the nose, which has been
    scarified.' Unfortunately, the Sanskrit of the Nepalese version is not unambiguous on the
    important point of whether or not the flap of grafted skin remains connected
    to its original site on the cheek. However, Ḍalhaṇa \citep[81]{vulgate} clarifies the 
    meaning of the vulgate here by stating that one should supply the word 'flesh' when 
    reading 'connected,' thus indicating that he understood the flesh to be connected to the 
    face.} % 
%
Then the \seEngOnly{apramatta}{undistracted} physician, 
    should quickly put it back together so that it is 
    \seEngOnly{sādhubaddha}{well joined}.
    \label{well-joined}
\end{sloka}  


\item[21] 
    % 30.
\begin{sloka}
Having carefully observed that it has been sewn up properly, he should then fasten
it along with two tubes.\footnote{Ḍalhaṇa noted that the two tubes should be made
    of \seEngOnly{nala}{reed} or the \seEngOnly{eraṇḍapatranāla}{stalk of the leaf of 
    castor oil
    plant} (on \Su{1.16.21}{81}). They should not be made of lead or betel nut because
    the weight will cause them to slip down.} Having caused it to be
    raised,\footnote{The Sanskrit term \dev{unnāmayitvā} in 1.16.21 is non-Pāṇinian.}
        the powder of \seEngOnly{pattāṅga}{sappanwood},\footnote{Caesalpinia sappan, 
        L. %(AVS
            % 1.323, IMP 2.847f.).
            For \dev{pattāṅga} there are manuscript variants \dev{pattrāṅga}
            (\MScite{Kathmandu NAK 5-333}) and \dev{pattaṅga} (\MScite{Kathmandu NAK
            1-1079}).  Also, \MScite{Kathmandu KL 699} (f.\,14r:1) has \dev{pattrāṅga} in
            a verse in 1.14 (cf.\ \Su{1.14.36}{66}). The text known to Ḍalhaṇa  has
            \dev{pataṅga} (\Su{1.16.29}{81}) and this term is propagated in modern
            dictionaries.} %\se{yaṣṭīmadhuka}
            {liquorice} %{Glycyrrhiza glabra, L.}{AVS 3.84, NK \#1136},
            and Indian barberry.\footnote{Berberis aristata, DC. % (Dymock 1.65, NK 
                % \#685, GJM 562, IGP 141).
                Ḍalhaṇa understands it as \se{rasāñjana}{elixir salve}
                (\cite[81]{vulgate}).}\q{añjana} should be sprinkled on it.

\end{sloka}    
\item[22] 
\begin{sloka}
 The wound should be covered properly with \seEngOnly{picu}{cotton} and should be
moistened repeatedly with sesame oil.  Ghee should be given to the man to drink. 
His digestion being complete, he should be oiled and purged in accordance with
the instructions specific to him.\footnote{The expression \dev{svayathopadeśa}
    is ungrammatical but supported in all available witnesses.}
\end{sloka}
    
\item[23] %32.
\begin{sloka}
And once healed and really come together, what is left of that \se{vadhra}{slice
    of flesh} should then be trimmed.\footnote{The vulgate transmission has lost the
    word \dev{vadhra} and replaced it with \dev{ardha} “half,” which makes little
    sense in this surgical  context.} If it is \seEngOnly{hīna}{reduced}, however, one 
    should
    make an effort to stretch it, and one should make its overgrown flesh
    smooth.\footnote{Ḍalhaṇa  accepted a verse following this, \Su{1.16.32}{81}, which
        points out that the procedure for joining the nose is similar to that of joining
        the lips without fusing the ducts. He noted that earlier teachers did not think
        this statement on the nose and lips was made by sages, but he included it because
        it was accepted by Jejjaṭa, Gayadāsa and others, although they did not comment on
        it because it was easy to understand. Cakrapāṇidatta also did not comment on this
        additional verse \citep[133]{acar-1939}.}

\end{sloka}    
    
\end{translation}    
