\RequirePackage{snapshot}
% !TeX encoding = UTF-8
% !TeX spellcheck = en_GB
\documentclass[12pt]{article}
\usepackage{xelatex-generic-biblatex-indexing,
    booktabs, % better-looking tables
    wasysym, %for the blob symbols in the table
    CC-license-footer,
 %   coffeestains,
}
\hangfootnotes

\addbibresource{incremental_SS_Translation.bib}

\usepackage{hyperref}
\usepackage[
    text={Draft of \today\ for private study only},
    color={[gray]{0.95}},
    scale=.27]{draftwatermark}

% just print hyperlinked “URL” in the bibliography, not the long, ugly URL.
\DeclareFieldFormat{url}{%
    \ifhyperref
   % {\textsc{url: }\href{#1}{URL}}
    {\href{#1}{\textsc{url}}}
    {\textsc{url: }\nolinkurl{#1}}} %DW bug here
\DeclareFieldFormat{eprint:ark}{%
%    \textsc{ark:}\space
    \ifhyperref  
    {\href{https://n2t.net/#1}{\nolinkurl{#1}}}
    {\nolinkurl{#1}}}


\setlist[enumerate]{noitemsep} % enumitem global setting for \items
\setlist[itemize]{noitemsep} 

%% Make a new environment for the translation with tweaked spacing.
%% The old-fashioned LaTeX way:
%\newenvironment{translation}
%    {\begin{itemize}[noitemsep,
%            left=0pt .. \parindent]}
%    {\end{itemize}}

%% Make a new environment for the translation with tweaked spacing.
%% Do it the enumitem way:
\newlist{translation}{enumerate}{2}
\setlist[translation]{label=\textbullet,
   noitemsep,
    left=0pt .. \parindent}


\newcommand{\Su}[2]{SS~#1
    \citep[#2]{vulgate}}
% references to the Suśrutasaṃhitā
% #1 = canonicau
% #2 = page of Ācārya's third edition
% later, when we've got our own numbers for our edition, we'll add them 
% as the second argument.

\newcommand{\AS}[2]{AS #1 
    \citep[#2]{atha-1980}} % references to the Saṅgraha, 
% #1 = canonical
% #2 = page of Aṭhāvale's edition

\newcommand{\Ah}[2]{Ah #1 
    \citep[#2]{kunt-1902}} % references to the Saṅgraha, 
% #1 = canonical
% #2 = page of Kunte and Navare's edition


%
%\newcommand{\item}[1]{$^{#1}$}

%\newenvironment{translation}{\list }{\endlist }

\renewcommand{\SS}{\emph{Suśrutasaṃhitā}}

\title{A Translation of the New Edition of the \SS}
\author{Jason Birch and Dominik Wujastyk}
\date{\texttt{Draft of \today}\\ \copyright\ Jason Birch and Dominik Wujastyk}

\let\se\saneng  % just make typing a bit quicker

\begin{document}
       \input{sanskrit-hyphenations}
    
    %\nocite{forb-1856}
    \maketitle
    
    % \coffeestainA{0.9}{0.85}{-25}{5cm}{1.3cm}
    
        \noindent The \href{http://sushrutaproject.org}{Suśruta Project} is
producing
\href{https://saktumiva.org/wiki/wujastyk/susrutasamhita/01-su.su/provisional-edition_sutrasthana}{a
 new Sanskrit text edition} of the \emph{Suśrutasaṃhitā} based on the early 
Nepalese manuscripts.  As we gradually transcribe and edit the manuscripts, we are 
producing this new translation of the classic work.
\MSnocite{Kathmandu, KL 699}
\MSnocite{Kathmandu, NAK 1-1079}
\MSnocite{Kathmandu, NAK 5-333}
        
        \tableofcontents
        
        \newpage
        \section{The manuscripts and editions used in the vulgate editions by 
        Yādavaśarma Trivikrama Ācārya}
    
        Yādavaśarma Trivikrama Ācārya produced three successive editions of the
\SS\ with the commentary of Ḍalhaṇa, in 1915, 1931 and 1938.  These
editions, especially the last, are generally considered the most scholarly
and reliable editions of the work, and have been constantly reprinted up
to the present day.
        
The 1915 edition was based on three manuscripts.  The 1931 edition used another
seven plust two printed editions.  For his final 1938 edition, Ācārya used a
further three manuscripts.\footnote{The following account of the sources is
paraphrased from \citeauthor{susr-trikamji3}'s own account of his sources
\citep[22]{susr-trikamji3}.}  These sources are described as follow, with an
overview in Table~\ref{dw:tableofeds}.
      
        \subsection{The sources of the 1915 edition}
        
        \begin{enumerate}
            \item[1] Calcutta, Royal Asiatic Society.  Covers the \emph{sūtra, nidāna, śārīra and 
            kalpa sthāna}s.  
            
            \item [2] Jaipur, Pandit Gaṅgādharabhaṭṭaśarman, lecturer at the Royal 
            Sanskrit University.  Covers the \emph{cikitsāsthāna} and the \emph{uttaratantra}.
            
            \item [3]  Bundi, my great friend the royal physician Paṃ.\ Śrīprasādaśarman  
            Covers the \emph{uttaratantra}.
        \end{enumerate}
        
        \subsection{The sources of the 1931 edition}

\begin{enumerate}

    \item[1] Vārāṇasī, professor of literature, the great Gaurīnāthapāṭhaka.  With the 
    \emph{Nibandhasaṅgraha}. Covers the \emph{nidānasthāna} and \emph{uttaratantra}.
 
    \item [2]  Ahmedabad.  My friend Sva.\ Vā.\ Vaidya Raṇachoḍalāla Motīlālaśarman.  
    With the \emph{Nibandhasaṅgraha}.  Covers the \emph{śārīrasthāna}.
    
    \item [3] From the personal library of my great friend Sva.\ Vā.\ Vaidya
Murārajīśarman. Extremely old. No commentary.  Covers the \emph{śārīrasthāna}.
    
    \item [4]  Puṇe, BORI library.  With the \emph{Nibandhasaṅgraha}. Covers the
\emph{śārīrasthāna}.\footnote{Not one of the three MSS of the
\emph{śārīrasthāna} described in \cite{shar-vaid}.}
    
    \item [5]  Puṇe, BORI library.  With the \emph{Nibandhasaṅgraha}. Complete.  
    With some damaged folia.
    
    \item [6]  Bombay, Asiatic Society.  Incomplete.\footnote{Possibly \MScite{Mumbai, 
    AS B.I.3} or \MScite{Mumbai, AS B.D.109} \citep[v.\,1, \# 212 and 
    213]{vela-1930}.  But both these have the \emph{Nibandhasaṅgraha}.  The first 
    covers only the \emph{śārīrasthāna}; the second may be complete, but Velankar calls it 
    only “disorderly.”}
    
    \item [7] Varanasi, the private library of Vaidya Tryambakaśāstrī.  Covers the 
    \emph{cikitsāsthāna}.  The variant readings of this MS were compiled by Prof.\ 
    Guruprasādaśāstrī and supplied to Ācārya.
    
    \item [8]  A printed edition together with the commentary 
    \emph{Suśrutasandīpanabhāṣya} by Professor Hārāṇacandra Cakravārtti. Complete work.
    This is the 1910 Calcutta edition numbered “t” by \citet[IB, 
    312]{meul-hist}.\footcite{bhat-1917}
    
    \item [9] A printed edition of the first 43 chapters of the
\emph{sūtrasthāna}, printed in Bengali script, with the commentaries
\emph{Bhānumatī}, \emph{Nibandhasaṅgraha}, edited by Vijayaratnasena and
Niśikāntasena. This is the 1886 Calcutta edition numbered “g” by \citet[IB,
311]{meul-hist}.\footcite{sena-1886}

\end{enumerate}

   
        \subsection{The sources of the 1938 edition}
        % \coffeestainC{1}{1}{180}{0}{-5 mm}
        \begin{enumerate}
    \item [1]  Gwalior, from the library of my great friend Paṃ.\ Rāmeśvaraśāstrin Śukla. 
    Covers the \emph{sūtra, nidāna, śārīra, cikitsā and kalpasthāna}s.
     
            \item[2] Bikaner, from the library of the Royal Palace, supplied by Paṃ.\ 
            Candraśekharaśāstrin. Contains the commentary 
            \emph{Nyāyacandrikāpañjikāvyākhyā} by Gayadāsa.  Covers the 
            \emph{nidānasthāna}.  This is almost certainly \MScite{Bikaner Anup 
            4390}.\footnote{See Dominik Wujastyk, “MS Bīkāner AnupLib 4390.” \emph{Pandit}. 
            <\url{http://panditproject.org/entity/108068/manuscript}>.}
            
            \item [3] Kathmandu, located in the private library of the Royal Guru Hemarāja 
            Śarman.  An extremely old palm-leaf manuscript. Readings from this MS were 
            compiled by Paṃ Nityānandaśarman Jośī and sent to Ācārya. Covers from the 
            beginning of the work to the end of the ninth chapter of the \emph{cikitsāsthāna}.
        \end{enumerate}

    \begin{table}
        \begin{tabular}{c|ccc|ccccccccc|ccc}
            \toprule
              \multicolumn{16}{c}{\emph{Manuscripts (\newmoon) and print editions 
($\circ$)}} \\
 \emph{edition}            &\multicolumn{3}{c}{1915}
&                \multicolumn{9}{c}{1931} 
  &              \multicolumn{3}{c}{1938} \\
                 \emph{sthāna}  & 1 & 2 & 3 & 1 &2  &3  &4  &5  &6  &7  &8  &9  &1  
                 &2 &3 \\
            \midrule

             \emph{sū}. &  \newmoon&  &  &
               &  &  &  & \newmoon & ? &  & $\circ$ & $\circ^\textrm{\footnotesize(1--43)}$ &  
             \newmoon & &\newmoon \\
            
             \emph{ni}. &\newmoon  &  &  &
              \newmoon &  &  &  &  \newmoon&  ?&  & $\circ$ &  &  
              \newmoon&\newmoon & \newmoon\\
            
             \emph{śā}. &  \newmoon&  &  &
              & \newmoon & \newmoon & \newmoon & \newmoon &  ? &  &  
              $\circ$&  &  
              \newmoon& &\newmoon \\
            
             \emph{ci}. &  & \newmoon &  &
               &  &  &  &\newmoon & ? &  \newmoon&$\circ$  &  &
              \newmoon & &\newmoon $^\textrm{\footnotesize(1--9)}$ \\
            
             \emph{ka}.  &\newmoon  &  &  &
               &  &  &  &\newmoon  &  ?&  & $\circ$ &  &  
             \newmoon  & & \\

            \emph{utt}.  &  & \newmoon &\newmoon  &
            \newmoon  &  &  &  & \newmoon & ? &  & $\circ$ &  &  
            & & \\
            \bottomrule
       
        \end{tabular}  
       \caption{The sources of Yādavaśarman T. Ācārya's three vulgate editions.}
           \label{dw:tableofeds}
        \end{table}
        
        \subsection{Evaluation}
        
        Estimates show that there are approximately 230 extant manuscript
witnesses for the \emph{Suśrutasaṃhitā}.\footnote{This figure is arrived
at by summing the MSS mentioned in \cite{ncc} and in the \cite{ngmcp}. The
real figure could be many scores higher.}  Many of these manuscripts cover only
one or more or its chapters.  Nevertheless, this is an order of magnitude
more evidence than was considered by Ācārya for his vulgate editions.
        
While the descriptions provided by Ācārya of his source materials seems at first
to be moderately comprehensive, Table~\ref{dw:tableofeds} reveals the real paucity
of textual basis for these editions.  Apparently, fifteen manuscripts were
consulted.  However, we quickly find that two of the sources were other people's
printed editions, and one of those covered less than a quarter of the work.  That
reduces the manuscript base to 13 manuscripts.   Ācārya does not appear to have
seen two of the manuscripts at all, having been sent collations prepared for him
by others (7 of 1931 and 3 of 1938).  Thus, Ācārya's final edition was based on
the personal consultation of eleven manuscripts.   One of them, no.\,6 of 1931,
remains mysterious. Only a single manuscript covers the whole of the
\emph{Suśrutasaṃhitā}, no.\,5 of the 1931 edition.  Manuscript 1 of 1938 is the
next most complete, but it omits the \emph{uttaratantra}, which comprises a third
of the work.  Manuscript 1 of the 1915 edition is next, but it still omits both of
the longest chapters, and thus offers less than half the work.  For the rest, the
evidence is spotty, with each part of the work being supported by only between
four and eight manuscripts, excluding the printed editions.

Two sources stand out for their historical importance.  The first is no.\,3 of
1931, which Ācārya calls “extremely old.”  It covered the \emph{śārīrasthāna}
only, and unfortunately we know nothing of the later history of this manuscript. 
The second is no.\,3 of 1938, which is one of the important Nepalese manuscripts
being considered in the present project. Ācārya's remarks do not allow us
to identify this manuscript probatively. It should be \MScite{Kathmandu NAK 5-333}
\citep[\S 2.3]{kleb-2021b}, but that manuscript covers the whole work, not just up to the
ninth chapter of the \emph{cikitsāsthāna}. Perhaps Ācārya only received collations
for this portion of the MS and did not know that the MS was a witness for the
whole work.
        
\subsection{The 1939 edition}        

In 1939, Yādavaśarman Trivikrama Ācārya produced an edition of the
\emph{sūtrasthāna} of the \emph{Suśrutasaṃhitā} that was published by the Swami
Laxmi Ram ayurvedic centre in Jaipur, and printed at the famous Nirṇayasāgara
Press in Mumbai.\footnote{\cite{acar-1939}. On the Swami Laxmi Ram centre, see
\cite{hofe-2007}.}  The text was edited on the basis of the following sources:
\subsubsection{For the Bhānumatī}
\begin{enumerate}
    \item Printed ed. Siglum \dev{mu} for \dev{mudrita}.  Had the Bhānumatī up to chapter 
    Su.sū.40.
    \item IOLR MS provided through BORI \dev{ha} for \dev{hastalikhita}. Had the Bhānumatī, 
    to the end of Su.sū. 
    \end{enumerate}
\subsubsection{For the Suśrutasaṃhitā}
\begin{enumerate}
    \item Palm leaf MS from Hemarājaśarman's library, \dev{tā} for \dev{tāḍapatra}
    \item His own published edition. \dev{ḍa} for \dev{ḍalhaṇasaṃmataḥ pāṭhaḥ}
    \item Hārāṇacandra Cakravarti's published edition with his own commentary. \dev{hā}
\end{enumerate}
   
   Ācārya noted in his introduction that the manuscripts containing the Ḍalhaṇa's
commentary all came with the root-text of the \SS\ too, and thus the main \SS\
text reflected the readings chosen by Ḍalhaṇa.  But the MSS of the
\emph{Bhānumatī} contained the commentary alone and had many explanations based
on different readings of the root-text than those of Ḍalhaṇa.  In many of these
cases it was hard to know what the text that Cakrapāṇidatta had before him. 
But it seemed that Cakrapāṇidatta had a text before him that had much in common
with the text of the Nepalese manuscript.\footcite[3--4]{acar-1939} 


\newpage
    
    \section{Sūtrasthāna, adhyāya 1}
    
    % \coffeestainB{0.7}{1}{-30}{18 pt}{-50 pt}
    
\begin{translation}

    \item[1] Now I shall narrate the chapter on the origin of this
knowledge.\footnote{Ḍalhaṇa understood the word "\se{veda}{knowledge}" as
specifically "medical knowledge." He said that the word "longevity"
(\emph{āyur}) \ssaneng{āyur}{life, longevity} had been elided.
%    
%    Notes Dec 8:
%    Dominik: N's ādhyāyaṃ corruption of H's nāmādhyāyaṃ: possible evidence that N was created 
%after H
%    Check: Ācārya 1931 footnote on vedotpattim
%    Commentary Ḍalhaṇa (c. 1200 CE) notes āyur dropped from veda in vedotpattim
%    
%    
After this opening statement, later manuscripts and commentaries include
the attribution, "as the venerable Dhanvantari stated."  The absence of this
statement in the early Nepalese MSS is highly noteworthy because it removes
the outer narrative frame of the \SS\
\parencites[148]{wuja-2013}[\S\,3.1.2]{kleb-2021b}.  On the figure of Dhanvatari in 
medical literature, see \cite[IA 358--361]{meul-hist}.} %     <!-- Notes Dec 8:
%    Dom: note the omission of Dhanvantari, which is in the edition.
%    On Dhanvantari, see Meulenbeld HIM, authorities associated with Suśruta. He's
% an authority on
%surgery or toxicology-->
    
    \item[2] Now, as is well-known, Aupadhenava, Vaitaraṇa, Aurabhra, Puṣkalāvata,
Karavīra, Gopurarakṣita, Bhoja, Suśruta and others addressed Lord Divodāsa,
king of Kāśi, the best of the immortals, who was in his ashram surrounded by
an entourage of sages.\footnote{On these persons, see \cite[IA
361--363, 369\,ff.]{meul-hist}. The authority Bhoja does not appear in the list as
published in the vulgate edition \citep[1]{susr-trikamji2}, and was not
included in \cite{meul-hist} amongst “authorities mentioned in the \SS.” 
\citeauthor{meul-hist} gathered textual evidence about Bhoja at \cite[IA
690--691]{meul-hist}. \citet{kleb-2021a} has discussed these authors in the
context of an anonymous commentary on the \SS\ that cites them.}

\nocite{emen-1969}
    
%    Notes Dec 8:
%    Dom: Check these names in Meulenbeld 
%    Bhoja is an early lost authority on medicine. Not the same person as King Bhoja, commentator 
%on the Yogasūtras.
%    Ḍalhaṇa's comm. mentions Bhoja as also included in prabhṛtayaḥ: so the version of the text he 
%was using did not mention Bhoja, but he was aware of him: His provenance makes it possible that 
%he knew the Nepalese version of the SŚ
    
    \item[3]
%O Lord, after seeing people who are assailed by the impingements of various pains 
%caused by 
%physical, mental and accidental diseases, who have the support of friends [but] 
%feeling as if they 
%were alone, and acting frantically, shouting out, we have been distressed. 

“O Lord, distress arose in our minds after witnessing people thrashing about with
cries, assailed by different kinds of \se{vedanābhighāta}{pain and injury}, 
feeling helpless in spite 
of having friends, because of diseases arising from the body, the mind and
external sources.


    
%    Notes Dec 8:
%    āgantu - caused by something from outside the body
%    abhighāta - threats, impingements 
%    vedanābhighāta - tatpuruṣa 
%    anātha - among a list of people who shouldn't be treated.
%    Ḍalhaṇa- sanātha: samitra someone with a friend -->

    \let\uncertain\texttt
    
\item[4]    
“To quell the illnesses of those who seek happiness and for our own purpose of
prolonging life, we desire \se{āyurveda}{the science of life} that is being
taught.  Welfare\ssaneng{śreyas}{welfare}, both in this world and in
    the next, depends upon it. Therefore, we have come to the Lord in pupillage." %
% reading bhagavan (voc.) and tam (it, ayurveda, masc. acc.)

% we think upasannāḥ smaḥ is probably wrong, but we can't see how to improve it.
% upapannā sma ?

\item[5] The Lord said to them:

“Welcome to you!  My children, all of you are beyond reproach and worthy 
to be taught.
   
    
    \item[6] 
%    "As is well-known in this world, before creating people, Brahmā composed 
%    what is called Āyurveda.\footnote{The relative pronoun \emph{yad}, that has 
%no  
%    correlative \citep[\P 461]{spei-1886}, is omitted.} 
%    It is taught as part of the \emph{Atharvaveda}, in hundreds of 
%    thousands of verses and a thousand chapters and, after observing the short 
%    lifespan and low intelligence of people, made it again in eight parts. 
%    %infer tat as the object of kṛtavān
%    
    
    “As is well known, Ayurveda is the name of what is said to be the subsidiary
part of the Atharvaveda.   Before creating people, Svayambhū composed it in
hundreds of thousands of verses and a thousand chapters and, after observing the
short lifespan and low intelligence of people, he presented it again in eight
parts.\footnote{Svayambhū is another name for Brahmā, the creator.}
    
    \item[7] “Surgery, treatment of body parts above the clavicle, general medicine, 
    knowledge of spirits, care of children, and the disciplines of antidotes, rejuvenation 
    and aphrodisiacs.
    % why do some of the auxiliaries end in tantra? Dom: Some were disciplines that had a separate life outside āyurveda. The others were more particular to vaidyas.
  
        \item[8] “Now,  a collection of the characteristics of each component of 
        Āyurveda.
    
        \item[9] “Among them, [the component] called surgery has the goal of 
        extracting 
        various grasses, wood, stone, dust, iron (?), soil, bone, hair, nails, discharge of 
        pus, malignant wounds and foreign bodies inside the womb, and of determining 
        the application of surgical instruments, knives, caustics and fire by means of 
        sixty definitions.
        %Ḍalhaṇa seems to read duṣṭavraṇāntar, and glosses antar as madhyāt. He then reads garbhaśalya (HIM - foetuses stuck in the womb). Ḍalhaṇa is aware of the reading ṣaṣṭyā vidhānaiḥ (following uddhraṇārtha), and says some explain it as apatarpaṇādyai rakṣāvidhānāntair dvivraṇīyoktair ity arthaḥ. However, mss. 699 and 533 read abhi° not vi°
        
        \item[10] “[The component] named the doctrine of treating body parts above 
        the clavicles has the aim of curing diseases situated above clavicles that is,  
        diseases located in ears, eyes, mouth, nose and so on.
        
        \item[11] “[The component] called general medicine has the goal of curing 
        illnesses established in the whole body and [diseases] such as fever, tumour, 
        swelling, hemorrhagic disorders, insanity, epilepsy, urinary diseases, diarrhoea 
        and the like.
        
        \item[12] “[The component] called knowledge of spirits is for appeasing
demons by pacification rites and making food offerings for those whose
minds have been possessed by gods, their enemies,\footnote{Dānavas.  The
insertion marks (\emph{kākapada}s) below the text at this point appears to
be by the original scribe.} Gandharvas, Yakṣas, demons, deceased
ancestors, Piśācas, Vināyakas, \footnote{The vulgate doesn't have
\emph{vināyaka}s but does add \emph{asura}s, probably under the influence
of Ḍalhaṇa.  Cite Paul Courtright, Ganesha book.} Nāgas and evil spirits
that possess children. % Notes: vināyaka is omitted from the vulgate. In
% Mahābhārata, etc. It refers to
%a class of demons.-->
          
        
        \item[13] “[The component] called care of children is for bearing children and 
        purifying defects in a wet-nurse's milk, and curing diseases that have arisen 
        from bad breast milk and demons.
        
        \item[14] “[The component] called the discipline of toxicology is for
[knowing] the signs of poison from snake and insect bites and for
neutralising various combinations of poisons.\footnote{The scribal
insertion marks (crosses) above the line at this point in MS K appear to
be in a later hand and their referent is lost in the damaged part of the
folio.  Although MSS N and H include \se{lūtā}{spiders} and
\se{sarīsṛpa}{creepy-crawlies} in the list, it does seem that MS K had
a shorter list, and the vulgate edition adds \se{mūṣika}{rodents}.}
        
        \item[15] “[The component] called the discipline of rejuvenation is 
        maintaining 
        youth, bringing about a long life and mental vigour and for curing diseases.
        
        % Got to here 2021-01-13
        
        \item[16] “[The component] called the discipline of aphrodisiacs brings about 
        the 
        increase, purity, accumulation and  production of semen for those whose semen 
        is minimal, bad, depleted, and dry [respectively] and for inducing an erection.
        
        \item[17] “Thus, this Āyurveda is taught with eight components."
        
        "Among these [components], tell us which is for whom."
        
        \item[18] They said, "After you have conveyed the knowledge of surgery, 
        teach 
        us everything."
        
        \item[19] He said, "so be it."
        
        \item[20] They then said, "Having considered the view of all of us, when we 
        are 
        unanimous, Suśruta will question you. We too will learn what is being taught to 
        him."
        
        \item[21] He said, "so be it.
        
        \item[22] “Now, as is well-known, the aim of Āyurveda is eliminating the 
        disease of one who have been assailed by disease and protecting the healthy;  
        āyurveda is [that knowledge] in which they find a long life, or that by which 
        long life is known. Learn its best component (i.e., surgery), which is being 
        taught in accordance with tradition, perception, inference and analogy.
        
        \item[23] "For this component is first, the most important, because it is 
        referred to first; it cures wounds and joins together the most important thing, 
        Yajña's head. For, just as it has been said of old, 'the head that had been cut off 
        by Rudra was joined again by the two Aśvins.'
        
        \item[24] "And also, of the eight disciplines of Āyurveda, [surgery] alone is 
        the best because of the quick action of its \se{kriyā}{procedures}, its 
        application of blunt 
        instruments, knives, caustics and fire, and it is common to all disciplines.
        
        \item[25] "Therefore, [surgery] is eternal, meritorious, leads to heaven, 
        brings renown, bestows a long life, and affords a livelihood.
        
        \item[26] "Brahmā said this, 'Prajāpati learned it. From him, the Aśvins. From 
        the Aśvins, 
        Indra. From Indra, I. In this world, I will transmit to those who desire it for the benefit of 
        people.' 
        
        [There a verse about this.].\footnote{This is an expansion 
        of the scribe's abbreviation \emph{bha} for \emph{bhavati cātra ślokaḥ} 
        “There is a verse about this” (sometimes plural).\label{bha}}
        
        \item[27]           
        \begin{sloka}
            For, I (i.e., Brahmā) am Dhanvantari, the first god, the remover of old age, 
            pain and 
            death of mortals.\\ Having understood surgery, the best of the great 
            knowledge 
            systems, I arrived on earth again to teach it here.
        \end{sloka}    
        
        % draft tr.
        
        \item[28] In this context, as far as this discipline is concerned, a 
        \se{puruṣa}{human being} is called an amalgam of the five elements 
        and the embodied soul.  This is where \se{kriyā}{procedures} apply. This 
        is the locus. 
        
        Why?
        
        Because of the duality of the world, the world is twofold: the stationary
and the moving. Its \se{ātmaka}{nature} is twofold, depending on the
preponderance of Agni and Soma.\footcite[See][]{wuja-2004}  Alternatively,
it can be considered as being fivefold.  The multitude of beings in it are
fourfold: they are termed “sweat-born, stone-born, caul-born and
egg-born”.\footnote{This fourfold classification of beings is paralleled
with closely-related vocabulary in  \emph{Bhelasaṃhitā} 4.4.4	
\parencites[206]{kris-2000}[81]{mook-1921}.}  Where they are concerned, the
human being is the main thing; others are his support.  Therefore, the
\se{puruṣa}{human being} is the locus.

\item[29]  Diseases are said to be the conjunction of the person and 
\se{duḥkha}{suffering}.
 There are four of them: invasive, bodily, mental and inherent.  The invasive ones 
 are caused by an injury.  The bodily ones are based on food, caused by 
 \se{vaiṣamya}{irregularities} in wind, bile, phlegm and blood.\footnote{Note 
 that four humoral substances are assumed here.} 
 
The \se{mānasa}{mental} ones, caused by 
\se{icchā}{desire} and 
\se{dveṣa}{hatred}, 
include:  
\se{krodha}{anger}, 
\se{āśoka}{grief}, 
\se{dainya}{misery}, 
\se{harṣa}{overexcitement}, 
 \se{kāma}{lust}, 
 \se{viṣāda}{depression},
 \se{īrṣyā}{envy},
 \se{asūyā}{jealousy},
 \se{mātsarya}{malice}, 
 and
 \se{lobha}{greed}.
 
 The \se{svābhāvika}{inherent} ones are hunger, thirst, old age, death, 
 sleep and  those of the \se{prakṛti}{temperament}.
 
 These too are \se{adhiṣṭhāna}{located} in the mind and body.
 
\se{lekhana}{Scarification},
\se{bṛṃhaṇa}{nourishment},
\se{saṃśodhana}{purification},
 \se{saṃśamana}{pacification},
 \se{āhāra}{diet} and
 \se{ācāra}{regimen}, 
 properly employed, bring about their cure.
    
 
 \item [30] Furthermore, food is the  \se{mūla}{root} of living beings as well
as of \se{bala}{strength}, \se{varṇa}{complexion} and 
\se{ojas}{vital
    energy}. It \se{āyatta}{depends on} the six \se{rasa}{flavours}.
Flavours, furthermore, have substances as their \se{āśrayin}{substrate}.  And
substances are \se{oṣadhī-}{remedies}.\footnote{Pāṇini 6.3.132 provides that
the final vowel of the noun \emph{oṣadhi} may be lengthened
(\emph{$\rightarrow$oṣadhī}) under certain conditions.  These conditions require
that the word be used in a Vedic mantra and not in the nominative.  Neither
condition is met in this passage, yet the author uses the form \emph{oṣadhī}. 
This form is in fact not uncommon in medical literature as well as in epics,
purāṇas, smṛtis, and other parts of Sanskrit literature.} There are 
two types:
\se{sthāvara}{stationary} and \se{jaṅgama}{moving}.



\item [31]  Of these, there are four types of stationary ones:
\se{vanaspati}{fruit trees}, \se{vṛkṣa}{flowering trees},
\se{oṣadhi}{herbs} and \se{vīrudh}{shrubs}.\footnote{Ca.sū.1.71--72 
also
describes these four types of medicinal plant in similar terms but with slightly
differing names: \emph{oṣadhi} is a plant that ends after fruiting, \emph{vīrudh}
is a plant that branches out, \emph{vanaspati} is a tree with fruit, and
\emph{vānaspatya} is a tree with fruit and flowers.}
Amongst these, the “fruit trees” have fruit but no flowers.\footnote{The MSS agree 
in reading \emph{phalavantyaḥ} “having flowers” which is grammatically 
non-standard. This form is also found in the  \emph{Viṣṇudharmottarapurāṇa} 
(1.92.27, \cite[1.92.27][56r]{sarm-1912}).}  The “flowering trees” 
have flowers and fruit.  The “herbs” die when the fruit is ripe. “Shrubs” put out 
shoots.

% \citep{ober-2003} didn't have anything on phalavantyo that I could find quickly.


   \item[32]  As is well known, moving remedies are also of four types: those
\se{jarāyuja}{born in in a caul}, those \se{aṇḍaja}{born from eggs},
those \se{svedaja}{born of sweat}, and \se{udbhid}{shoots}. Amongst
these, those born in a caul include \se{paśu}{animals}, humans, and
\se{vyāla}{wild animals}.  Birds, \se{sarīsṛpa}{creepy-crawlies} and
snakes are “born of eggs.” \se{kṛmi}{Worms}, \se{kunta}{small insects}
and \se{pipīlika}{ants} and others are born of sweat.\footnote{The word
\emph{kunta}, though marked as “lexical” in most dictionaries, is in fact found
in literature, commonly as a compound with \emph{pipīlika}; the compound
sometimes seems to be understood a type of ant (\emph{tatpuruṣa} compound)
rather than as a pair of insects (\emph{dvandva} compound).}  Shoots include
\se{indragopa}{red velvet mites} and \se{maṇḍūka}{frogs}.\footnote{On
\emph{indragopa}, see \cite{lien-1978}.}|

\item[33] In this context, among the stationary remedies, 
\se{tvak}{skin}, 
\se{patra}{leaves}, 
\se{puṣpa}{flowers}, 
\se{phala}{fruits},
\se{mūla}{roots},
\se{kanda}{bulbs},
\se{kṣīra}{sap},
\se{niryāsa}{resin},
\se{sāra}{essence},
\se{sneha}{oil}, 
and
\se{svarasa}{juice extract}\footnote{On \se{svarasa}{juice extract} see
CS 1.1.73, 1.4.7; Ḍalhaṇa on \Su{4.10.12}{450}.} 
are useful; among the moving remedies 
\se{carman}{pelt}, hair, nails, and 
\se{rudhira}{blood} and so forth. 

 \item[34] And \se{pārthiva}{earthen products} include gold and 
 silver.\footnote{The flow of concepts in the treatise seems to be interrupted here.}
 
\item[35] The \se{kālakṛta}{items created by time} are \se{samplava}{clusters} 
as far as wind and \se{nivāta}{no wind}, heat and shade, darkness and light
and the cold, hot and \se{varṣā}{rainy seasons} are concerned. 
The divisions of time are the
\se{nimeṣa}{blink of the eye}, a
\se{kāṣṭhā}{trice}, 
\se{kalā}{minutes}, 
\se{muhūrta}{three-quarters of an hour}, a
\se{ahorātra}{day and night}, a
\se{pakṣa}{fortnight}, a
\se{māsa}{month}, a
\se{ṛtu}{season}, a
\se{ayana}{half-year}, a
\se{saṃvatsara}{year},
and
\se{yuga}{yuga}.\footnote{These units are presented at 
\Su{1.6.5}{24} and discussed by \citet[\S\,59]{haya-2017}.}


\item[36]  These naturally cause 
\se{sañcaya}{accumulation}, 
\se{prakopa}{irritation}, 
\se{upaśama}{pacification} 
and 
\se{pratīkāra}{alleviation} of the \se{doṣa}{humours}. And they have
    \se{prayojanavat}{practical purposes}.
    
\medskip[There are verses about this:]\footnote{See footnote \ref{bha}.}

\item[37] 
\begin{sloka}
    This fourfold category is taught by physicians as a cause for the agitation and 
quelling of bodily diseases.%
%
\footnote{On the topic of the “group of four,” the commentator Ḍalhaṇa
considers them to be “food, behaviour, earthen products and items created by
time.”  He refers to the author of the lost commentary entitled \emph{Pañjikā},
and to Jejjaṭa \citep[IA, 372--3, 192]{meul-hist}.  In his view, these early
commentators  do not agree that the \se{caturvarga}{fourfold grouping} refers
to the quartet of \se{sthāvara}{stationary}, \se{jaṅgama}{moving},
\se{pārthiva}{earthen products} and \se{kālakṛta}{items created by 
time} \citep[9a]{vulgate}.}
\end{sloka}

\item[38] \begin{sloka}
    There are two kinds of invasive diseases. Some certainly\footnote{The
text uses an archaic interjection here, \emph{ha}.} \se{\mbox{nipat-}}{affect} the
mind, others the body. Their \se{kriyā}{treatment} is of two kinds too.
\end{sloka}

\item[39]\begin{sloka}
    For those that affect the body there is \se{śārīravad}{physical} 
therapy, whereas for those that affect the mind there is the 
\se{varga}{collection} of desirable sensory experiences like sound that 
bring \se{sukha }{comfort}.

\end{sloka}
%Those that affect the body have therapy that is  
%\se{śārīravat}{physical}, whereas for those of the mind it is 
 
\item [40] 

\se{evam}{Along these lines}, this brief explanation of the 
\se{catuṣtaya}{four factors}
is given: \begin{itemize}
\item    
\se{puruṣa}{human being},
\item
\se{vyadhi}{disease},
\item
\se{oṣadhi}{remedies},
\item
\se{kriyākāla}{the time for therapies}.
\end{itemize}
In this context, 
\begin{itemize}
    \item from the mention of the word “human,” the collection of 
substances that arise from it, such as the elements, and the 
\se{vikalpa}{particulars} of its major 
and minor \se{aṅga}{parts} such as 
\se{tvak}{skin}, 
\se{māṃsa}{flesh}, 
\se{sirā}{ducts}, 
\se{snāyu}{sinews}, 
\se{asthi}{bones} and 
\se{sandhi}{joints}
are meant.
 \item
From the mention of “diseases,” all diseases 
caused by
wind, bile, phlegm,
\se{sannipāta}{congested humours},
\se{āgantu}{external factors} and 
\se{svabhāva}{inherent factors} are \se{vyākhyāta}{intended}.
\item
From the mention of “remedies,”
 there is the teaching of 
 substances,
 tastes, 
 potencies,
 post-digestive tastes.
 \item
From the mention of 
“\se{kriyā}{procedures},”
\se{karman}{therapies} such as oiling
and 
\se{chedya}{excision} are taught.
\item
From the mention of the word “time,” every single teaching about the times for
procedures is meant. 

\end{itemize}
\medskip[There is a verse about this:]\footnote{See footnote \ref{bha}.}

\item[41]
 
\begin{sloka}
This seed of medicine has been declared in brief.  Its explanation will be given in one 
hundred and twenty chapters.\footnote{This is the number of chapters in the first 
five sections of the work, namely the  \emph{Sūtra-, Nidāna-, Śārīra-, Cikitsā- 
\emph{and} 
Kalpa-sthāna}s. These have 46, 16, 10, 40 and 8 chapters respectively.  The 
\emph{Uttaratantra} has 66 chapters.}
\end{sloka}

%[There are verses on this:]\q{is this bha
%really in the MSS?}

\item [42] There are one hundred and twenty chapters in five
\se{adhyāya}{sections}.\footnote{On \emph{viṃśa} in the sense of “greater by 20”
see P.5.2.46 \emph{śadantaviṃśateś ca}.}  In that regard, having divided them,
according to their subject matter, into the Ślokasthāna, the Nidāna, the Śārīra,
the Cikitsita and the Kalpa, we shall mention this in the
Uttaratantra.\footnote{The end of this sentence reads oddly.  The vulgate edition
adds an object: “[we shall mention] the remaining topics [in the Uttara]” which
smooths out the difficulty, but this is supported in none of the Nepalese MSS.  At
the start of the Uttaratantra \citep[1.3--4ab]{susr-trikamji3} there is indeed a
statement that picks up the point about there being 120 chapters.}

\medskip[There is a verse about this:]\footnote{See footnote \ref{bha}.}

\item[43]    
    \begin{sloka}
        Someone who reads this eternal proclamation of the King of
Kāśī, that was declared by Svayambhu, will have good karma on earth, will
be respected by kings and upon death will achieve the world of Śakra.
    \end{sloka}
    
\end{translation}    
    
    \newpage
    
     \section{Sūtrasthāna, adhyāya 2}
  
  
  \subsection{Literature}
  \cites[IA, 204]{meul-hist}{prei-2007}[82--83, \emph{et passim}]{wujad-2012}. 
    
    \begin{translation}    
        \item [1] 
        
    \end{translation}
    
    \newpage
    
    % % % % % % % % % % % % % % % % % % % % % % % SS 1.28
    \section{Sūtrasthāna, adhyāya 16}
    % Adhyāya 16  
    
    
    \begin{translation}    
        
        \item [1] Now we shall expound the method for
        the piercing of the ear. 
        
        \item [2] One may pierce a child's ears for the purpose of preserving and 
        decorating.
     
     During the first half of the sixth or seventh lunar month,
during a renowned day, half day, hour and constellation the child who has
\se{kṛtamaṅgala}{received a benediction}, -- \se{svastivācana}{blessings
    pronounced} -- should be placed on the lap of a wet-nurse.\footnote{The
syntax here is unclear. The expression \emph{svastivācana} may have been
a gloss inserted into the text at an earlier period to clarify
\emph{maṅgala}.  But as it stands, it is not syntactically connected to
the rest of the sentence.  The vulgate records a reading in which the
words are united in a compound that reads more naturally.} For a boy, do
the right one first; for a girl do the left one. Use a needle on a
thin ear; an \se{ārā}{awl} on a thick one.


     \item [3]  If there is excess blood or pain one should know that it was pierced
     in the wrong place. The absence of side-effects is a sign that it has been pierced 
     in the right place.\footnote{From here on, witness K is missing a folio, so the text 
     is contstructed on the basis of witnesses N and H.}
     
     \item [4] In this context, if an ignorant person accidentally pierces a 
     \se{sirā}{duct} there will 
     be fever, burning, \se{śvayathu}{swelling}, pain, \se{granthi}{lumps}, 
     \se{manyāstambhā}{paralysis 
     of the nape of the neck}, 
     \se{apatānaka}{convulsions}, headache or pain in the ear. 
     
\item[5]     Having removed the \se{varti}{wick} in it, one should smear it with
a paste of the roots of 
barley, 
liquorice, 
\se{mañjiṣṭhā}{Indian madder}, and the
\se{gandharvahasta}{castor oil tree},
thickened with honey and ghee.

When it has healed well, one should pierce it again.

\item[6] One should treat the properly-pierced ear by sprinkling it raw sesame
oil.   After every three days one should apply a thicker \se{varti}{wick} and
sprinkle oil right on it.\footnote{The manuscripts support the reading
\emph{sthūlatarīṃ} that is either a non-standard form or a scribal error.}

\item[7]
Once the ear is free from humours or side-effects, one should 
loosen it with a light \se{pravardhanaka}{dilator} in order to enlarge it.

\item[8]

\begin{sloka}

An person's ear enlarged in this way can split in two, either as a result of the 
humours or as a result of a blow.  

Listen to me about the \se{sandhāna}{joins} it 
can have. 

\end{sloka}

Here, there are, in brief, fifteen ways of mending the ear.  They are as follows:
Rim-join, Lotus-splittable, Dried Flesh, Fastening, Cheek-ear, Take away,
Ready-Split, Multi-joins, Door-hinge, Half door-hinge, 
\se{saṃkṣipta}{Compressed}, Reduced-ear,
Creeper-ear, Stick-ear, and Crow's lip.

In this context, among these, 
\begin{description}

    \item[\mdseries``Rim-join'' (\emph{nemīsandhānaka}):]
    both flaps are wide, long, and equal.

    \item[\mdseries``Lotus-splittable'' (\emph{utpalabhedyaka}):]
    both flaps are round, long, and equal.

    \item[\mdseries``Dried flesh'' (\emph{vallūraka}):]
    both flaps are short, round, and equal.

    \item[\mdseries``Fastening'' (\emph{āsaṅgima}):]
    one flap is longer on the inside.

    \item[\mdseries``Cheek-ear'' (\emph{gaṇḍakarṇa}):]
    one flap is longer on the outside.\footnote{See fig.~\ref{fig:gandakarna}.}

    \item[\mdseries``Take-away'' (\emph{āhārya}):]
    the flaps are missing, in fact, on both sides.

    \item[\mdseries``Ready-split'' (\emph{nirvedhima}):]
the flaps are like a \se{pīṭha}{dais}.

    \item[\mdseries``Multi-joins'' (\emph{vyāyojima}):]
    one flap is small, the other thick, one flap is equal, the other unequal.

    \item[\mdseries``Door-hinge'' (\emph{kapāṭasandhika}):]
    the flap on the inside is long, the other is small.

    \item[\mdseries``Half door-hinge'' (\emph{ardhakapāṭasandhika}):]
    the flap on the outside is long, the other is small.
\end{description}
    
    `These ten \se{vikalpa}{options} for \se{sandhi}{joins} of the ear should be
bound.  They can mostly be explained as resembling their names.\footnote{This
is an odd assertion, given the strangeness of the names.}  The five below are
not so successful:
    
% got to here. 2021-03-12    

The five from \se{saṃkṣipta}{compressed} on are incurable.  Among these, 
“compressed” has a dry ear canal and the other flap is small.   “Reduced ear” has 
flaps that have no base and have wasted flesh on their edges. “Creeper-ear” has 
flaps that are thin and uneven. “Stick-ear” has \se{granthita}{lumpy} flesh and the 
flaps are stretched thin and have \se{stabdha}{stiff} \se{sirā}{ducts}.  “Crow-lip” 
has a flap 
without flesh with \se{saṃkṣipta}{compressed} tips and little blood. Even when 
they are bound up, they do not heal because they are hot, inflamed, 
\se{srāva}{suppurating}, or swollen.

\item[10]  

% 15
A person wishing to perform any of these joins should therefore gather together the
supplies prepared according to the recommendations of the `Preparatory
Supplies' chapter.\footnote{SS.1.5.}  And in particular, he should gather
\se{surāmaṇḍa}{decanted liquor}, milk, water,
\se{dhānyāmla}{fermented rice-water}, and \se{kapālacūrṇa}{powdered 
earthenware crockery}.  

Next, he should prepare the woman or man who have had the ends of their hair tied 
up, have eaten lightly, and are firmly supported by qualified 
attendants.

Then, he should ready the \se{bandha}{bindings} and carry out the procedure with
\se{chedya}{cutting}, \se{bhedya}{splitting}, \se{lekhya}{scarification}, or
\se{vyadhana}{piercing}, examine the blood of the ear to know whether it is 
\se{duṣṭa}{tainted} or not.  Then if it is tainted by wind, the ear should be
bathed with \se{dhānyāmla}{fermented rice-water} and water; if tainted by choler, 
then cold water
and milk should be used; if tainted by phlegm, then \se{surāmaṇḍa}{decanted 
    liquor} and water
should be used, and then he should scarify it again.


Then, arranging the join in the ear so that it is neither proud, depressed, nor
uneven, one should make the join. Next, after anointing it with honey and ghee,
each ear should be bandaged with \se{picu}{cotton} and \se{prota}{gauze}, and
bound up with a thread, neither too tightly nor too loosely.  Then the earthenware
powder should be sprinkled on, and \se{ācārika}{medical advice} should be given.
And he should supplement with food as taught in  the `Two Wound'
chapter.\footnote{SS.4.1.}

\item[11]
\begin{sloka}
    One should avoid rubbing, sleeping during the day, exercise, overeating,
    sex, getting hot by a fire, or the effort of speaking.
\end{sloka}

\item[12]

% 17
One should not make a join when the blood is too pure, too copious, or too
thin.\footnote{The vulgate reads “impure” for the Nepalese “too pure,” which would
appear to make better medical sense.  Emending the text to \emph{nāśuddha-} for
\emph{nātiśuddha-} in the Nepalese recension would yield the same meaning as the
vulgate.} For when the ear is tainted by wind, then it is
\se{raktabaddha}{obstructed by blood}, unhealed and will peel. When tainted with
choler, is becomes \se{gāḍha}{pinched}, \se{pāka}{septic} and red.  When tainted
by phlegm, it will be \se{stabdha}{stiff} and itchy.  It has excessively copious
\se{srāva}{suppuration} and is \se{puffed up}{śopha}.  It has it has a small
amount of \se{kṣīṇa}{wasted} flesh and it will not grow.\footnote{This passage
exemplifies numerous small changes }

\item[13] When the ear is properly healed and there are no complications,  one may
very gradually start to expand it.  Otherwise, it may be \se{saṃrambha}{angry},
burning, septic or painful.  It may even be split open again.


\item [14]


Now, massage for the healthy ear, in order to enlarge it. 

\newcommand{\animal}[4]{#1 (\emph{#2}\footnote{#3 (#4)})}
\newcommand{\plant}[4]{#1 (\emph{#2}\footnote{#3 (#4)})}

One should gather as much as one can of the following: a
\animal{monitor lizard}{godhā}{Varanus bengalensis, Schneider}{Daniel
    1983:58},
\se{pratuda}{scavenging} and \se{viṣkira}{seed-eating} birds, and
creatures that live in marshes or water,\footnote{For such classifications,
see \citet{zimm-jung} and \citet{smit-clas}.} fat, marrow, milk, and sesame oil, and
white mustard oil.  
% note: think more about the compound structure here.
Then cook the oil with 
\newcommand\skt[2]{#1 (#2)}
an
\skt{admixture}{prativāpa} of the following:
\plant{purple calotropis}{arka}{Calotropis gigantea, (L.) R. Br.}{ADPS 52, AVS
    1.341, NK \#427, Potter 57, ID 306},
\plant{white calotropis}{alarka}{Calotropis procera, (Ait.) R. Br.}{NK
    \#428, GIMP 46b, ID 306},
\plant{country mallow}{balā}{Sida cordifolia, L.}{ADPS 71, NK \#2297},
\plant{`strong Indian mallow'}{atibalā}{Abutilon indicum, (L.) Sweet; Sida
    rhombifolia, L.?}{NK \#11, IGP ,4 1080; NK \#2300},
%\plant{Indian sarsaparilla}{anantā}{Hemidesmus indicus, (L.) R.
%  Br. \textnormal{and} Cryptolepis buchanani, Roemer \&
%  Schultes}{ADPS 434, AVS 3.141, NK \#1210},
%the \skt{Indian sarsaparillas}{sārive}
\plant{country sarsaparilla}{anantā}{Hemidesmus indicus, (L.) R. Br.}{ADPS 434,
    AVS 3.141--5, NK \#1210}
%and
%\plant{black creeper}{pālindī}{Ichnocarpus frutescens, (L.)
%    R.Br. \textnormal{or} Cryptolepis buchanani, Roemer \&
%    Schultes}{AVS 3.141, 3.145, 3.203, NK \#1283, \#1210, ADPS
%    434}),
%
%%\plant{prickly chaff-flower}{apāmārga}{Achyranthes aspera,
%%    L.}{GJM 524f., IMP 1.39, ADPS 44f., IMP 3.2066f., Dymock 3.135},
%\plant{Withania}{aśvagandhā}{Withania somnifera (L.) Dunal}{IMP
%    5.409f., Dymock 2.566f., Chevallier 150.},
\plant{beggarweed}{vidāri}{Desmodium
    gangeticum (L.) DC}{Dymock 1.428, GJM 602, cf.\ NK
    \#1192; ADPS 382, 414 and IMP 2.319, 4.366 are confusing},
%\plant{giant potato}{kṣīraśukla  $\rightarrow$ kṣīravidārī}{Ipmoea mauritiana,
%    Jacq.}{ADPS 510, AVS 3.222, IMP 3.1717ff.},
liquorice (\emph{madhuka}),
\plant{hornwort}{jalaśūka $\rightarrow$ jalanīlikā}{Ceratophyllum
    demersum, L.}{IMP 2371, AVS 2.56, IGP 232},\footnote{This name is not
certain: in fact, the commentator Ḍalhaṇa notes that some people interpret
it as a poisonous, hairy, air-breathing, underwater creature.}
\skt{items having the `sweet' savour}{madhuravarga},\footnote{The
items which exemplify the `sweet' savour \label{kakolyadi}
(\emph{madhuravarga}) are enumerated at SS.1.42.11.} and
\plant{`milk flower'}{payasyā  $\rightarrow$ vidārī}{Pueraria tuberosa (Willd.) 
DC.}{ADPS
    510, IMP 1.792f., AVS 4.391; not Dymock 1.424f. See GJM supplement 444,
    451, IMP 1.187, but IMP 3.1719 = Ipmoea mauritiana, Jacq.}.
%
This should then be deposited in a well-protected spot.

\item[15]% 20
\begin{sloka}
    
The wise man who been sweated should rub the \se{mardita}{massaged} ear with 
it. 
Then it will be free of complications, and will enlarge properly and be strong.
\end{sloka}

\item[16]
% 22cd-23
Ears which do not enlarge even when sweated and oiled, 
should be scarified
at the \se{apāṅga}{edge of the hole}, but not outside it.  


\item[17]
In this tradition, experts know countless repairs to ears.  So a 
 physician who is \se{suniviṣṭa}{very intent} on working in this way 
 \se{yojayed}{may repair} them.
 
 \item[18]
 % 25
 If an ear has grown hair, has a nice hole, a firm join, and is strong and
 even, well-healed, and free from pain, then one can enlarge it slowly.
 
 
 \item[19]
 
 %28, 29.
 Now I shall describe the proper method of repairing a severed nose.
 First, take from the trees a leaf the same size as the man's nose and hang it
 on him. 
 
 \item[20] Next, having cut a \se{vadhra}{slice of flesh} with the same
measurements off the cheek, the end of the nose is then scarified.\footnote{The
vulgate reads \se{baddham}{bound, connected} for \se{vadhra}{slice of flesh}.
This is a critical variant from the surgical point of view.  If the slice remains
connected, it will have a continuing blood supply.  This is one of the effective 
techniques that so astonished surgeons witnessing a similar operation in Pune in
the eighteenth century \citep[see][67--70]{wuja-roots3}.}
  
 Then the \se{apramatta}{diligent} physician, 
 should quickly \se{pratisandhā-}{put it back together} so that it is 
 \se{sādhubaddha}{well joined}.


 \footnote{Or
`\ldots\ off the cheek, it is fixed to the end of the nose, which has been
scarified'. The Sanskrit text is unfortunately not unambiguous on the
important point of whether or not the flap of grafted skin remains connected
to its original site on the cheek.} 

        % 30.
        Having carefully observed that it has been well sown up,
        two tubes should be fixed in place.  Then, having lifted them up,\footnote{The 
        Sanskrit here, \emph{unnāmayitvā} is  non-Pāṇinian.}
        the powder of
        \plant{sappanwood}{pattāṅga}{Caesalpinia sappan, L.}{AVS 1.323, IMP 
        2.847f.},\footnote{For {pattāṅga} there are manuscript variants 
\emph{pattrāṅga} (MS H) and \emph{pattaṅga}    (N).  We read with H and K 
(f.\,14r:1) on \citet[1.14.36]{vulgate}. The vulgate reads \emph{pataṅga} 
and this reading is propagated in modern dictionaries.}
        \plant{liquorice}{yaṣṭīmadhuka}{Glycyrrhiza glabra, L.}{AVS 3.84, NK \#1136},
        and
        \plant{Indian barberry}{añjana}{Berberis aristata, DC.}{Dymock 1.65, NK
            \#685, GJM 562, IGP 141}\q{añjana}
        should be applied to it.
        
       \item[22] 
     The wound should be covered properly with \skt{cotton}{picu} and should be
     moistened repeatedly with sesame oil.  Ghee should be given to the man to
     drink.  His digestion being complete, he should be oiled and purged in
     accordance with the instructions specific to him.\footnote{The expression 
     \emph{svayathopadeśa} is ungrammatical but supported in all available 
     witnesses.}   
     
     \item[23] %32.
And once healed and really come together, what is left of its \se{vadhra}{flesh}
should then be trimmed. If it is  \se{hīna}{reduced}, however, one should make an
effort to stretch it , and one should make its overgrown flesh smooth.
     
     
    \end{translation}    



\section{Uttaratantra, adhyāya 16 (17 in the vulgate)}


\subsection{Literature}

Survey of this chapter and the existing research 
on it to 2002: \cites[IA, 305--306]{meul-hist}.

History of couching in India: \cite{wujad-2019,
    desh-2000,
    desh-1999,
    bret-1826,
    leff-2020,
    shas-kaly,
    jack-1884,
    scot-1817,
    elli-1918,
    hend-1895}.


\subsection{Translation}

\begin{translation}
    
    \item[1]  Now I shall explain the \se{pratiṣedha}{counteraction} of diseases
located in the \se{dṛṣṭi}{pupil}.
    
    \item[2]  There are three \se{sādhya}{curable}, three \se{asādhya}{incurable},
and six \se{yāpya}{mitigatible} diseases located in peoples eyes. Among these,
three are \se{sādhya}{curable}. Amongst these three, %\item[3]
the \se{pratīkāra}{remedy} has been stated for the one called
“\se{dhūmadarśin}{seeing smoke}”.\footnote{This disease and its cure are 
described earlier (SS.6.7.39 and SS.6.10.16 
\citep[609 and 614]{vulgate} respectively). The latter part of this verse is
hard to construe and the text here may have been altered at an early period.}

\item[3--5ab]  When the eye is \se{vidagdha}{inflamed} by bile and when it is 
inflamed 
by phlegm, one should apply the method for removing bile and phlegm, using 
\se{nasya}{nasal medicines}, 
\se{seka}{irrigation},
\se{añjana}{application of collyrium},
\se{ālepa}{liniment}, and 
\se{puṭapāka}{medicines cooked in a crucible},
together with a \se{tarpaṇa}{balm},\footnote{These therapies are described in 
SS.6.18 
\citep[633--640]{vulgate}.} but not 
\se{śastrakṣata}{cutting with a blade}.\q{where is cutting with a knife related to 
removing bile or phlegm.}\footnote{Dalhaṇa interprets this as 
\se{sirāvedha}{blood-letting}, which is discussed in SS.1.14 \citep[]{vulgate}.}

One should drink \se{sarpis}{ghee} prepared with the \se{triphalā}{three fruits} 
and in the first [case where the problem is bile], and \se{traivṛta}{prepared with 
turpeth} in the 
latter [case, of phlegm].

And ghee \se{tailvaka}{prepared with tilvaka} is wholesome in both cases, or else 
aged ghee on its own.
% Viburnum nervosum D. Don.  Chunekar 185-186.

%gairika Hellwig 140-141
\item[5cd--7ab]

In a collyrium, these four \se{yoga}{compounds} are beneficial in both cases:
\begin{itemize}
 
    \item \se{gairika}{ochre}, \se{saindhava}{Sind salt}, \se{kṛṣṇā}{long pepper} 
    and the \se{maṣī}{black soot} from cow's teeth;\q{maṣī burned charcoal. Find 
    refs.}

\item \se{gomāṃsa}{Cow's flesh}, \se{marica}{black pepper}, \se{śirīṣa}{siris}
and \se{manaḥśilā}{red arsenic};

\item 
\se{vṛnta}{stalk} from a
\se{kapittha}{wood apple} with
\se{madhu}{honey};\footnote{\se{kapittha}{wood apple} in this verse is ablative 
singular or accusative plural, neither of which construe obviously.}

\item
or the the fruits of the \se{svayaṃgupta}{velvet bean}.
\end{itemize}

\item [8]

The physician should make a collyrium with ground up 
\se{kupyaka}{metal},\footnote{A metal other than gold or silver, according to 
\cite[1.217]{josi-maha}.  Perhaps lead, which is used in making contemporary 
collyrium.}
\se{aśoka}{Asoka tree},
\se{śālā}{Sal tree},
\se{amra}{mango},
\se{priyaṃgu}{beautyberry},
\se{nalina}{Indian lotus},
\se{utpala}{blue lotus}, 
together with
\se{hareṇu}{hareṇu},
\se{āmalaka}{emblic},
\se{pathyā}{myrobalan},
\se{pippali}{long pepper}.  
It should be combined with ghee and 
\se{kṣaudra}{honey}.


\item[9--10]

Also, when bile and phlegm have developed, the physician should apply
\se{hareṇu}{hareṇu} with the \se{svarasa}{expressed juice} of the 
flowers
from \se{amra}{mango} and \se{jambū}{Jambu} trees.

Then this collyrium, \se{vipakva}{matured} with ghee and \se{kṣaudra}{honey},
should then be applied.

\item[10--11ab]

\se{kiñjalka}{Filaments} of \se{nalina}{Indian lotus} and \se{utpala}{blue lotus},
with \se{gairika}{ochre}, and the juice of \se{gośakṛt}{cow-dung} are a collyrium
in the form of a  \se{guḍikā}{pill}.  This is good for both day and night
blindness.

\item[11cd--12ab]
 
 \se{rasāñjana}{Elixir-salve}, \se{kṣaudra}{honey}, ghee, \se{tālīśa}{scramberry},
together with gold and ochre, with the \se{gośakṛt}{juice of cow-dung} are for an
eye afflicted with bile.

\item [12cd--13] 

Alternatively, wise physician should first grind together \se{śīta}{elixir-salve}
and \se{sauvīraka}{stibnite}, \se{bhāvita}{infused} with \se{rasa}{the
    blood of birds and animals}.\footnote{This is Ḍalhaṇa's preferred interpretation
of \emph{rasa} “juice” in this context.  He also notes that some take
\se{śīta}{elixir-salve} to be camphor.}  Then he mixes it with the bile of a
tortoise or with \se{rauhita}{extract of rohu carp}.  It should always be used
with powdered collyrium to quell the bile.

\item[14]

Thus, a collyrium of \se{kārśmarī}{white teak} flowers, \se{madhuka}{liquorice},
\se{dārvī}{tree turmeric}, \se{lodhra}{lodh tree} and \se{rasāñjana}{elixir salve}
is always good as a collyrium in this case.


\item [15]

Alternatively, for those who cannot see during the day, this \se{guḍikā}{pill},
with sandalwood, is recommended: \se{nadīja}{salt}, conch shell and the three
spices, collyrium, \se{manaḥśilā}{realgar}, the two
\se{rajana}{turmerics}\footnote{Turmeric (Curcuma longa \textit{Linn.}) and tree
turmeric (Berberis aristata DC).  The term \emph{rajana} is unusual; the normal
term is \emph{rajanī}. \emph{Rajana} occurs in \emph{Suśrutanighaṇṭu} 158 in the
sense of Ferula asafoetida, Linn.} and \se{yakṛdrasa}{liver
    extract}.\footnote{This verse appears as no.\,27 in the vulgate.}

\item [16] One should grind up \se{srotoja}{kohl},\footnote{Glossed by Ḍalhaṇa as
a kind of collyrium.  Cf.\ \cite[2.M13]{nadk-1954} and \cite[197--198]{shar-1982}}
and \se{saindhava}{Sind salt} and long pepper and also \se{hareṇu}{hareṇu}.  
Such wicks with goats urine are good in a collyrium for
\se{kṣaṇadāndhya}{night blindness}.

\item [17--18ab] Alternatively, in such a case, grind together
\se{kālānusāriva}{Indian sarsaparilla}\footnote{There are two forms of
\emph{sārivā} mentioned widely in Āyurvedic literature, the white and the black.
Ideas on the identity of the black form are particularly fluid.  See
\citet[434--438]{adps} for a clear discussion.} long pepper, \se{nāgara}{dried
    ginger} and honey, the leaf of the \se{tālīśapatra}{scramberry}, the two
\se{rajana}{turmerics}, a conch shell and \se{yakṛdrasa}{liver extract}. %
Then shade-dried wicks take away \se{ruj}{illness}.


\item[18cd--19ab]

Wicks made of \se{manaḥśilā}{red arsenic},
\se{abhayā}{chebulic myrobalan},
\se{vyoṣa}{the three spices}.
\se{sāriva}{Indian sarsaparilla}, 
\se{samudraphena}{cuttlefish bone}, 
combined with goat's milk are good.

\item[19cd--21ab]

One should cook a \se{kṣaudrāñjana}{honey collyrium} either in the juices of
\se{gomūtra}{cow's urine}, and bile, \se{madirā}{spirits}, \se{yakṛt}{liver}, and
\se{dhātrī}{emblic} or else in the juice of the \se{yakṛt}{liver} of something
different, or else with the extract of the \se{triphalā}{three fruits}. %
One of these should be mixed with cow urine, ghee and \se{arṇavamala}{cuttle
    fish}\footnote{At SS 6.12.31, Ḍalhaṇa glossed \emph{arṇavamala} as
\se{samudraphena}{cuttlefish bone}. It may be worth considering whether the
unusual term \emph{arṇavamala} “ocean-filth” might refer to ambergris.} with long
pepper, honey and \se{kaṭphala}{box myrtle}. It is placed in sea salt and stored
in a bamboo tube.

\item[21cd--22]

One should cook the liver of a sheep, the ghee of a goat, with long pepper and Sindh 
salt, honey and the juice of emblics.  Then one should store it properly in a catechu 
box.  Prepared thus, the honey collyrium is good. 


\item[23]

Alternatively, a collyrium that is \se{hareṇu}{hareṇu} mixed with
\se{māgadhī}{long pepper}, the bone and the marrow of a goat, 
\se{elā}{cardamom}
and liver, together with liver extract, is good for eyes afflicted by
phlegm.\footnote{On the identities of \emph{elā} and \emph{hareṇu}
\citet[511\,ff]{watt-comm} described the former as “true” or “lesser” or “Malabar”
cardamom, Elettaria cardamomum, Maton \& White, in contrast to the “greater”
cardamom is Amomum subulatum (that he discusses on p.\,65) that is commonly 
used
as an inferior substitute for E. cardamomum. \citet[467\,f]{sing-1972} provided an
interesting discussion of \emph{hareṇu}, noting that the term refers to two
substances, first the \emph{satīna} pulse (Pisum sativum, Linn.), and second an
unknown fruit such as perhaps a Vitex. They noted, “None of the text commentators
have attempted to disclose the nature of its source plant,” although Ḍalhaṇa
described it as aromatic and identical to \emph{reṇukā} (SS.ci.2.75).%
% Sanskrit: {hareṇu}, Latin: {Amomum subulatum, Roxb.?}, References: {PVS
%Caraka 2.734, AVS 1.128, NK #154} %
}

\item[24] 

Over a fire, one should cook the \se{yakṛt}{liver} of a \se{godhā}{monitor lizard} 
prepared
with \se{antra}{entrails} and stuffed with \se{māgadhi}{long pepper}.  As is well 
known, \se{yakṛt}{liver} which is \se{niṣevita}{used} with collyrium certainly 
destroys night blindness.


\item [25] 

After preparing both a \se{plīhan}{spleen} and a liver on a spit, one should eat them
both with ghee and oil.\footnote{We read the locative as if an instrumental; if the
locative were intended then it would be the spit that would be coated with oil and
ghee.}

\item [25cd--26ab]

As is well known, there are  six diseases that can be \se{yāpya}{alleviated};
\se{tatra}{in those cases} one should release the blood by bloodletting.

And for the sake of wellbeing one should also purge using aged ghee 
\se{upahita}{combined} with 
purgative \se{aṅga}{aids}. 

\item[26cd--27]

When an eye-disease is \se{pavanodbhava}{caused by wind} they say that
\se{pañcāṅgulataila}{castor oil} mixed with milk is good.\footnote{Ḍalhaṇa says
that the unexpressed topic of this recipe is \se{timira}{partial blindness}.} In
the case of diseases of \se{śonita}{blood} and \se{pitta}{bile}, one should drink
ghee with the three fruits; it is particularly
cleansing.\footnote{\se{śonita-pitta, rakta-pitta}{Blood-bile} is a
widely-recognized disease in ayurveda, but the compound here is definitely dual,
which rules out that interpretation. One would expect blood-bile because the
previous verse } In the case of phlegm, a purgative by means of
\se{trivṛt}{turpeth} is recommended. In the case of all three humours,
\se{sugandhi}{sandal} in oil is prepared with it (turpeth).\footnote{The expression 
“\se{tailasugandhi}{the fragrant one in oil}” is puzzling. The word \emph{sugandhi} 
has different referents in the \emph{Nighaṇṭu} literature but is not common as a 
noun in the extant literature. “Sandal” is just one of its possible meanings.}
% Ḍalhaṇa (commentary on SS.ci.24.7) equates  
%\emph{trivarga}, i.e., \emph{triphalā}, with \emph{trisugandhi}.

\item[28]

In cases of \se{timira}{partial blindness}, aged ghee is recommended.  It is good if 
it is kept in an iron vessel.

\item [28cd--29ab]

One should know that ghee with the three mylobalans is always good, and it is made
with what is called \se{meṣaviṣāṇa}{periploca of the woods}.

A man who is suffering from partial blindess should lick the finely-ground three fruits 
mixed with ghee \se{sapāṇa}{off his hand}.\footnote{“Off his hand” translates the 
adverbial \emph{sapāṇam}, an unusual word. Ḍalhaṇa reproduces a reading close 
to the Nepalese recention but says that Jejjaṭa rejects it and so he also does 
\citep[627]{susr-trikamji3}.}

\item[29cd]

Alternatively, someone afflicted by phlegm should apply them (the three fruits)
mixed with oil and \se{pragāḍha}{steeped} in honey.

\item[30]

The very best oil, well-cooked with a decoction of cow-dung, is good in cases of 
partial blindness, taken as an errhine.

In cases caused by bile, ghee by itself is good, as is oil when it arises from wind and 
blood.

\item[31]
And in the case of wind one should apply
\se{trivṛt}{turpeth} based on
\se{atibalā}{strong mallow},
and \se{balā}{country mallow}
in an \se{nasya}{errhine}.\footnote{“Based on” translates \emph{-āśrita} 
“depending on” which does not construe easily here.  The vulgate has \emph{śṛta} 
“cooked” which makes easier sense but is not supported by the Nepalese MSS.}

Ghee which has been extracted from milk cooked with the meat of aquatic 
creatures and those from marshlands should be prescribed.

\item[32]
   
 \dag An \se{puṭākhya}{enclosed roasting} with  Sindh salt and the product of the
meat of a \se{kravyabhuj}{carnivore} and a \se{eṇa}{deer}, is combined with 
honey
and ghee.\footnote{Ḍalhaṇa notes \citep[628a]{vulgate} that \emph{puṭāhvaya} 
(see
verse 35 below) is a synonym for \emph{puṭapāka}, and that the process is
described in the \emph{Kriyākalpa} chapter, i.e., SS.6.18.33--38
\citep[635]{vulgate}.  On the \emph{puṭa} process in the 
\emph{Suśrutasaṃhitā},
which is earlier and different than that of \emph{rasaśāstra} literature, see the
discussion by  \citet[83]{wujad-2019}:
 \begin{quote}
    The term ‘enclosed roasting’ (\emph{puṭapāka}) does occur in
    the \emph{Suśrutasaṃhitā} in the context of eye treatments, but
    designates a method of obtaining juice from substances by
    wrapping them in leaves pasted with earth and cooking
    the bolus on charcoal to finally extract a juice.
\end{quote}
} % end of footnote

\se{vasā}{Fat} from a horse, a vulture, a snake, and a \se{tāmracūḍa}{cock},
combined with \se{madhūka}{mahua} is always good in a 
collyrium.\dag\footnote{This
verse contain irresolvable difficulties. There are no significant variants in the
Nepalese MS transmission, but the text is ungrammatical.   The vulgate reads
substantially differently but we have nevertheless made some emendations in line
with it and read the verse as two sentences.}
   
   \item [33]
   
   Having 
   \se{niṣevita}{prepared}
   a collyrium made of 
   \se{srotas}{kohl}
   and gradually combine it with
   \se{rasa}{juices}, milk and ghee.\footnote{Ḍalhaṇa specifies that the juices are 
   meat soups of various animals \citep[628]{vulgate}.}
   
For thirty days, this collyrium is put in the mouth of a black snake that is
covered with \se{kuśa}{kuśa grass}.

\item [34]

Next, a collyrium that is milk containing \se {māgadhī}{long pepper},
\se{kṣāraka}{lye} and \se{saindhava}{Sindh salt} that has been repeatedly 
prepared
with the mouth of a black snake, is good in the case of \se{rāgin
    timira}{bloodshot blindness}.\footnote{Ḍalhaṇa describes this blindness as a type
of \emph{kāca} disease caused by wind  \citep[628]{susr-trikamji3}. The expression
“bloodshot blindness” is an attempt to capture the idea of a blind eye that is
dyed or coloured (not colour-blindness). This verse is quite different from the
vulgate and also syntactically challenging.}
      
\item [35]

They say that ghee may be produced from that and combined with sweet herbs is 
good as an errhine for eye-diseases caused by bile.

And here,  a \se{tarpaṇa}{balm} is good that is a combination that is the flesh of
wild animals \se{puṭāhvaya}{taken hot}.\footnote{The expression
\se{puṭāhvaya}{taken hot} is a guess.  }


\item[36]

And 
\se{manaḥśilā}{realgar} mixed with 
\se{rasāñjana}{elixir salve} and honey is a 
\se{dravāñjana}{liquid collyrium} which is, in this case, combined with 
\se{madhūka}{mahua}.\footnote{The expression \se{dravāñjana}{liquid collyrium} is
only known from Ḍalhaṇa's comments on SS.6.17.11ab \citep[626]{vulgate}.  The
recipe in the present collyrium is different from that discussed by Ḍalhaṇa.}

Alternatively, experts on this say that finely ground \se{tuttha}{blue vitriol}
extracted from a gold mine is the \se{samāñjana}{“same collyrium”}.\footnote{The
expression \se{samāñjana}{“same collyrium”} is a hapax legomenon glossed
inexplicably by Ḍalhaṇa as “a collyrium with an equal amount of fermented barley”
(\emph{tulyasauvīrāñjana}) \citep[628]{vulgate}.}

\item [37]

Conch mixed with equal parts of sheep's horn and \se{añjana}{stibnite} removes the
impurity of the \se{kāca}{glassy opacity} because of the \se{añjana}{application
    of collyrium}.\footnote{The ablative “from collyrium” is hard to construe, but 
    Ḍalhaṇa uses this term and phrase in his commentary on \Su{6.17.41ab}{629}.}



The \se{rasa}{extracts} produced from 
a\se{palāśa}{flame of the forest},
\se{rohīta}{Rohīta tree},\footnote{Probably \emph{Soymida febrifuga} A. Juss.}
\se{madhūka}{mahua},
ground with the \se{agra}{supernatant layer} of the \se{madira}{spirits} is 
applied. 

\item[38]

 Alternatively, one should cook an errhine with 
\se{uśīra}{cuscus grass},
\se{lodhra}{lodh tree},
\se{triphalā}{the three fruits},
\se{priyaṅgu}{beauty berry} to pacify eye diseases caused by 
phlegm.\footnote{Ḍalhaṇa invokes a \se{paribhāṣā}{general rule} to indicate that 
this mixture should be cooked with sesame oil. }



One should apply smoke of the bark of
\se{vidaṅga}{embelia},
\se{pāthā}{velvet leaf},
\se{kinihī}{white siris},
and
\se{iṅgudī}{desert date}; and
\se{uśīra}{cuscus grass} alone.

\item[39] 

A ghee that is \se{bhāvita}{cooked} from a decoction of a
\se{vanaspati}{non-flowering tree}\footnote{These are fig trees.  The 
\emph{Sauśrutanighaṇṭu} (252) specifies the Uḍumbara. Cf.\ the classification in
CS.1.1.71--72, 1.8, \emph{et passim}.}
% cross-reference the description in Su.su.
as well as 
\se{haridrā}{turmeric}
and 
\se{nalada}{spikenard}
is good in a 
\se{tarpaṇa}{balm}. 


% \coffeestainD{0.4}{0.5}{90}{3 cm}{4 cm}

Alternatively, one may have an \se{puṭapāka}{enclosed roasting} done with
\se{jāṅgala}{arid-land animals}\footnote{On this term, see SS.1.35.42 
\citep[157]{vulgate} and the discussion by
\citet[25--31]{zimm-1999}.} 
and a plentiful amount of \se{māgadha}{long
    pepper}, Sindh salt and honey.

\item[40] %

A \se{kriyā}{treatment} with 
\se{manaḥśilā}{realgar}, the three spices, conch, honey, along with Sindh salt, 
\se{kāsīsa}{green vitriol} and \se{rasāñjana}{elixir salve}.\footnote{Ḍalhaṇa 
glosses  \se{kriyā}{treatment} specifically as \se{rasakriyā}{inspissation} 
\citep[629]{vulgate}.}

They say that an 
\se{rasāñjana}{elixir salve} combined with 
myrobalans, 
treacle and 
dried ginger
is good.\footnote{We emend \emph{hite} to \emph{hitam}, against the MSS.}

\item[41]

Alternatively, a collyrium that has been prepared many times in the eight types of
urine\footnote{See SS mūtravarga}\q{find ref.} is put into water with the three 
fruits. Having
stored it in the mouth of a \se{niśācara}{nocturnal creature}\footnote{Ḍalhaṇa
glosses \se{niśācara}{nocturnal creature} as “vulture,” although elsewhere in the
SS it is more commonly interpreted as a spirit or demon.  In the present context, 
following verses 33 and 34, it is probably a snake.} one should place it in a
\se{salilotthita}{conch} for two months.\footnote{We interpret
\se{salilotthita}{“water-born”} as “conch” in line with \emph{jalodbhava}, but the
term is uncertain.}


\item[42]

One should apply that collyrium 
together with the flowers of
\se{madhūka}{mahua}
and
\se{śigru}{horseradish tree}
when [the disease] is caused by all [the humours].

But alternatively, all treatments apply when blood is the cause.  The procedure
that removes bile is good when there is \se{mlāyin}{blue dot
    cataract}.\footnote{The vulgate follows Ḍalhaṇa in glossing \emph{mlāyin} as
\emph{parimlāya}.  The description of this condition at SS.6.7.27--28 appears to
refer to “blue dot” or “cerulean” cataract.  \emph{$\surd$mlai} derivatives can 
mean “dark” or “black.”), which is normally a different ailment.}\q{Check
    out these refs.}

\item[43] For one who has a humour, the physician should consider the rule in
all humoral cases and then smear the ointment on the face.\footnote{The vulgate
edition omits part of this verse (ab) combining earlier and later passages.}

The treatment that is good for removing \se{syanda}{watery eye} should be 
properly
applied in all these humoral cases, according to the individual.\footnote{The term
\se{syanda}{watery eye} refers to the specific disease \emph{abhiṣyanda}.  See
SS.6.6.5, 1.46.51, etc.}

\item[44] % tr. 47, p. 201

The physician should not employ substances in errhines etc., when the humours intensify, 
and also when disease spreads.  And further, in the \emph{Kalpa}, there is a good deal more 
said about collyriums, and that should be considered and then applied.\footnote{Ḍalhaṇa 
notes that \emph{Kalpa} means the Uttaratantra adhyāya 18 \citep[633\,ff]{vulgate}.}

\item[45] 
Someone who uses matured ghee, the three fruits, \se{śatāvarī}{wild asparagus}, 
as well as
\se{mudga}{mung beans}, emblic and barley
has nothing to fear from cases of severe \se{timira}{blindness}.

\item[46] Blindness is dispelled by milk prepared with wild asparagus or in
emblics, or again \se{yavaudana}{cooked barley} followed by the water of three
fruits with plenty of ghee.

\item[47]

When there is \se{rāgiṇi timire}{bloodshot blindness}, the wise physician should
not cut a vein.  A humour \se{utpīḍita}{injured} by the instrument rapidly
destroys vision.

\item[48] 

\se{araga timira}{Non-bloodshot blindness} in the first \se{paṭala}{layer} is
treatable.  And \se{rāgiṇi timire}{bloodshot blindness} in the second layer, with
difficulty.
%\footnote{Although the SS says \se{kṛcchra}{with difficulty}, the
%implication is that it is \se{asādhya}{impossible to treat} (cf.\ SS 6.17.2
%above).} 
And in the third layer it is \se{yāpya}{mitigable}.


\item [49]

I shall explain the therapy for success when there is a \se{liṅganāśa}{cataract}
caused by phlegm.   It may be white, like a full moon, an umbrella, a
\se{muktā}{pearl} or a \se{āvarta}{spiral}.

\item [50] Or it may be uneven, thin in the middle, streaked or have excessive
\se{prabha}{shine}. A \se{doṣa}{humour} in the pupil may be characterized as being
painful or having blood.\footnote{In the vulgate, and in parallel passages in the
AS, the reading “\se{bhavet}{it may be}” is replaced with the negative “\se{na
ced}{if, then not}” (cf.\ \AS{utt.17.1--3}{712}).   These characteristics are then
read as conditions that preclude surgery;  for the Nepalese recension, they are
simply descriptions of the appearance of a cataract.}

\item [51--52]

At a time that is neither too hot or too cold, 
the patient who has been oiled and sweated
is restrained and seated, looking symmetrically at his own nose.

The wise physician should \se{muktvā}{separate} two white sections from the
\se{kṛṣṇa}{black part} and from the \se{apāṅga}{outer corner of the eye}. Having
\se{$\surd$pīḍ}{pressed} properly into the eye,\footnote{We understand the
locative \emph{nayane} as the place of pressing; other interpreters take it as an
accusative dual.  The idea is that the eye is held steady by the surgeon.} at the
\se{daivakṛte}{naturally occurring} \se{chidra}{hole} with the \se{śalākā}{probe}
made of copper or iron, with a tip like a barley-corn that is held by a steady
hand with the middle finger, forefinger and thumb,
the left one with the right hand and the other one contrariwise.

When the piercing is done, there is the simultaneous issue of a drop of liquid and a 
sound.\footnote{Ḍalhaṇa interprets \se{samyak}{simultaneous} rather as “proper,” 
referring to the proper kind of incision.}

\item [55]

The expert should moisten the exact place of piercing  with a woman's breast-milk.  Then he 
should scratch the \se{dṛṣṭimaṇḍala}{circuit of the pupil} with the tip of the 
\se{śalākā}{probe}.\footnote{The anatomy of the eye is described in \Su{6.1.14--16}{596}  
The disks or 
\emph{maṇḍala}s are the circuits or disks of the eye.}

\item[56]
Without injuring, gently pushing the phlegm in the circuit of the pupil against the nose, he 
should remove it by means of \se{ucchiṅgana}{sniffing}.\footnote{Ḍalhaṇa describes 
\se{ucchiṅgana}{sniffing} at 
\Su{6.19.8}{641}, clearly intending inward sniffing.}


% fill in later

\item[57]



Whether the humour is \se{styāna}{solid} or \se{cala}{liquid}, one should apply
sweating to the eye externally, with \se{bhaṅga}{leaves} that remove wind, after
fixing the \se{sūcī}{needle} properly.\footnote{We interpret \emph{bhaṅga} as
leaves, following the usage elsewhere in this sthāna \Su{4.32.9, 6.11.5}{513, 614}
where \emph{bhaṅga} means \se{pallava}{shoots}. A similar procedure is
described at \AS{6.17.25}{716a}, where sweating of the eye is done by means of the
leaves of a castor-oil plant.}


\item[58]

But if the humour cannot be destroyed or if it comes back, one should apply the
\se{vyadha}{piercing} once again, with appropriate oils and so on.


\item[59]

%athābhramukte ca harir yathā dṛṣṭiḥ prakāśate 
% athābhramukto va harir

Now the \se{dṛṣṭi}{pupil} shines like the \se{hari}{sun} in a cloudless sky; then, 
when objects become visible, one may slowly remove the 
\se{śalākā}{probe}.\footnote{There are many problems with the MS readings and 
interpretation of this half-verse.  We have inferred “sky” and emended from 
“\se{agramukta}{free from the point}” to “\se{abhramukta}{free from clouds}”.  The latter 
meaning is supported (in different words) by the vulgate and occurs elsewhere in Sanskrit 
literature.}

\item[60]

Having smeared ghee on the eye, one should cover it with a bandage.  Then, he must lie 
down supine in a house free from disturbances.\footnote{Ḍalhaṇa explains disturbances 
specifically as dust, smoke, drafts and sunlight \Su{6.17.67}{631a}.} 

\item[61]
At that time, he should not belch, cough, sneeze, spit or shiver.  Afterwards there should be 
\se{yantraṇā}{restrictions} as in the case of someone who has drunk 
oil.\footnote{Ḍalhaṇa glosses “\se{yantraṇā}{restrictions}” as having a controlled diet and 
the other restrictions appropriate to someone who is taking oil as a preparation before 
further 
therapy (\Su{6.17.68}{631}). These restrictions are also described at \Su{6.18.28}{635} 
and   \Ah{1.16.25cd}{249}.}


\item[62]

Every three days one should wash it with 
\se{kaṣāya}{decoctions} that remove wind.
After three days, one should sweat the eye externally because of the danger of wind. 

\item[63]
Having restrained himself in this way for ten days he should thereafter take a beneficial 
\se{karma}{regimen} that clears the \se{dṛṣṭi}{pupil} and also he should take light food in 
measure.

\end{translation}

\subsubsection{[Complications]}

\begin{translation}
    
\item[64] 

When there is a \se{vilocana}{misshapen eyeball}, the eye may fill because of the
release of blood from a vein.\footnote{The condition of “misshapen eye” is referred
to briefly in \Su{6.61.9}{800}, where Ḍalhaṇa glosses it as
“\se{vakrabhrūnetra}{bent brow and eye}.”  The vulgate's reading of “\se{śonitena}{with 
blood}” is easier to construe.}

% many verses missing in Nepalese recension.

A hard probe leads to \se{śūla}{shooting pain}, a thin to
\se{doṣapariplava}{unsteadiness of the humours},\footnote{There is a medically
significant difference here from the vulgate, which reads “a \se{khara}{rough}
probe” not a “thin” probe.}

\item[65]

a thick-tipped probe leads to a large wound, and a sharp one may cause harm in many ways;
a very irregular one may cause a discharge of water, a \se{sthirā}{rigid} one brings about 
a \se{kriyāsaṅga}{loss of function}.\footnote{This translation of \se{kriyāsaṅga}{loss of 
function} is given on the basis of Ḍalhaṇa's gloss of \emph{kriyāsaṅgakarin} 
as “\se{gamanādikriyāvināśakarī}{causing the destruction of actions such as moving}”
at \Su{3.8.19}{382}.}

\item[66]

Therefore, one should make a good probe that is free from these defects.

\end{translation}

\subsubsection{[Characteristics of the probe]}

\begin{translation}

\item[\empty ] 
The probe should be eight finger-breadths long and in the middle it is wrapped with thread 
and is as thick as a thumb joint.  It is shaped like a bud at both \se{vaktra}{ends}.

\item[67]

A commendable probe should be made of silver, iron or 
\se{śātakumbhī}{gold}.\footnote{The vulgate reads “\se{tāmra}{copper}” in place of 
“silver.”}


\end{translation}

\subsubsection{[Complications]}

\begin{translation}
\item[\empty ] 

Redness, swelling, lumps, \se{coṣa}{driness},
\se{budbuda}{bubbling},\footnote{Ḍalhaṇa glosses “\se{budbuda}{bubbling}” as
“\se{māṃsanirgama}{prolapse} that looks like bubbles.”} \se{sūkarākṣitā}{pigs'
    eye},\footnote{The expression “pigs' eye” appears to be a \emph{hapax}.  It is
glossed as “\se{adhodṛṣṭitva}{downward vision}” by Ḍalhaṇa.},
\se{adhimantha}{irritation}, etc.\ and other diseases arise from faults in the
piercing,

\item[69--70]

or even from bad behaviour. One should treat them each accordingly.

Listen to me once again about compounds for painful red eyes.

\se{gairikaḥ}{Red chalk}, 
\se{śārivā}{Indian sarsaparilla},
\se{dūrvā}{panic grass},
and ghee ground with barley.

\item[71]

This face ointment is to be used for quelling pain and redness.  Or else it
may be taken combined with the juice of \se{mātuluṅga}{citron} with sesame gently
fried, mixed with \se{siddhārthaka}{white mustard}.\footnote{On the adverbial use
of \se{mṛdu}{gently}, see \cite{gomb-1979}.}  This is immediately beneficial when
someone is looking for relief.

\item[72]

A paste with 
\se{payasyā}{Holostemma},\footnote{The identity of \emph{payasyā} is debated
\citep[538]{sing-1972}, and was already in doubt at the time of Ḍalhaṇa but  likely 
candidates may be those suggested by Ḍalhaṇa,   who suggests either
\emph{arkapuṣpī} or \emph{kṣīrakākolī}, that may be \emph{Holostemma adakodien} 
Schult.\ 
and \emph{Leptadenia reticulata}  (Retz.) Wight \& Arn.\ 
\citep[195-196]{adps}.  The \emph{Sauśrutanighaṇṭu} glosses it as \emph{kṣīrikā} or 
\emph{arkapuṣpikā} \citep[v.\,307]{suve-2000}.} 
%
\se{śārivā}{Indian sarsaparilla}, \se{patra}{cassia cinnamon},
\se{mañjiṣṭhā}{Indian madder}, and \se{madhukair}{liquorice} stirred with goat's
milk, pleasantly warmed, is said to be healthy.\footnote{The expression
“\se{ajākṣīrārdita}{stirred with goat's milk}” is difficult.  It may be connected
with the rare root \emph{ard} documented by \citet[15]{whit-root}. Cf.\
$\surd$\emph{ard gatau} (\emph{Dhātupāṭha} 1.56).}

\item[73]

Alternatively, it can be made in this way with Himalayan cedar, Himalayan cherry and dried 
ginger.
Or, in the same way, with grapes, liquorice and the Lodh tree mixed with Sindh salt.

\item[74]

Alternatively, goats' milk with the Lodh tree, Sindh salt, red grapes and liquorice, cooked, 
should be used in irrigation because it removes pain and redness. 
% śritam for śṛtam

\item[75]



\end{translation}

% got to here

\newpage
        
% % % % % % % % % % % % % % % % % % % % % % % SS 1.28
\section{Sūtrasthāna, adhyāya 28}
% Adhyāya 28  
    
\begin{translation}    
    \item [1] Thus, living creatures and their strength,
\se{varṇa}{complexion} and \se{ojas}{energy} are rooted in food.  That
(food) depends on the six \se{rasa}{flavours}. Thus, the flavours depend
on \se{dravya}{substance}, and substances depend on medicinal herbs. 
There are two kinds of them (herbs):  stationary and mobile.\footnote{\Su{1.1.28}{7}, tr.\ 
\cite[I, 
21]{shar-susr}.}

\end{translation}



%\section{Nidānasthāna}
%Cf.\ \cite{adri-engl}.
\nocite{*} % include everything from the bib file in the bibliography

\newpage % now the end matter
    \printbiblist[heading=biblistintoc]{shorthand} % the Abbreviations (shorthands)
    
    \printbibliography[notkeyword=edition,
        notkeyword=shorthand,
        heading=bibintoc]
    
    \newpage
    {\footnotesize
            \printindex[lexical]
    \par}

    \indexprologue{\emph{\footnotesize The numbers after the colon refer to pages
        in this document.}} 

    \printindex[manuscripts]

\end{document}
