% !TeX root = incremental_SS_Translation.tex
%\setmainfont[Script=Devanagari]{Tiro Devanagari Sanskrit} % just for testing

\section{Nidānasthāna, adhyāya 1: On the diagnosis of diseases caused 
by wind \dev{वातव्याधिनिदानम्} }

\subsection{Literature}

Existing research on this chapter to 2002: \volcite{IB}[354--369]{meul-hist}.

\subsection{Translation}

\begin{translation}
    
\item [1] And now we shall explain the chapter about diagnosis of diseases
caused by wind\footnote{ Appropriate word for vātavyādhi? Diseases
    caused by wind or rheumatism?}.\q{diseases caused by wind.  Don't use 
    modern interpretative terminology.  But what you can do is have that 
    discussiona about terminology in a footnote or in your introductory remarks.}
    
    \item[2]
    
    \item[3] Dhanvantari, the foremost of the upholders of righteousness, who
    emerged with nectar, Suśruta asks after touching/holding his 
    feet.\q{subject-verb-object}
    
    \item[4] O King\footnote{ H and N both mss read \dev{भूपते }instead of 
    \dev{कौपनैः} in the
        vulgate.}! (Perhaps divodāsa)\q{Don't put guesses in the main text.  
        Footnote them.} the best of the orators! Let us know
        about the naturalized and disordered form of wind, its places in the
        body and types of the diseases caused by its contamination.
        
        \item[5--9ab] On hearing his words, the venerable sage replied that being
        independent, constant and omnipresent this wind is revealed as self-born
        and supreme being. It is situated in the form of life-force in all
        beings and worshiped by all worlds. It is the cause of origin, continued
        existence and destruction of beings. It is unmanifest though manifests
        in/through action, cold, dry, light in weight, variable, moving
        horizontally with two attributes i.e., sound and tangibility\footnote{
            According to Ḍalhaṇa, it has power to divide humours, fluids, feces
            etc. moving inside the body and it is the cause to the disease in the
            limbs. It carries humours, chyle, semen/7 fluids? and feces further in
            the body. The wind which is moving outside is holding the earth and
            body. (\dev{सा चास्य शक्तिः 
            शरीरदोषमूत्रपुरीषादिविभागोऽवयवसंस्थानका(क)रणं
            दोषधातुमलसंवहनादिश्च, शरीराद्बहिस्तु संचरतो धरणीधारणादिः  }      Su 
            1938:257)}.
            Having all chief qualities which are sattva, rajas and tamas but
            predominated by rajas. It has inconceivable power. It is inducer of
            humours\footnote{ Ḍalhaṇa suggests \dev{नेता=प्रेरक} (Su 1938:257)} and
                distinguished in the group of diseases\footnote{ Ḍalhaṇa suggests
                    \dev{राट्=राजते} not  \dev{राजा}}. 
                    \emph{\textbf{It move}\textbf{s
                            quickly, moves again and again, }}stays in stomach 
                            and
                    intestine.
                    
                    \item[9cd] Now, listen to the description of wind which moves inside 
                    the 
                    body.
                    
                    \item[10] Unvitiated wind makes possible objects of senses connect 
                    with
                    intellect. It maintains a state of equilibrium between the humours,
                    semen/7 fluids? and Gastric fluid and actions done by body, speech 
                    and
                    intellect bring to one's right place\footnote{ According to Ḍalhaṇa,
                        \dev{सम्पत्तिः=सम्पन्नता }at 1.6.3 (Su1938:23). Ḍalhaṇa 
                        commented that 
                        Gayadāsa
                        reads \dev{`इन्द्रियार्थोपसंप्राप्तिं' }but not written here because 
                        of being
                        detailed. (\dev{गयदासाचार्यस्तु इमं श्लोकं `इन्द्रियार्थोपसंप्राप्ति' 
                        इत्यादि 
                        कृत्वा पठति,
                        स च विस्तरभयान्न लिखितः) }But H and N mss suggest 
                        `\dev{इन्द्रियार्थोपसम्पत्तिः}'}.
                        
                        \item[11] Just as the five types of bile have been described based 
                        on their
                        name, place and their actions, similarly, one type of air is of five
                        types based on name, place, action and diseases.
                        
                        \item[12] Five types of wind:
                        
                        \begin{enumerate}
                            \def\labelenumi{\arabic{enumi}.}
                            \item
                            \begin{quote}
                                Vital wind (\emph{prāṇa})
                            \end{quote}
                            \item
                            \begin{quote}
                                \emph{udāna }
                            \end{quote}
                            \item
                            \begin{quote}
                                \emph{samāna}
                            \end{quote}
                            \item
                            \begin{quote}
                                \emph{vyāna}
                            \end{quote}
                            \item
                            \begin{quote}
                                \emph{apāna}
                            \end{quote}
                        \end{enumerate}
                        
                        \begin{quote}
                            above five types of wind remain in their equilibrium and hold 
                            the
                            body\footnote{ Ḍalhaṇa suggests \dev{स्थान=साम्य, 
                            यापयन्ति=धारयन्ति} (The
                                manuscripts all read \dev{प्राणोदानः समानश्च व्यानोपानस्तथैव 
                                च । 
 }                               against the
                                vulgate's \dev{प्राणोदानौ समानश्च व्यानश्चापान एव च । } I 
                                think 
                                \dev{प्राणोदानौ,
                                व्यानापानौ} or \dev{व्यानश्चापान एव च } should be read)}.
                            \end{quote}
                            
                            \item[13--14ab] The wind that flows through the mouth is 
                            called the 
                            vitality
                            (prāṇa), which holds the body. It propels down food inside the 
                            stomach
                            and engages with the gastric fluid\footnote{ Ḍalhaṇa suggests 
                            head,
                                chest, throat and nose as locations of prāṇa. (Sus1938:259) 
                                Gayadāsa
                                suggests \dev{अग्नि} for \dev{प्राण}.}. Unvitiated Vital wind 
                                mostly 
                                causes hiccups,
                                asthma etc. diseases.
                                
                                \item[14cd--15]. The wind which flows upwards in the body, 
                                the best among all
                                five winds is called udāna. Singing, speech etc. individual 
                                things done
                                by the same wind. Unvitiated udāna wind mostly causes 
                                diseases above the
                                collar bone e.g., nose, eyes, head and ears\footnote{ 
                                Ḍalhaṇa suggests
                                    it also causes diseases like cough etc. 
                                    (\dev{चकारादन्यादपि प्राणोदानौ,
                                    व्यानापानौ
                                    कासादीन् करोति ।})}.
                                    
                                    \item [16--17ab] The samāna wind flows in stomach and 
                                    duodenum. It helps
                                    gastric fluids in the digestion of food and separates the 
                                    substances
                                    produced from it e.g., chyle, impurities, urine and feces. 
                                    Unvitiated
                                    samāna wind causes diseases like a chronic enlargement 
                                    of spleen
                                    (gulma), weak digestion, and diarrhea.
                                    
                                    \item[17cd--18] The vyāna wind moves inside the whole 
                                    body 
                                    and circulates
                                    chyle and expels sweat and blood outside the body. It 
                                    helps in the
                                    movements of limbs in every way. Contaminated vyāna 
                                    wind causes all
                                    diseases occurring in the body.
                                    
                                    \item[19--20ab] Staying in the abdomen, the apāna wind 
                                    propels wind of body,
                                    feces, urine, semen, womb and menstruation to come 
                                    out of the body at
                                    their proper time. Contaminated apāna wind causes 
                                    terrible diseases that
                                    occur in the bladder and anus.
\end{translation}