% !TeX root = incremental_SS_Translation.tex
\section{Kalpasthāna, adhyāya 4}

\subsection{Introduction}
The fourth chapter of the Kalpasthāna of the \emph{Suśrutasaṃhitā} addresses 
the topic of snake bites and snake venom.  Unusually for the Nepalese version of 
the \SS, the discussion is framed as a question from Suśruta to the wise 
Dhanvantari.  Suśruta's questions are about the number of snakes, how they 
are classified, the symptoms of their bites and the pulses or stages of poisoning 
experienced by a victim of snakebite and related topics.  The taxonomy of snakes 
is presented in a presentational variant form in Figure~\ref{snakes}.


    
\subsection{Literature} 

A brief survey of this chapter's contents and a detailed assessment of the
existing research on it to 2002 was provided by Meulenbeld.\footcite[IA,
292--294]{meul-hist} There also exists a herpetological literature from
colonial India as well as more recent studies of snakes in the context of
religious life.\footnote{E.g., \cite{ewar-1878, wall-1921}. 
    \citet[75--124]{wall-1913} provided a useful analysis of the medical effects
    of snake envenomation in India arranged by the varied symptomology of
    different snakes.  He also discussed the difference between toxic effects
    and fright (69--75) and also the difficulties arising out of uncertainty
    about the effects of snake-bite (124--126).  \citet{doni-2015} provided a
    good survey of snakes as protagonists in religious literature from the
    \emph{Atharvaveda} through the epics, \emph{Purāṇas} and Buddhist
    literature. \citet[31--33 \emph{et passim}]{slou-2016} discussed the \SS's
    \emph{Kalpasthāna} as a precursor and influence on later Tantric traditions
    of snake-bite interpretation and therapy. } %
    %Translations of this chapter
    % since 2000 have appeared by
    %\textcites[131--139]{wuja-2003}[3,
    % 1--15]{shar-1999}{srik-2002}.\footnote{For a
    %    bibliography of translations to 2002, including Latin (1847),
    % English
    % (1877),
    %Gujarati (1963)
    %    and Japanese (1971), see \cite[IB, 314--315]{meul-hist}.}
    
A discussion of this chapter specifically in the light of the Nepalese
manuscripts was published by Harimoto.\footcite[101--104]{hari-2011} After a
close comparative reading of lists of poisonous snakes, Harimoto concluded
that, “the Nepalese version is internally consistent while the [vulgate]
editions are not.”  Harimoto showed how the vulgate editions had been
adjusted textually to smooth over inconsistencies, and gave insights into
these editorial processes.\footnote{The two editions that Harimoto noted,
    \cite{vulgate} and \cite{bhat-1889}, present identical texts.}


\subsection{Translation}

\begin{translation}
    \item[1] Now we shall explain the \se{kalpa}{procedure} relating to the
knowledge concerning the venom in those who have been bitten by
snakes.\footnote{The \emph{Sarvāṅgasundarī}, commenting on
    \Ah{1.16.17}{246}, glossed \dev{kalpa} as \dev{prayoga}.}
    
    \item[3] Suśruta, grasping his feet, questions the wise Dhanvantari, the 
    expert in all the sciences.
    
    \item[4]
    
    “My Lord, please speak about the number of snakes, and their divisions,
the symptoms of someone who has been bitten, and the knowledge
concerning the \se{vega}{successive shocks} of poisoning”.\footnote{The
    expression “successive shocks” translates \dev{vega}, which is other
    contexts may mean “(natural) urge.”  Here, it is rather the discrete
    stages or phases of physiological reaction to envenomation.  Cf.\ the
    symptoms of cobra poisoning described by \citet[80]{wall-1913}.}
        
    \item[5]
    
    On hearing his query, that distinguished physician spoke.
    
    “The venerable snakes such as Vāsukī and Takṣaka are uncountable. 
    
\item[6--9ab]

“They are snake-lords who support the earth, as bright as the ritual fire,
ceaselessly roaring, raining and scorching. They hold up the earth, with its
oceans, mountains and continents. If they are angered, they can destroy the
whole world with a breath and a look.  Honour to them. They have no role
here in medicine.

“The ones that I shall enumerate in due order are those mundane
ones with poison in their fangs who bite humans.\footnote{The next few
    verses are discussed in detail by \citet[101--104]{hari-2011}, who shows
    that in the taxonomy of snakes, the Nepalese version of the \SS\ has greater
    internal coherence than the vulgate recension.}


\end{translation}

    \begin{figure}
        \centering
        \Tree [.Snakes{ (80)}  
        [.Darvīkara {26 kinds} ]
        [.Maṇḍalin  {22 kinds} ]  
        [.Rājimant  {10 kinds} ]   
        [.Nirviṣa     {12 kinds} ]  
    [.Vaikarañja [.{3 kinds} {7 kinds} ] ]  ]
         \caption{The taxonomy of snakes in the vulgate, \Su{5.4.9--13ab}{571}.}
         \label{snakes1}
\end{figure}
\begin{figure}
\centering          
            \Tree [.Snakes{ (80)}  
            [.Darvīkara {26 kinds} ]
            [.Maṇḍalin  {26 kinds} ]  
            [.Rājimant  {13 kinds} ]   
            [.Nirviṣa     {12 kinds} ]  
            [.Vaikarañja {3 kinds} ]  ]
        \caption{The taxonomy of snakes in the Nepalese version.}
        \label{snakes2}
        \end{figure}
    
    \begin{translation}
        \item[9cd--10]    
        
        “There are eighty kinds of snakes and they are divided in five ways:
        Darvīkaras, Maṇḍalins, Rājimants, and Nirviṣas.  And Vaikarañjas that are
        traditionally of three kinds.\footnote{\citet{hari-2011} translated these
            names as “hooded,” “spotted,” “striped,” “harmless,” and “hybrid.” Figure 
            \ref{snakes1} shows the taxonomy described in the vulgate text; Figure 
            \ref{snakes2} shows the different and more logical division of the Nepalese 
            version of the \SS.}
            
    \item [11] 
    
    “Of those, there are twenty and six hooded snakes, and the same number
of Maṇḍalins are known.\q{Or “There are 20 phaṇins and 6 maṇḍalins.  The
    same number are known. There are 13 Rājīmants.”  Or even, “there are 20
    Phaṇins and six of them are Maṇḍalins.” Are phaṇins really the same as
    darvīkaras?}  There are thirteen Rājīmants.\footnote{The phrasing of
    this śloka is awkward.}
    
    \item [12]
    
    “There are said to be twelve Niriviṣas and, according to tradition, three 
    Vaikarañjas.
    
    \item [13--14ef]
    
“If they are trodden on, bad or provoked or looking for food, those very
angry snakes will bite.  And that is said to happen in three ways:
\se{sarpita}{serpented}, \se{darita}{torn} and thirdly \se{nirviṣa}{without
    venom}.  Some experts want to add “hurt by the snake's body”.\footnote{This 
    might refer to constriction.  The phrase reads like a commentarial addition 
    rather than the main text of the \SS.}

\item[15--16]

“An aroused snake may make one, two or more deep \se{pada}{imprints} of its
teeth, without much blood.\footnote{\label{pada-snakes} The word
    \dev{udvṛtta} “aroused” was glossed by Ḍalhaṇa at \Su{5.4.15}{571} as
    \dev{unmoṭya}, a word not found as such in standard dictionaries
    (\cite{moni-sans,apte-prac,mayr-kurz,josi-maha}. Semantic considerations
    suggest that the word is not related to $\surd$\emph{muṭ} “break” or
    \emph{mūta/mūṭa} “woven basket.” Perhaps it is related to the Tamil
    \texttamil{மோடி} (\emph{mōṭi},) whose meanings include “arrogance, grandeur,
    display” \citep[\#5133]{burr-1984} or to faintly-documented forms like
    \emph{moṭyate} “is twisted” \citep[\#10186]{CDIAL}. Ḍalhaṇa's \dev{unmoṭya}
    may thus mean “twisting up” or “making an arrogant display.” \par Note that
    \dev{pada} “imprints” is being used in the same sense as in \Su{1.13.19}{57}
    when describing the marks on the body where a knife scarifies the skin
    before leeching. See footnote \ref{pada-leeches}.} They have a
    \se{cuñcumālaka}{little row of puncture-marks},\footnote{The usual
        dictionary lexeme is \dev{cañcu} “beak, billl,” not \dev{cuñcu} as in the
        Nepalese witnesses.} making them abnormal. They are close together and
        swollen.  The physician can recognize this as “\se{sarpita}{serpented}.”

    
    % 
    %https://global.oup.com/us/companion.websites/9780190200886/student/chapter10/gline/quotation/
\end{translation}